### 2021-06-02 paper/paper.tex
set up paper template from _Genetics_ Overleaf

\documentclass[9pt,twocolumn,twoside]{gsajnl_modified}
% Use the documentclass option 'lineno' to view line numbers

\articletype{inv}

\title{Template for preparing your submission to \textit{GENETICS} using Overleaf}

\author[]{\Large Jesse D. Bloom}

\affil[]{Fred Hutchinson Cancer Research Center}
\affil[]{Howard Hughes Medical Institute}
\affil[]{Seattle, WA, USA}

\keywords{Keyword; Keyword2; Keyword3; ...}

\runningtitle{} % For use in the footer 
\runningauthor{}

\begin{abstract}
The abstract should be written for people who may not read the entire paper, so it must stand on its own. The impression it makes usually determines whether the reader will go on to read the article, so the abstract must be engaging, clear, and concise. In addition, the abstract may be the only part of the article that is indexed in databases, so it must accurately reflect the content of the article. A well-written abstract is the  most effective way to reach intended readers, leading to more robust search, retrieval, and usage of the article.

Please see additional guidelines notes on preparing your abstract below.
\end{abstract}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\marginmark
\firstpagefootnote

\correspondingauthoraffiliation{}{Corresponding author: \href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org}}
\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page

\lettrine[lines=2]{\color{color2}T}{}his \textit{Genetics} journal template is provided to help you write your work in the correct journal format. Instructions for use are provided below. Note that by default line numbers are present to aid reviewers and editors in reading and commenting on your manuscript. To remove line numbers, remove the \texttt{lineno} option from the \verb|\documentclass| declaration.

\section{Guide to using this template in Overleaf}

This template is provided to help you prepare your article for submission to the \textit{Genetics}.

\section{Author Affiliations}

For the authors' names, indicate different affiliations with the symbols: $\ast$, $\dagger$, $\ddagger$, $\S$. After four authors, the symbols double, triple, quadruple, and so forth as required.

\section{Your Abstract}

In addition to the guidelines provided in the example abstract above, your abstract should:

\begin{itemize}
\item provide a synopsis of the entire article;
\item begin with the broad context of the study, followed by specific background for the study;
\item describe the purpose, methods and procedures, core findings and results, and conclusions of the study;
\item emphasize new or important aspects of the research;
\item engage the broad readership of GENETICS and be understandable to a diverse audience (avoid using jargon);
\item be a single paragraph of less than 250 words;
\item contain the full name of the organism studied;
\item NOT contain citations or abbreviations.
\end{itemize}

\section{Introduction}

For the introduction, authors should be mindful of the broad readership of the journal. The introduction should set the stage for the importance of the work to a generalist reader and draw the reader in to the specific study. The scope and impact of the work should be clearly stated.

In individual organisms where a mutant is being studied, the rationale for the study of that mutant must be clear to a geneticist not studying that particular organism. Similarly, study of particular phenotypes should be justified broadly and not on the basis of interest for that organism alone. General background on the importance of the genetic pathway and/or phenotype should be provided in a single, well-reasoned paragraph near the beginning of the introduction.

Authors are encouraged to:

\begin{itemize}
\item cite the supporting literature completely rather than select a subset of citations;
\item provide important background citations, including relevant review papers (to help orient the non-specialist reader);
\item to cite similar work in other organisms.
\end{itemize}

\section{Materials and Methods}
\label{sec:materials:methods}

Manuscripts submitted to \textit{GENETICS} should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist. If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.  For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis. If working with mutants indicate how many independent mutants were isolated. If working with populations indicate how samples were collected and whether they were random with respect to the target population.

\subsection{Statistical Analysis}

It is important to indicate what statistical analysis has been performed; not just the name of the software and options selected, but the method and model applied. In the case of many genes being examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided. The type of correction applied should be clearly stated. It should also be clear whether the p-values reported are raw, or after correction. Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections. In large scale data exploration studies (e.g. genome wide expression studies) a clear and complete description of the replication structure must be provided.

\subsection{Data Availability}

At the end of the Materials and Methods section, include a statement on reagent and data availability. Please read the Data and Reagent Policy before writing the statement. Make sure to list the accession numbers or DOIs of any data you have placed in public repositories. List the file names and descriptions of any data you will upload as supplemental information. The statement should also include any applicable IRB numbers. You may include specifications for how to properly acknowledge or cite the data.

For example: Strains are available upon request. File S1 contains detailed descriptions of all supplemental files. File S2 contains SNP ID numbers and locations. File S3 contains genotypes for each individual. Sequence data are available at GenBank and the accession numbers are listed in File S3. Gene expression data are available at GEO with the accession number: GDS1234. Code used to generate the simulated data is provided in file S4.

\section{Results and Discussion}

The results and discussion should not be repetitive. The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context. Authors are encouraged to cite recent work relevant to their interpretations. Present and discuss results only once, not in both the Results and Discussion sections. It is sometimes acceptable to combine results and discussion. The text should be as succinct as possible. Heed Strunk and White's dictum: "Omit needless words!"

\section{Additional guidelines}

\subsection{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement. Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.

\subsection{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes". Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).

\subsection{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.  Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII. Write the names of strains in roman except when incorporating specific genotypic designations. Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme. Do not use "+" to indicate wild type. Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.

\subsection{Cross References}
Use the \verb|\nameref| command with the \verb|\label| command to insert cross-references to section headings. For example, a \verb|\label| has been defined in the section \nameref{sec:materials:methods}.

\section{In-text Citations}

Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization,Falush16}

\section{Examples of Article Components}
\label{sec:examples}

The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.

\section{First level section header}

Use this level to group two or more closely related headings in a long article.

\subsection{Second level section header}

Second level section text.

\subsubsection{Third level section header:}

Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.

\section{Figures and Tables}

Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

\subsection{Sample Figure}

Figure \ref{fig:spectrum} shows an example figure.

\begin{figure}[htbp!]
\centering
\includegraphics[width=\linewidth]{example-figure}
\caption{Example figure from \url{10.1534/genetics.114.173807}. Please include your figures in the manuscript for the review process. You can upload figures to Overleaf via the Project menu. Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).  Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.  RGB, CMYK, and Grayscale are all acceptable. Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process. Figures should be 10-20 cm in width and 1-25 cm in height. Graph axes must be exactly perpendicular and all lines of equal density.
Label multiple figure parts with A, B, etc. in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight. Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented. All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used. Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.  
}%
\label{fig:spectrum}
\end{figure}

\subsection{Sample Video}

Figure \ref{video:spectrum} shows how to include a video in your manuscript.

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{example-figure}
\caption{Example movie (the figure file above is used as a placeholder for this example). \textit{GENETICS} supports video and movie files that can be linked from any portion of the article - including the abstract. Acceptable formats include .asf, avi, .wav, and all types of Windows Media files.   
}%
\label{video:spectrum}
\end{figure}

\subsection{Sample Table}

Table \ref{tab:shape-functions} shows an example table. Avoid shading, color type, line drawings, graphics, or other illustrations within tables. Use tables for data only; present drawings, graphics, and illustrations as separate figures. Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.

Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.

\begin{table*}[htbp]
\centering
\caption{\bf Students and their grades}
\begin{tableminipage}{\textwidth}
\begin{tabularx}{\textwidth}{XXXX}
\hline
Student & Grade\footnote{This is an example of a footnote in a table. Lowercase, superscript italic letters (a, b, c, etc.) are used by default. You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
\hline
Alice & 82\% & 1 & Performed very well.\\
Bob & 65\% & 3 & Not up to his usual standard.\\
Charlie & 73\% & 2 & A good attempt.\\
\hline
\end{tabularx}
  \label{tab:shape-functions}
\end{tableminipage}
\end{table*}

\section{Sample Equation}

Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
\begin{equation}
S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i
\label{eq:refname1}
\end{equation}
denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

\bibliography{references}

\end{document}

### 2021-06-02 paper/paper.tex
start writing results about deleted project

@@
 
-\title{Template for preparing your submission to \textit{GENETICS} using Overleaf}
+\title{Recovery of a deleted deep sequencing data set sheds new light on the early spread of SARS-CoV-2 in Wuhan}
 
@@
 
-\keywords{Keyword; Keyword2; Keyword3; ...}
+\keywords{}
 
@@
 \begin{abstract}
-The abstract should be written for people who may not read the entire paper, so it must stand on its own.
-The impression it makes usually determines whether the reader will go on to read the article, so the abstract must be engaging, clear, and concise.
-In addition, the abstract may be the only part of the article that is indexed in databases, so it must accurately reflect the content of the article.
-A well-written abstract is the  most effective way to reach intended readers, leading to more robust search, retrieval, and usage of the article.
-
-Please see additional guidelines notes on preparing your abstract below.
+Abstract here
 \end{abstract}
@@
 \thispagestyle{firststyle}
-\marginmark
+%\marginmark
 \firstpagefootnote
@@
 
-\lettrine[lines=2]{\color{color2}T}{}his \textit{Genetics} journal template is provided to help you write your work in the correct journal format.
-Instructions for use are provided below.
-Note that by default line numbers are present to aid reviewers and editors in reading and commenting on your manuscript.
-To remove line numbers, remove the \texttt{lineno} option from the \verb|\documentclass| declaration.
+\lettrine[lines=2]{\color{color2}T}{}he origins...
 
-\section{Guide to using this template in Overleaf}
+\section{Results}
 
-This template is provided to help you prepare your article for submission to the \textit{Genetics}.
+\subsection{Identification of a SARS-CoV-2 deep sequencing data set that has been removed from the NCBI's Sequence Read Archive}
+During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed all SARS-CoV-2 Illumina sequencing data available as of late March 2020 from the Sequence Read Archive (SRA), which is a repository of deep sequencing data maintained by the NIH's National Center for Biotechnology Information (NCBI).
+The first supplementary table of \citet{farkas2020insights} lists all of the deep sequencing data sets available from the SRA as of March 30, 2020.
+This Excel table from \citet{farkas2020insights} is available through the journal \textit{PeerJ} at \url{https://peerj.com/articles/9255/#supp-2}, and I have digitally archived a copy in the Wayback Machine at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
 
-\section{Author Affiliations}
+Half of the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as containing data from GridION sequencing of SARS-CoV-2 PCR amplicons.
+The table indicates that the project contains a total of 141 SRA accessions, each corresponding to a different sequencing run.
+Because I had never encountered any other mention of this deep sequencing project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
+Searching for ``PRJNA612766'' in the NCBI's SRA search box was also unsuccessful, returning a message of ``No items found.''
 
-For the authors' names, indicate different affiliations with the symbols: $\ast$, $\dagger$, $\ddagger$, $\S$. After four authors, the symbols double, triple, quadruple, and so forth as required.
+I then searched for individual sequencing run accessions in the NCBI's SRA search box.
+These searches returned messages indicating that sequencing runs had been removed (Figure~\ref{fig:acc_removed}).
 
-\section{Your Abstract}
+\begin{figure}[htbp!]
+\centering
+\fbox{\includegraphics[trim={0.2in 0 0 9in},clip,width=\linewidth]{figures/acc_removed.png}}
+\caption{Searching for individual accessions from deleted deep sequencing project PRJNA612766 revealed they had been removed from the SRA.
+Shown is the result of searching for ``SRR11313485'' to reach the \url{https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
+A copy of this result has been digitally archived on the Wayback Machine at \url{https://web.archive.org/web/20210502131630/https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
+}%
+\label{fig:acc_removed}
+\end{figure}
 
-In addition to the guidelines provided in the example abstract above, your abstract should:
+\subsection{The deleted data set contains sequencing of viral samples collected early in the Wuhan outbreak}
 
-\begin{itemize}
-\item provide a synopsis of the entire article;
-\item begin with the broad context of the study, followed by specific background for the study;
-\item describe the purpose, methods and procedures, core findings and results, and conclusions of the study;
-\item emphasize new or important aspects of the research;
-\item engage the broad readership of GENETICS and be understandable to a diverse audience (avoid using jargon);
-\item be a single paragraph of less than 250 words;
-\item contain the full name of the organism studied;
-\item NOT contain citations or abbreviations.
-\end{itemize}
-
-\section{Introduction}
-
-For the introduction, authors should be mindful of the broad readership of the journal.
-The introduction should set the stage for the importance of the work to a generalist reader and draw the reader in to the specific study.
-The scope and impact of the work should be clearly stated.
-
-In individual organisms where a mutant is being studied, the rationale for the study of that mutant must be clear to a geneticist not studying that particular organism.
-Similarly, study of particular phenotypes should be justified broadly and not on the basis of interest for that organism alone.
-General background on the importance of the genetic pathway and/or phenotype should be provided in a single, well-reasoned paragraph near the beginning of the introduction.
-
-Authors are encouraged to:
-
-\begin{itemize}
-\item cite the supporting literature completely rather than select a subset of citations;
-\item provide important background citations, including relevant review papers (to help orient the non-specialist reader);
-\item to cite similar work in other organisms.
-\end{itemize}
-
-\section{Materials and Methods}
-\label{sec:materials:methods}
-
-Manuscripts submitted to \textit{GENETICS} should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist.
-If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.
-For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis.
-If working with mutants indicate how many independent mutants were isolated.
-If working with populations indicate how samples were collected and whether they were random with respect to the target population.
-
-\subsection{Statistical Analysis}
-
-It is important to indicate what statistical analysis has been performed; not just the name of the software and options selected, but the method and model applied.
-In the case of many genes being examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided.
-The type of correction applied should be clearly stated.
-It should also be clear whether the p-values reported are raw, or after correction.
-Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.
-In large scale data exploration studies (e.g.
-genome wide expression studies) a clear and complete description of the replication structure must be provided.
-
-\subsection{Data Availability}
-
-At the end of the Materials and Methods section, include a statement on reagent and data availability.
-Please read the Data and Reagent Policy before writing the statement.
-Make sure to list the accession numbers or DOIs of any data you have placed in public repositories.
-List the file names and descriptions of any data you will upload as supplemental information.
-The statement should also include any applicable IRB numbers.
-You may include specifications for how to properly acknowledge or cite the data.
-
-For example: Strains are available upon request.
-File S1 contains detailed descriptions of all supplemental files.
-File S2 contains SNP ID numbers and locations.
-File S3 contains genotypes for each individual.
-Sequence data are available at GenBank and the accession numbers are listed in File S3.
-Gene expression data are available at GEO with the accession number: GDS1234.
-Code used to generate the simulated data is provided in file S4.
-
-\section{Results and Discussion}
-
-The results and discussion should not be repetitive.
-The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context.
-Authors are encouraged to cite recent work relevant to their interpretations.
-Present and discuss results only once, not in both the Results and Discussion sections.
-It is sometimes acceptable to combine results and discussion.
-The text should be as succinct as possible.
-Heed Strunk and White's dictum: "Omit needless words!"
-
-\section{Additional guidelines}
-
-\subsection{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement.
-Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.
-
-\subsection{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes".
-Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).
-
-\subsection{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.
-Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII.
-Write the names of strains in roman except when incorporating specific genotypic designations.
-Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme.
-Do not use "+" to indicate wild type.
-Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.
-
-\subsection{Cross References}
-Use the \verb|\nameref| command with the \verb|\label| command to insert cross-references to section headings.
-For example, a \verb|\label| has been defined in the section \nameref{sec:materials:methods}.
-
-\section{In-text Citations}
-
-Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization,Falush16}
-
-\section{Examples of Article Components}
-\label{sec:examples}
-
-The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
-
-\section{First level section header}
-
-Use this level to group two or more closely related headings in a long article.
-
-\subsection{Second level section header}
-
-Second level section text.
-
-\subsubsection{Third level section header:}
-
-Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
+\subsection{Recovery of the data from the Google and Amazon clouds}
 
 \section{Figures and Tables}
-
-Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
 
@@
 \includegraphics[width=\linewidth]{example-figure}
-\caption{Example figure from \url{10.1534/genetics.114.173807}.
-Please include your figures in the manuscript for the review process.
-You can upload figures to Overleaf via the Project menu.
-Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).
-Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.
-RGB, CMYK, and Grayscale are all acceptable.
-Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process.
-Figures should be 10-20 cm in width and 1-25 cm in height.
-Graph axes must be exactly perpendicular and all lines of equal density.
-Label multiple figure parts with A, B, etc.
-in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight.
-Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented.
-All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used.
-Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.
+\caption{captions  
 }%
 \label{fig:spectrum}
-\end{figure}
-
-\subsection{Sample Video}
-
-Figure \ref{video:spectrum} shows how to include a video in your manuscript.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=\linewidth]{example-figure}
-\caption{Example movie (the figure file above is used as a placeholder for this example).
-\textit{GENETICS} supports video and movie files that can be linked from any portion of the article - including the abstract.
-Acceptable formats include .asf, avi, .wav, and all types of Windows Media files.
-}%
-\label{video:spectrum}
 \end{figure}
@@
 
-\section{Sample Equation}
-
-Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-\begin{equation}
-S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i
-\label{eq:refname1}
-\end{equation}
-denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
-
 \bibliography{references}

### 2021-06-02 paper/paper.tex
finished first Results section

@@
 \subsection{Identification of a SARS-CoV-2 deep sequencing data set that has been removed from the NCBI's Sequence Read Archive}
-During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed all SARS-CoV-2 Illumina sequencing data available as of late March 2020 from the Sequence Read Archive (SRA), which is a repository of deep sequencing data maintained by the NIH's National Center for Biotechnology Information (NCBI).
-The first supplementary table of \citet{farkas2020insights} lists all of the deep sequencing data sets available from the SRA as of March 30, 2020.
-This Excel table from \citet{farkas2020insights} is available through the journal \textit{PeerJ} at \url{https://peerj.com/articles/9255/#supp-2}, and I have digitally archived a copy in the Wayback Machine at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
+During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed SARS-CoV-2 deep sequencing data from the Sequence Read Archive (SRA), which is a repository maintained by the NIH's National Center for Biotechnology Information (NCBI).
+The first supplementary table of \citet{farkas2020insights} lists all SARS-CoV-2 deep sequencing data sets available from the SRA as of March 30, 2020.
+This table is available in Excel format at \url{https://peerj.com/articles/9255/#supp-2}, and I have digitally archived a copy in the Wayback Machine at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
 
-Half of the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as containing data from GridION sequencing of SARS-CoV-2 PCR amplicons.
-The table indicates that the project contains a total of 141 SRA accessions, each corresponding to a different sequencing run.
-Because I had never encountered any other mention of this deep sequencing project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
-Searching for ``PRJNA612766'' in the NCBI's SRA search box was also unsuccessful, returning a message of ``No items found.''
+Half the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as involving GridION sequencing of SARS-CoV-2 amplicons.
+The table indicates the project contains 141 SRA accessions, each corresponding to a different sequencing run.
+Because I had never encountered any other mention of this sequencing project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
+Searching for ``PRJNA612766'' in the NCBI's SRA search box returned a message of ``No items found.''
 
@@
 
-\begin{figure}[htbp!]
+\begin{figure}[]
 \centering
 \fbox{\includegraphics[trim={0.2in 0 0 9in},clip,width=\linewidth]{figures/acc_removed.png}}
-\caption{Searching for individual accessions from deleted deep sequencing project PRJNA612766 revealed they had been removed from the SRA.
+\caption{Accessions from deep sequencing project PRJNA612766 have been removed from the SRA.
 Shown is the result of searching for ``SRR11313485'' to reach the \url{https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
@@
 
+The SRA is designed to serve as a permanent archive of deep sequencing data.
+The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed''~\citep{SRA_deletion}.
+The documentation states that data can only be deleted by directly e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
+An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows e-mails between NCBI staff and the lead author of a paper on pangolin coronaviruses to which the journal \textit{Nature} added an editor's note on November 11, 2020, due to concerns about the identity of the underlying samples~\citep{chan2020single}.
+
+\begin{figure}[]
+\centering
+\fbox{\includegraphics[trim={0.8in 0.8in 1.1in 1in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
+\caption{Example of the process used to delete SRA data.
+The image shows an excerpt of e-mail exchanges between the lead author of the pangolin coronavirus paper \citet{xiao2020isolation} and SRA staff.
+The full e-mails are available at \url{https://usrtk.org/wp-content/uploads/2020/12/NCBI-Emails.pdf}.
+}
+\label{fig:pangolin_emails}
+\end{figure}
+
+Taken together, these facts suggest that subsequent to March 30, 2020, someone must have submitted an e-mail request to the NCBI requesting deletion of the SARS-CoV-2 deep sequencing project PRJNA612766.
+
 \subsection{The deleted data set contains sequencing of viral samples collected early in the Wuhan outbreak}
@@
 \section{Figures and Tables}
-
-\subsection{Sample Figure}
-
-Figure \ref{fig:spectrum} shows an example figure.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=\linewidth]{example-figure}
-\caption{captions  
-}%
-\label{fig:spectrum}
-\end{figure}
 

### 2021-06-03 paper/paper.tex
second results subsection

@@
 The first supplementary table of \citet{farkas2020insights} lists all SARS-CoV-2 deep sequencing data sets available from the SRA as of March 30, 2020.
-This table is available in Excel format at \url{https://peerj.com/articles/9255/#supp-2}, and I have digitally archived a copy in the Wayback Machine at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
+I have digitally archived a copy of this table in the Wayback Machine at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
 
-Half the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as involving GridION sequencing of SARS-CoV-2 amplicons.
-The table indicates the project contains 141 SRA accessions, each corresponding to a different sequencing run.
+Half the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as Oxford Nanopore sequencing of SARS-CoV-2 amplicons.
+The table indicates this project contains 141 SRA accessions, each corresponding to a different sequencing run.
 Because I had never encountered any other mention of this sequencing project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
@@
 The SRA is designed to serve as a permanent archive of deep sequencing data.
-The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed''~\citep{SRA_deletion}.
-The documentation states that data can only be deleted by directly e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
-An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows e-mails between NCBI staff and the lead author of a paper on pangolin coronaviruses to which the journal \textit{Nature} added an editor's note on November 11, 2020, due to concerns about the identity of the underlying samples~\citep{chan2020single}.
+The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by directly e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
+An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows e-mails between NCBI staff and the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to which the journal \textit{Nature} added an editor's note on November 11, 2020, due to concerns about the identity of the underlying samples~\citep{chan2020single}.
 
@@
 \fbox{\includegraphics[trim={0.8in 0.8in 1.1in 1in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
-\caption{Example of the process used to delete SRA data.
-The image shows an excerpt of e-mail exchanges between the lead author of the pangolin coronavirus paper \citet{xiao2020isolation} and SRA staff.
-The full e-mails are available at \url{https://usrtk.org/wp-content/uploads/2020/12/NCBI-Emails.pdf}.
+\caption{Example of the process to delete SRA data.
+The image shows an excerpt of e-mails between the lead author of the pangolin coronavirus paper \citet{xiao2020isolation} and SRA staff.
+The full e-mails are at \url{https://usrtk.org/wp-content/uploads/2020/12/NCBI-Emails.pdf}.
 }
@@
 
-Taken together, these facts suggest that subsequent to March 30, 2020, someone must have submitted an e-mail request to the NCBI requesting deletion of the SARS-CoV-2 deep sequencing project PRJNA612766.
+Taken together, these facts indicate that subsequent to March 30, 2020, someone must have e-mailed the NCBI to reques deletion of the SARS-CoV-2 deep sequencing project PRJNA612766.
 
 \subsection{The deleted data set contains sequencing of viral samples collected early in the Wuhan outbreak}
+The metadata in the first supplementary table of \citet{farkas2020insights} indicates that the samples in the deleted project PRNJA612766 were collected by Aisu Fu and Renmin Hospital of Wuhan University.
+Google searching for these terms revealed the samples were related to a study posted as pre-print on \textit{medRxiv} in early March of 2020~\citep{Wang2020medRxiv}, and subsequently published in the journal \textit{Small} in June of 2020~\citep{Wang2020small}.
 
-\subsection{Recovery of the data from the Google and Amazon clouds}
+The study describes an approach to diagnose infection with SARS-CoV-2 and other respiratory viruses by Oxford Nanopore sequencing.
+This approach involved reverse-transcription of total RNA from swab samples, followed by PCR with specific primers that generated amplicons covering portions of the viral genome.
+These amplicons were then sequenced on an Oxford Nanopore GridION, and infection with SARS-CoV-2 was diagnosed if the resulting sequencing yielded a sufficiently large number of reads aligning to the SARS-CoV-2 genome.
+Importantly, the study notes that this approach yields genetic information about the sequence of the virus as well enabling diagnosis of infection.
 
-\section{Figures and Tables}
+The pre-print~\citep{Wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
+The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be approved by a centralized group to ensure that scientific publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
+The final published manuscript~\citep{Wang2020small} updated the description of these 45 samples to say that they were obtained ``early in the epidemic (January 2020).''
+Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
+In addition, they both state that the approach was then later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
 
-\subsection{Sample Table}
+There is complete concordance between the deep sequencing accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the biological samples described by \citet{Wang2020medRxiv}.
+Specifically, there are 89 sequencing accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
+The number of accessions is approximately twice the number of early epidemic samples because each sample has data for two nanopore sequencing runtimes except one sample (B5) with just one runtime.
+ There are 31 sequencing accessions corresponding to the 16 samples collected in February from Renmin Hospital patients, with these samples named R01, R02, etc.
+ Again, the number of accessions is approximately twice the number of samples because all but one sample (R04) has data for two different sequencing runtimes.
+ In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for the initial benchmarking and development of the approach.
+ Together, these samples and controls account for all 241 accessions listed for PRJNA612766 in the supplementary table of \citet{farkas2020insights}.
 
-Table \ref{tab:shape-functions} shows an example table.
-Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
-Use tables for data only; present drawings, graphics, and illustrations as separate figures.
-Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
+Neither the pre-print~\citep{Wang2020medRxiv} or published manuscript~\citep{Wang2020small} contain any corrections or notes that indicate a scientific reason for deleting the study's sequencing data from the SRA.
 
-Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
-
-\begin{table*}[htbp]
-\centering
-\caption{\bf Students and their grades}
-\begin{tableminipage}{\textwidth}
-\begin{tabularx}{\textwidth}{XXXX}
-\hline
-Student & Grade\footnote{This is an example of a footnote in a table.
-Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
-You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
-\hline
-Alice & 82\% & 1 & Performed very well.\\
-Bob & 65\% & 3 & Not up to his usual standard.\\
-Charlie & 73\% & 2 & A good attempt.\\
-\hline
-\end{tabularx}
-  \label{tab:shape-functions}
-\end{tableminipage}
-\end{table*}
+\subsection{Recovery of the deleted sequencing data from the Google cloud}
 

### 2021-06-03 paper/paper.tex
first part of recovery of deleted data results

@@
 % Use the documentclass option 'lineno' to view line numbers
+
+\usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
+
+\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
 
@@
 \caption{Example of the process to delete SRA data.
-The image shows an excerpt of e-mails between the lead author of the pangolin coronavirus paper \citet{xiao2020isolation} and SRA staff.
+The image shows excerpted e-mails between the lead author of the pangolin coronavirus paper~\citep{xiao2020isolation} and SRA staff.
 The full e-mails are at \url{https://usrtk.org/wp-content/uploads/2020/12/NCBI-Emails.pdf}.
@@
 
-\subsection{The deleted data set contains sequencing of viral samples collected early in the Wuhan outbreak}
-The metadata in the first supplementary table of \citet{farkas2020insights} indicates that the samples in the deleted project PRNJA612766 were collected by Aisu Fu and Renmin Hospital of Wuhan University.
+\subsection{The deleted data set contains sequencing of viral samples collected early in the Wuhan epidemic}
+The metadata in the first supplementary table of \citet{farkas2020insights} indicates that the samples in deleted project PRNJA612766 were collected by Aisu Fu and Renmin Hospital of Wuhan University.
 Google searching for these terms revealed the samples were related to a study posted as pre-print on \textit{medRxiv} in early March of 2020~\citep{Wang2020medRxiv}, and subsequently published in the journal \textit{Small} in June of 2020~\citep{Wang2020small}.
 
-The study describes an approach to diagnose infection with SARS-CoV-2 and other respiratory viruses by Oxford Nanopore sequencing.
+The study describes an approach to diagnose infection with SARS-CoV-2 and other respiratory viruses by nanopore sequencing.
 This approach involved reverse-transcription of total RNA from swab samples, followed by PCR with specific primers that generated amplicons covering portions of the viral genome.
-These amplicons were then sequenced on an Oxford Nanopore GridION, and infection with SARS-CoV-2 was diagnosed if the resulting sequencing yielded a sufficiently large number of reads aligning to the SARS-CoV-2 genome.
-Importantly, the study notes that this approach yields genetic information about the sequence of the virus as well enabling diagnosis of infection.
+These amplicons were then sequenced on an Oxford Nanopore GridION, and infection with SARS-CoV-2 was diagnosed if the resulting sequencing yielded a sufficient number of reads aligning to the SARS-CoV-2 genome.
+Importantly, the study notes that this approach yields information about the sequence of the virus as well enabling diagnosis of infection.
 
@@
 
-\subsection{Recovery of the deleted sequencing data from the Google cloud}
+\subsection{Recovery of deleted sequencing data from the Google cloud} 
+As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
+In addition, none of the sequencing runs could be accessed using the command-line tools of the SRA toolkit.
+For instance, as of June 3, 2020, running the commands \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} both returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313395'``.
+
+However, the SRA has begun storing all data on the Google and Amazon clouds.
+While inspecting the SRA's data access web interface for other sequencing accessions, I noticed that the SRA files were often available from links to the cloud such as \texttt{https://storage.googleapis.com/nih-sequence-read-archive/run/<ACCESSION>/<ACCESSION>} where \texttt{<ACCESSION>} is the accession.
+
+Based on the hypothesis that deletion of sequencing runs by the SRA might not delete files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
+This approach was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
+I have digitally archived this URL on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
 

### 2021-06-03 scripts/sra_file_info.py
use `vdb-dump` to get object creation timestamp and info timestamp for all SRA files

@@ -1,0 +1,44 @@
+"""Implements ``snakemake`` rule `sra_file_info`."""
+
+
+import os
+import re
+import subprocess
+
+import pandas as pd
+
+
+sra_files = snakemake.input.sra_files
+accessions = [os.path.splitext(os.path.basename(sra_file))[0]
+              for sra_file in sra_files]
+
+print(f"Getting information for {len(sra_files)} SRA files.")
+
+records = []
+for acc, f in zip(accessions, sra_files):
+
+    obj_ts = subprocess.run(['vdb-dump', '--obj_timestamp', f],
+                            capture_output=True,
+                            text=True).stdout.strip()
+
+    info = subprocess.run(['vdb-dump', '--info', f],
+                          capture_output=True,
+                          text=True).stdout
+    info_m = re.search('TIME\s+\:\s\w+\s\((\d{2}/\d{2}/\d{4} \d{2}\:\d{2})\)',
+                       info)
+    if not info_m:
+        raise ValueError(f"did not match info:\n{info}")
+    info_date = info_m.group(1)
+
+    records.append((acc, f, obj_ts, info_date))
+
+print(f"Writing results to {snakemake.output.csv}")
+(pd.DataFrame(records,
+              columns=['accession', 'sra_file', 'object_timestamp',
+                       'info_timestamp'])
+ .assign(object_timestamp=lambda x: pd.to_datetime(x['object_timestamp']),
+         info_timestamp=lambda x: pd.to_datetime(x['info_timestamp']),
+         )
+ .to_csv(snakemake.output.csv, index=False)
+ )
+

### 2021-06-03 paper/paper.tex
Results thru recovery of sequencing data

@@
 As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
-In addition, none of the sequencing runs could be accessed using the command-line tools of the SRA toolkit.
+In addition, none of the sequencing runs could be accessed using the command-line tools of the SRA Toolkit.
 For instance, as of June 3, 2020, running the commands \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} both returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313395'``.
@@
 Based on the hypothesis that deletion of sequencing runs by the SRA might not delete files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
-This approach was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
+This strategy was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
 I have digitally archived this URL on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
+
+I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 of 45 early epidemic samples that were SARS-CoV-2 positive, as well as the 16 hospital samples from February (SRA files for SRR11313490 and SRR11313499 were not accessible via the cloud).
+These SRA files are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
+I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for each SRA file.
+For all files, the object timestamp is February 15, 2020, and the time is March 16, 2020.
+Although the SRA Toolkit does not clearly document the definition of these two properties, my \emph{guess} is that the object timestamp may refer to when the SRA file was created from a FASTQ file uploaded to the SRA, and the time may refer to when the accession was made public.
+
+\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for many samples}
 

### 2021-06-04 paper/paper.tex
add coverage plots to paper

@@
 
-Half the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as Oxford Nanopore sequencing of SARS-CoV-2 amplicons.
+Half the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as nanopore sequencing of SARS-CoV-2 amplicons.
 The table indicates this project contains 141 SRA accessions, each corresponding to a different sequencing run.
@@
 The SRA is designed to serve as a permanent archive of deep sequencing data.
-The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by directly e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
-An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows e-mails between NCBI staff and the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to which the journal \textit{Nature} added an editor's note on November 11, 2020, due to concerns about the identity of the underlying samples~\citep{chan2020single}.
+The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
+An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows e-mails between NCBI staff and the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation}.
 
@@
 The image shows excerpted e-mails between the lead author of the pangolin coronavirus paper~\citep{xiao2020isolation} and SRA staff.
-The full e-mails are at \url{https://usrtk.org/wp-content/uploads/2020/12/NCBI-Emails.pdf}.
+The full e-mails are at \url{https://usrtk.org/tag/pangolin-papers/}.
 }
@@
 
-Taken together, these facts indicate that subsequent to March 30, 2020, someone must have e-mailed the NCBI to reques deletion of the SARS-CoV-2 deep sequencing project PRJNA612766.
+Taken together, these facts indicate that subsequent to March 30, 2020, someone must have e-mailed the NCBI to request deletion of the SARS-CoV-2 deep sequencing project PRJNA612766.
 
@@
 There is complete concordance between the deep sequencing accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the biological samples described by \citet{Wang2020medRxiv}.
-Specifically, there are 89 sequencing accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
+There are 89 accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
 The number of accessions is approximately twice the number of early epidemic samples because each sample has data for two nanopore sequencing runtimes except one sample (B5) with just one runtime.
  There are 31 sequencing accessions corresponding to the 16 samples collected in February from Renmin Hospital patients, with these samples named R01, R02, etc.
- Again, the number of accessions is approximately twice the number of samples because all but one sample (R04) has data for two different sequencing runtimes.
- In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for the initial benchmarking and development of the approach.
+ Again, all but one sample (R04) has data for two different sequencing runtimes.
+ In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for development and benchmarking of the approach.
  Together, these samples and controls account for all 241 accessions listed for PRJNA612766 in the supplementary table of \citet{farkas2020insights}.
@@
 However, the SRA has begun storing all data on the Google and Amazon clouds.
-While inspecting the SRA's data access web interface for other sequencing accessions, I noticed that the SRA files were often available from links to the cloud such as \texttt{https://storage.googleapis.com/nih-sequence-read-archive/run/<ACCESSION>/<ACCESSION>} where \texttt{<ACCESSION>} is the accession.
+While inspecting the SRA's web interface for other sequencing accessions, I noticed that SRA files are often available from links to the cloud such as \texttt{https://storage.googleapis.com/nih-sequence-read-archive/run/<ACCESSION>/<ACCESSION>} where \texttt{<ACCESSION>}.
 
-Based on the hypothesis that deletion of sequencing runs by the SRA might not delete files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
+Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
 This strategy was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
-I have digitally archived this URL on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
+I have archived this URL on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
 
@@
 \subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for many samples}
+I aligned the recovered deep sequencing data to the SARS-CoV-2 genome, pooling the data when there were multiple accessions for a single biological sample (this was usually the case as most samples were sequenced using two runtimes as described above).
+Figure~\ref{fig:coverage} shows the sequencing coverage over the genome for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples.
+
+\begin{figure*}[]
+\centering
+\includegraphics[width=\linewidth]{../results/pileup/coverage_region.pdf}
+\caption{Sequencing depth over the SARS-CoV-2 genome from nucleotide site 21,563 to 29,674 for the 34 virus-positive early outbreak samples and the 16 samples from hospitalized patients in February.
+Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
+The dashed gray line is the minimum coverage required to call a consensus identity at a site.
+Note that the y-axis uses a symlog scale.
+An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see statistics is at {\color{red} add link}.
+A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
+}
+\label{fig:coverage}
+\end{figure*}
 
@@
 
+\clearpage
+\onecolumn
+\renewcommand{\thepage}{S\arabic{page}}
+\setcounter{page}{1}
+
+\section{Supplementary Material}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\linewidth]{../results/pileup/coverage_all.pdf}
+\caption{A version of Figure~\ref{fig:coverage} that shows coverage over the full length of the SARS-CoV-2 genome.
+An interactive version of this plot is at {\color{red} add link}.
+}
+\label{suppfigfig:coverage_all}
+\end{figure}
+
 \end{document}

### 2021-06-05 paper/paper.tex
start writing results on coverage

@@
 
-Half the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described in the table as nanopore sequencing of SARS-CoV-2 amplicons.
-The table indicates this project contains 141 SRA accessions, each corresponding to a different sequencing run.
+The majority the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described as nanopore sequencing of SARS-CoV-2 amplicons.
+The table indicates this project represents 241 of the 282 SARS-CoV-2 sequencing run accessions in the SRA as of March, 30, 2020.
 Because I had never encountered any other mention of this sequencing project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
 Searching for ``PRJNA612766'' in the NCBI's SRA search box returned a message of ``No items found.''
-
 I then searched for individual sequencing run accessions in the NCBI's SRA search box.
@@
 
-The SRA is designed to serve as a permanent archive of deep sequencing data.
+The SRA is designed as a permanent archive of deep sequencing data.
 The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
@@
 The metadata in the first supplementary table of \citet{farkas2020insights} indicates that the samples in deleted project PRNJA612766 were collected by Aisu Fu and Renmin Hospital of Wuhan University.
-Google searching for these terms revealed the samples were related to a study posted as pre-print on \textit{medRxiv} in early March of 2020~\citep{Wang2020medRxiv}, and subsequently published in the journal \textit{Small} in June of 2020~\citep{Wang2020small}.
+Google searching for these terms revealed the samples were related to a study posted as a pre-print on \textit{medRxiv} in early March of 2020~\citep{wang2020medRxiv}, and subsequently published in the journal \textit{Small} in June of 2020~\citep{wang2020small}.
 
@@
 This approach involved reverse-transcription of total RNA from swab samples, followed by PCR with specific primers that generated amplicons covering portions of the viral genome.
-These amplicons were then sequenced on an Oxford Nanopore GridION, and infection with SARS-CoV-2 was diagnosed if the resulting sequencing yielded a sufficient number of reads aligning to the SARS-CoV-2 genome.
+These amplicons were then sequenced on an Oxford Nanopore GridION, and infection with SARS-CoV-2 was diagnosed if the sequencing yielded a sufficient number of reads aligning to the SARS-CoV-2 genome.
 Importantly, the study notes that this approach yields information about the sequence of the virus as well enabling diagnosis of infection.
 
-The pre-print~\citep{Wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
+The pre-print~\citep{wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
 The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be approved by a centralized group to ensure that scientific publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
-The final published manuscript~\citep{Wang2020small} updated the description of these 45 samples to say that they were obtained ``early in the epidemic (January 2020).''
+The final published manuscript~\citep{wang2020small} from June of 2020 updated the description from ``early in the epidemic'' to ``early in the epidemic (January 2020).''
 Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
-In addition, they both state that the approach was then later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
+In addition, they both state that the approach was later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
 
-There is complete concordance between the deep sequencing accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the biological samples described by \citet{Wang2020medRxiv}.
+There is complete concordance between the accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the biological samples described by \citet{wang2020medRxiv}.
 There are 89 accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
 The number of accessions is approximately twice the number of early epidemic samples because each sample has data for two nanopore sequencing runtimes except one sample (B5) with just one runtime.
- There are 31 sequencing accessions corresponding to the 16 samples collected in February from Renmin Hospital patients, with these samples named R01, R02, etc.
- Again, all but one sample (R04) has data for two different sequencing runtimes.
+ There are 31 accessions corresponding to the 16 samples collected in February from Renmin Hospital patients, with these samples named R01, R02, etc.
+ Again, all but one sample (R04) have data for two different sequencing runtimes.
  In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for development and benchmarking of the approach.
@@
 
-Neither the pre-print~\citep{Wang2020medRxiv} or published manuscript~\citep{Wang2020small} contain any corrections or notes that indicate a scientific reason for deleting the study's sequencing data from the SRA.
+Neither the pre-print~\citep{wang2020medRxiv} or published manuscript~\citep{wang2020small} contain any corrections or notes that indicate a scientific reason for deleting the study's sequencing data from the SRA.
 
@@
 As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
-In addition, none of the sequencing runs could be accessed using the command-line tools of the SRA Toolkit.
+In addition, none of the runs could be accessed using the command-line tools of the SRA Toolkit.
 For instance, as of June 3, 2020, running the commands \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} both returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313395'``.
@@
 However, the SRA has begun storing all data on the Google and Amazon clouds.
-While inspecting the SRA's web interface for other sequencing accessions, I noticed that SRA files are often available from links to the cloud such as \texttt{https://storage.googleapis.com/nih-sequence-read-archive/run/<ACCESSION>/<ACCESSION>} where \texttt{<ACCESSION>}.
+While inspecting the SRA's web interface for other sequencing accessions, I noticed that SRA files are often available from links to the cloud such as \texttt{https://storage.googleapis.com/nih-sequence-read-archive/run/<ACCESSION>/<ACCESSION>}.
 
@@
 
-I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 of 45 early epidemic samples that were SARS-CoV-2 positive, as well as the 16 hospital samples from February (SRA files for SRR11313490 and SRR11313499 were not accessible via the cloud).
+I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February (files for SRR11313490 and SRR11313499 were not accessible via the cloud).
 These SRA files are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
@@
 For all files, the object timestamp is February 15, 2020, and the time is March 16, 2020.
-Although the SRA Toolkit does not clearly document the definition of these two properties, my \emph{guess} is that the object timestamp may refer to when the SRA file was created from a FASTQ file uploaded to the SRA, and the time may refer to when the accession was made public.
+Although the SRA Toolkit does not clearly document these two properties, my \emph{guess} is that the object timestamp may refer to when the SRA file was created from a FASTQ file uploaded to the SRA, and the time may refer to when the accession was made public.
 
 \subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for many samples}
-I aligned the recovered deep sequencing data to the SARS-CoV-2 genome, pooling the data when there were multiple accessions for a single biological sample (this was usually the case as most samples were sequenced using two runtimes as described above).
-Figure~\ref{fig:coverage} shows the sequencing coverage over the genome for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples.
+The approach described by \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
+The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
+In this paper, I focus on the region from 21,563 to 29,674 because this is a much longer contiguous sequence and because the amplicons were generated by a single rather than nested PCR, which reduces the risk of PCR artifacts.
 
@@
 \end{figure*}
+
+I aligned the recovered deep sequencing data to the SARS-CoV-2 genome, combining accessions for the same biological sample.
+Figure~\ref{fig:coverage} plots the sequencing coverage over the region of interest for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples; a comparable plot for the whole genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
 
@@
 }
-\label{suppfigfig:coverage_all}
+\label{suppfig:coverage_all}
 \end{figure}

### 2021-06-05 scripts/early_seq_subs.py
better filter early seqs and compute frac coverage

@@ -25,6 +25,7 @@ for addtl_prop in ['n_gapped_to_ref',
                    'n_ident_to_ref',
                    'n_subs_to_ref',
                    'n_ambiguous_to_ref',
+                   'frac_called',
                    'frac_called_in_region_of_interest',
                    'substitutions',
                    ]:
@@ -40,6 +41,9 @@ for seq in Bio.SeqIO.parse(snakemake.input.alignment, 'fasta'):
     else:
         assert len(seq) == length
         seqstr = str(seq.seq).upper()
+        data['frac_called'].append(
+                sum(nt in {'A', 'C', 'G', 'T'} for nt in seqstr)
+                / length)
         data['frac_called_in_region_of_interest'].append(
                 sum(nt in {'A', 'C', 'G', 'T'} for nt in
                     seqstr[start - 1: end]) / (end - start + 1))

### 2021-06-05 scripts/early_seq_subs.py
ignore muts at beginning or end of SARS-CoV-2 genome

@@ -31,6 +31,10 @@ for addtl_prop in ['n_gapped_to_ref',
                    ]:
     data[addtl_prop] = []
 
+ignore_muts_before = snakemake.params.ignore_muts_before
+ignore_muts_after = snakemake.params.ignore_muts_after
+print(f"Ignoring substitutions before {ignore_muts_before} or after {ignore_muts_after}")
+
 for seq in Bio.SeqIO.parse(snakemake.input.alignment, 'fasta'):
     if iseq == 0:
         if seq.id == ref_genome.id and (str(seq.seq).upper() ==
@@ -54,12 +58,10 @@ for seq in Bio.SeqIO.parse(snakemake.input.alignment, 'fasta'):
                                               for nt in seqstr))
         subs = [f"{ref_nt}{site}{seq_nt}" for site, (ref_nt, seq_nt)
                 in enumerate(zip(ref_seqstr, seqstr), start=1)
-                if ref_nt != seq_nt and seq_nt in ['A', 'C', 'G', 'T']]
+                if ref_nt != seq_nt and seq_nt in ['A', 'C', 'G', 'T']
+                and (ignore_muts_before <= site <= ignore_muts_after)
+                ]
         data['n_subs_to_ref'].append(len(subs))
-        assert length == sum(data[key][-1] for key in ['n_gapped_to_ref',
-                                                       'n_ident_to_ref',
-                                                       'n_subs_to_ref',
-                                                       'n_ambiguous_to_ref'])
         data['substitutions'].append(','.join(subs))
         head_entries = seq.description.split(' ', 1)[1]
         for entry in head_entries.split(', '):

### 2021-06-06 paper/paper.tex
narrow region of interest

@@
 \subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for many samples}
-The approach described by \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
+The approach described by \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites {\color{red} FIX REGION} 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
 The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.

### 2021-06-08 scripts/add_outgroup_to_alignment.py
build alignments with outgroups

@@ -1,0 +1,37 @@
+"""Implements ``snakemake`` rule `add_outgroup_to_alignment`."""
+
+
+import Bio.SeqIO
+
+import pandas as pd
+
+
+outgroup = snakemake.wildcards.outgroup
+start = snakemake.params.region['start']
+end = snakemake.params.region['end']
+assert start <= end
+
+print(f"Getting outgroup {outgroup} from {start} to {end} from {snakemake.input.comparator_map}")
+comparator_map = (
+    pd.read_csv(snakemake.input.comparator_map)
+    .set_index('site')
+    [outgroup]
+    .to_dict()
+    )
+outgroup_seq = ''.join(comparator_map[site] for site in range(start, end + 1))
+outgroup_seqrecord = Bio.SeqRecord.SeqRecord(seq=Bio.Seq.Seq(outgroup_seq),
+                                             id=outgroup,
+                                             name='',
+                                             description='',
+                                             )
+
+alignment = [outgroup_seqrecord]
+print(f"Reading alignment without outgroup from {snakemake.input.alignment}")
+for seq in Bio.SeqIO.parse(snakemake.input.alignment, 'fasta'):
+    assert outgroup != seq.id
+    alignment.append(seq)
+    assert len(outgroup_seqrecord) == len(alignment[-1]) == end - start + 1, (
+            f"{len(outgroup_seqrecord)=}\n{len(alignment[-1])=}\n{end - start + 1=}")
+
+print(f"Writing the {len(alignment)} sequences including outgroup to {snakemake.output.alignment}")
+Bio.SeqIO.write(alignment, snakemake.output.alignment, 'fasta')

### 2021-06-14 paper/paper.tex
progress in paper: up through table of substitutions

--- 2021-06-05 paper/paper.tex
@@
 
-\title{Recovery of a deleted deep sequencing data set sheds new light on the early spread of SARS-CoV-2 in Wuhan}
+\title{Recovery of a deleted SARS-CoV-2 deep sequencing data set from the early epidemic in Wuhan}
 
@@
 During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed SARS-CoV-2 deep sequencing data from the Sequence Read Archive (SRA), which is a repository maintained by the NIH's National Center for Biotechnology Information (NCBI).
-The first supplementary table of \citet{farkas2020insights} lists all SARS-CoV-2 deep sequencing data sets available from the SRA as of March 30, 2020.
-I have digitally archived a copy of this table in the Wayback Machine at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
+The first supplementary table of \citet{farkas2020insights} lists all SARS-CoV-2 deep sequencing data available from the SRA as of March 30, 2020.
 
-The majority the entries in this table refer to a deep sequencing project (BioProject PRJNA612766) by Wuhan University that is described as nanopore sequencing of SARS-CoV-2 amplicons.
+The majority of entries in this table refer to a project (BioProject PRJNA612766) by Wuhan University that is described as nanopore sequencing of SARS-CoV-2 amplicons.
 The table indicates this project represents 241 of the 282 SARS-CoV-2 sequencing run accessions in the SRA as of March, 30, 2020.
@@
 Searching for ``PRJNA612766'' in the NCBI's SRA search box returned a message of ``No items found.''
-I then searched for individual sequencing run accessions in the NCBI's SRA search box.
-These searches returned messages indicating that sequencing runs had been removed (Figure~\ref{fig:acc_removed}).
+I then searched for individual sequencing run accessions from the project in the NCBI's SRA search box.
+These searches returned messages indicating that the sequencing runs had been removed (Figure~\ref{fig:acc_removed}).
 
@@
 \caption{Accessions from deep sequencing project PRJNA612766 have been removed from the SRA.
-Shown is the result of searching for ``SRR11313485'' to reach the \url{https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
+Shown is the result of searching for ``SRR11313485.''
 A copy of this result has been digitally archived on the Wayback Machine at \url{https://web.archive.org/web/20210502131630/https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
@@
 The SRA is designed as a permanent archive of deep sequencing data.
-The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff with the reason for requesting the deletion~\citep{SRA_deletion}.
-An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows e-mails between NCBI staff and the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation}.
+The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
+An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
 
@@
 
-Taken together, these facts indicate that subsequent to March 30, 2020, someone must have e-mailed the NCBI to request deletion of the SARS-CoV-2 deep sequencing project PRJNA612766.
+Taken together, these facts indicate that subsequent to March 30, 2020, someone must have e-mailed the SRA to request deletion of SARS-CoV-2 deep sequencing project PRJNA612766.
 
@@
 The study describes an approach to diagnose infection with SARS-CoV-2 and other respiratory viruses by nanopore sequencing.
-This approach involved reverse-transcription of total RNA from swab samples, followed by PCR with specific primers that generated amplicons covering portions of the viral genome.
-These amplicons were then sequenced on an Oxford Nanopore GridION, and infection with SARS-CoV-2 was diagnosed if the sequencing yielded a sufficient number of reads aligning to the SARS-CoV-2 genome.
+This approach involved reverse-transcription of total RNA from swab samples, followed by PCR with specific primers to generate amplicons covering portions of the viral genome.
+These amplicons were then sequenced on an Oxford Nanopore GridION, and infection was diagnosed if the sequencing yielded sufficient reads aligning to the viral genome.
 Importantly, the study notes that this approach yields information about the sequence of the virus as well enabling diagnosis of infection.
@@
 Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
-In addition, they both state that the approach was later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
+In addition, both state that the approach was later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
 
-There is complete concordance between the accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the biological samples described by \citet{wang2020medRxiv}.
+There is complete concordance between the accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the samples described by \citet{wang2020medRxiv}.
 There are 89 accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
@@
 
-Neither the pre-print~\citep{wang2020medRxiv} or published manuscript~\citep{wang2020small} contain any corrections or notes that indicate a scientific reason for deleting the study's sequencing data from the SRA.
+Neither the pre-print~\citep{wang2020medRxiv} nor published manuscript~\citep{wang2020small} contain any correction or note that indicates a scientific reason for deleting the study's sequencing data from the SRA.
+I e-mailed both corresponding authors of \citet{wang2020medRxiv} to ask why they had deleted the deep sequencing data and to request details on the collection dates of the early outpatient samples, but received no reply.
 
@@
 In addition, none of the runs could be accessed using the command-line tools of the SRA Toolkit.
-For instance, as of June 3, 2020, running the commands \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} both returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313395'``.
+For instance, as of June 3, 2020, running \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313485'``.
 
@@
 This strategy was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
-I have archived this URL on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
+I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
 
 I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February (files for SRR11313490 and SRR11313499 were not accessible via the cloud).
-These SRA files are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
 I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for each SRA file.
@@
 
-\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for many samples}
+\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
 The approach described by \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
 The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
-In this paper, I focus on the region from 21,563 to 29,674 because this is a much longer contiguous sequence and because the amplicons were generated by a single rather than nested PCR, which reduces the risk of PCR artifacts.
+In this paper, analyze the region covering spike through ORF10 because this is a much longer contiguous sequence and because the amplicons were generated by a single rather than nested PCR, which reduces the risk of PCR artifacts.
+I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples did not have good coverage at the very termini of the region.
 
@@
 \centering
-\includegraphics[width=\linewidth]{../results/pileup/coverage_region.pdf}
-\caption{Sequencing depth over the SARS-CoV-2 genome from nucleotide site 21,563 to 29,674 for the 34 virus-positive early outbreak samples and the 16 samples from hospitalized patients in February.
+\includegraphics[width=\linewidth]{figures/coverage_region.pdf}
+\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early outbreak samples and the 16 samples from hospitalized patients in February.
 Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
@@
 Note that the y-axis uses a symlog scale.
-An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see statistics is at {\color{red} add link}.
+An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
 A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
@@
 
-I aligned the recovered deep sequencing data to the SARS-CoV-2 genome, combining accessions for the same biological sample.
-Figure~\ref{fig:coverage} plots the sequencing coverage over the region of interest for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples; a comparable plot for the whole genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
+I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same biological sample.
+Figure~\ref{fig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
+
+\begin{table*}[]
+{\small
+\begin{tabular}{lrll}
+\toprule
+sample &  fraction sites called (21570-29550) &           patient group &                                        substitutions relative to proCoV2 \\
+\midrule
+A4 &                               0.9827 &        early outpatient &                                                                          none \\
+C1 &                               0.9966 &        early outpatient &  G22081A (A=924, C=4, G=9), C28144T (C=6, T=1185), T29483G (C=1, G=45, T=1) \\
+C2 &                               0.9962 &        early outpatient &                                                 C29095T (C=1, G=1, T=751)  \\
+C9 &                               0.9536 &        early outpatient &                                  C28144T (C=3, T=823), G28514T (G=1, T=36)  \\
+D9 &                               0.9585 &        early outpatient &                                                     C28144T (C=4, T=1653)  \\
+D12 &                               0.9970 &        early outpatient &                                                     C28144T (C=8, T=2400)  \\
+E1 &                               0.9759 &        early outpatient &                                                           C28144T (T=125)  \\
+E5 &                               0.9758 &        early outpatient &              C24034T (A=5, C=3, T=74), T26729C (C=12), G28077C (C=142, G=4)  \\
+E11 &                               0.9877 &        early outpatient &                                 C25460T (C=2, T=246), C28144T (C=1, T=412)  \\
+F11 &                               0.9594 &        early outpatient &                             T25304A (A=9, T=1), C28144T (C=6, G=1, T=1328)  \\
+G1 &                               0.9959 &        early outpatient &                                   none                                        \\
+G11 &                               0.9677 &        early outpatient &                                  none                                         \\
+H9 &                               0.9941 &        early outpatient &                                                     C28144T (C=2, T=1254)  \\
+R11 &                               0.9987 &  hospital patient (Feb) &                               C21707T (T=401), C28144T (A=1, C=18, T=4265)  \\
+\bottomrule
+\end{tabular}
+}
+\caption{Samples from deleted project PRJNA612766 for which the SARS-CoV-2 sequence could be called at $\ge$95\% of sites between 21,570 and 29,550 and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
+Numbers in parentheses after each substitution give the deep sequencing reads with each nucleotide identity.
+}
+\end{table*}
+
+more text
+
+\section{Methods}
+
+\subsection{Code and data availability}
+The deleted SRA files recovered from the Google cloud are all available at  \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
+I have suffixed the file extension \texttt{.sra} to all of these files.
+
+\subsection{Archiving of key weblinks}
+I have digitally archived key weblinks in the Wayback Machine.
+These include:
+\begin{itemize}
+\item The first supplementary table of \citet{farkas2020insights}, which lists all SARS-CoV-2 accessions in the SRA as of March 30, 2020, is archived at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
+\end{itemize}
 
@@
 \centering
-\includegraphics[width=\linewidth]{../results/pileup/coverage_all.pdf}
+\includegraphics[width=\linewidth]{figures/coverage_all.pdf}
 \caption{A version of Figure~\ref{fig:coverage} that shows coverage over the full length of the SARS-CoV-2 genome.

### 2021-06-14 paper/paper.tex
added deltadist figurs

@@
 
-\title{Recovery of a deleted SARS-CoV-2 deep sequencing data set from the early epidemic in Wuhan}
+\title{Recovery of a deleted SARS-CoV-2 deep sequencing data set from the early Wuhan epidemic}
 
@@
 The table indicates this project represents 241 of the 282 SARS-CoV-2 sequencing run accessions in the SRA as of March, 30, 2020.
-Because I had never encountered any other mention of this sequencing project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
+Because I had never encountered any other mention of this project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
 Searching for ``PRJNA612766'' in the NCBI's SRA search box returned a message of ``No items found.''
@@
 \caption{Accessions from deep sequencing project PRJNA612766 have been removed from the SRA.
-Shown is the result of searching for ``SRR11313485.''
-A copy of this result has been digitally archived on the Wayback Machine at \url{https://web.archive.org/web/20210502131630/https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
+Shown is the result of searching for ``SRR11313485'' in the SRA search toolbar.
+This result has been digitally archived on the Wayback Machine at \url{https://web.archive.org/web/20210502131630/https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
 }%
@@
 An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
+Therefore, subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.
 
@@
 \caption{Example of the process to delete SRA data.
-The image shows excerpted e-mails between the lead author of the pangolin coronavirus paper~\citep{xiao2020isolation} and SRA staff.
+The image shows excerpted e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff.
 The full e-mails are at \url{https://usrtk.org/tag/pangolin-papers/}.
@@
 \end{figure}
-
-Taken together, these facts indicate that subsequent to March 30, 2020, someone must have e-mailed the SRA to request deletion of SARS-CoV-2 deep sequencing project PRJNA612766.
 
@@
 There are 89 accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
-The number of accessions is approximately twice the number of early epidemic samples because each sample has data for two nanopore sequencing runtimes except one sample (B5) with just one runtime.
+The number of accessions is approximately twice the number of early epidemic samples because each sample has data for two sequencing runtimes except one sample (B5) with just one runtime.
  There are 31 accessions corresponding to the 16 samples collected in February from Renmin Hospital patients, with these samples named R01, R02, etc.
- Again, all but one sample (R04) have data for two different sequencing runtimes.
- In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for development and benchmarking of the approach.
+ Again, all but one sample (R04) have data for two sequencing runtimes.
+ In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for benchmarking of the approach.
  Together, these samples and controls account for all 241 accessions listed for PRJNA612766 in the supplementary table of \citet{farkas2020insights}.
@@
 The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
-In this paper, analyze the region covering spike through ORF10 because this is a much longer contiguous sequence and because the amplicons were generated by a single rather than nested PCR, which reduces the risk of PCR artifacts.
-I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples did not have good coverage at the very termini of the region.
+In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by a single rather than nested PCR, which reduces the risk of PCR artifacts.
+I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the very termini of the region.
 
@@
 
-I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same biological sample.
+I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
 Figure~\ref{fig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
@@
 }
-\caption{Samples from deleted project PRJNA612766 for which the SARS-CoV-2 sequence could be called at $\ge$95\% of sites between 21,570 and 29,550 and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
+\caption{Samples for which the SARS-CoV-2 sequence could be called at $\ge$95\% of sites between 21,570 and 29,550, and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
 Numbers in parentheses after each substitution give the deep sequencing reads with each nucleotide identity.
+\label{tab:mutations}
 }
@@
 
-more text
+I called the sequence identity at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
+With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the nucleotide sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
+Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely using Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
+Although requiring a coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations are supported by at least 10 reads.
+In addition, the mutations called from the raw sequences in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.
+
+\subsection{Phylogenetic analysis of existing SARS-CoV-2 sequences from January of 2020 or earlier}
+To contextualize the sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
+
+Numerous evolutionary analyses have shown that known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
+However, attempts to infer the identity of the SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the sequences of the earliest reported isolates from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
+\caption{The reported collection date of each SARS-CoV-2 sequence versus its relative mutational distance from the RaTG13 bat coronavirus outgroup.
+Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
+The plot shows all sequences in GISAID collected no later than February 28, 2020, after performing the quality-control described in the Methods.
+Sequences that the joint WHO-China report~\cite{CITE} describes as being associated with the Wuhan Seafood Market are plotted with squares.
+Points are slightly jittered on the y-axis.
+Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
+Static versions of the plot with RpYN06 and RmYN02 outgroups are in Supplementary Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
+}
+\label{fig:deltadist_RaTG13}
+\end{figure*}
 
@@
 
+\begin{figure}[h!]
+{\bf \LARGE A} \\
+\centerline{\includegraphics[width=\linewidth]{figures/deltadist_RpYN06.pdf}}
+{\bf \LARGE B} \\
+\centerline{\includegraphics[width=\linewidth]{figures/deltadist_RmYN02.pdf}}
+\caption{Versions of Figure~\ref{fig:deltadist_RaTG13} but calculating the relative mutational distances using an outgroup of (A) RpYN06 or (B) RmYN02.
+}
+\label{suppfig:deltadist_RpYN06_RmYN02}
+\end{figure}
+
 \end{document}

### 2021-06-15 paper/paper.tex
latest paper text updates

@@
 However, attempts to infer the identity of the SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the sequences of the earliest reported isolates from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
+Although uncertainty remains about whether the proximal origin of SARS-CoV-2 was a zoonosis~\cite{CITE} or a lab accident~\cite{CITE}, all reasonable explanations agree that at a deeper level SARS-CoV-2 genome is largely derived from one or more bat coronaviruses~\cite{CITE}.
+One would therefore expect that if SARS-CoV-2 is ultimately derived from a bat coronavirus
+This fact is illustrated in Figure~\ref{fig:deltadist_RaTG13}.
 
@@
 The plot shows all sequences in GISAID collected no later than February 28, 2020, after performing the quality-control described in the Methods.
-Sequences that the joint WHO-China report~\cite{CITE} describes as being associated with the Wuhan Seafood Market are plotted with squares.
+Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
 Points are slightly jittered on the y-axis.

### 2021-06-15 paper/paper.tex
more writing of Results section existing sequences

--- 2021-06-14 paper/paper.tex
@@
 An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
-Therefore, subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.
+Subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.
 
@@
 Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
-In addition, both state that the approach was later applied to 16 additional samples collected on February 11 and 12 of 2020 from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
+In addition, both state that the approach was later applied to 16 additional samples collected on February 11--12, 2020, from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
 
@@
 \includegraphics[width=\linewidth]{figures/coverage_region.pdf}
-\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early outbreak samples and the 16 samples from hospitalized patients in February.
+\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early epidemic samples and the 16 samples from hospitalized patients in February.
 Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
@@
 
-I called the sequence identity at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
-With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the nucleotide sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
+I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
+With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
 Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely using Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
-Although requiring a coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations are supported by at least 10 reads.
-In addition, the mutations called from the raw sequences in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.
+Although requiring coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations have coverage $\ge$10.
+In addition, the mutations I called from the raw sequence data in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.
 
-\subsection{Phylogenetic analysis of existing SARS-CoV-2 sequences from January of 2020 or earlier}
-To contextualize the sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
+\subsection{Analysis of existing sequences emphasizes the perplexing discordance between collection date and distance to SARS-CoV-2's bat coronavirus relatives}
+To contextualize the viral sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
 
-Numerous evolutionary analyses have shown that known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
-However, attempts to infer the identity of the SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the sequences of the earliest reported isolates from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
+Evolutionary analyses have shown that known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
+However, attempts to infer this SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
+This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
+One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to the virus's bat coronavirus relatives---but this is not the case.
 
@@
 \end{figure*}
+
+This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest overall identity to SARS-CoV-2.
+The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than SARS-CoV-2 sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
+The discrepancy is especially pronounced for early Wuhan sequences collected from patients who had visited the Huanan Seafood Market, which was widely reported as an initial location of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
+All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original site of spread of a bat coronavirus into humans.
+Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Supplementary Figure~S\ref{suppfig:deltadist_RpYN06_RmYN02}).
+
+This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is chosen to be the one closest to the bat coronavirus outgroup.
+If we limit the analysis to sequences with at least two observation among strains collected no later than January 2020, there are three possible ways to root the tree 
+ (Figure~\ref{fig:tree_RaTG13} shows these rootings using a RaTG13 outgroup; Supplementary Figure~S\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained using RpYN06 or RmYN02 outgroups).
 

### 2021-06-15 paper/paper.tex
add tree figure

@@
 \usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
+\usepackage[export]{adjustbox}
 
@@
 An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
-A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
+A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Figure~\ref{suppfig:coverage_all}.
 }
@@
 I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
-Figure~\ref{fig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Supplementary Figure~S\ref{suppfig:coverage_all}.
+Figure~\ref{fig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
 
@@
 Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
-Static versions of the plot with RpYN06 and RmYN02 outgroups are in Supplementary Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
+Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
 }
@@
 The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than SARS-CoV-2 sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
-The discrepancy is especially pronounced for early Wuhan sequences collected from patients who had visited the Huanan Seafood Market, which was widely reported as an initial location of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
+The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
 All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original site of spread of a bat coronavirus into humans.
-Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Supplementary Figure~S\ref{suppfig:deltadist_RpYN06_RmYN02}).
+Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
 
-This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is chosen to be the one closest to the bat coronavirus outgroup.
-If we limit the analysis to sequences with at least two observation among strains collected no later than January 2020, there are three possible ways to root the tree 
- (Figure~\ref{fig:tree_RaTG13} shows these rootings using a RaTG13 outgroup; Supplementary Figure~S\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained using RpYN06 or RmYN02 outgroups).
+This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is chosen to be closest to the bat coronavirus outgroup.
+If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three possible ways to root the tree in this fashion (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
+
+ \begin{figure*}
+ \centerline{
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RaTG13_without_deleted_seqs.pdf}
+ \hspace{0.02\linewidth}
+ \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RaTG13_without_deleted_seqs.pdf}
+  \hspace{0.02\linewidth}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RaTG13_without_deleted_seqs.pdf}
+ }
+ \caption{
+ Phylogenetic trees of all SARS-CoV-2 sequences with multiple observations among viruses collected no later than January 31, 2020.
+ The three trees are identical except they are rooted to make the putative progenitor each of the three sequences with the highest identity to the RaTG13 bat coronavirus outgroup.
+The nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected (see key at bottom).
+The mutations on each branch are labeled, with mutations towards the nucleotide identity in the RaTG13 outgroup in purple.
+The labels at the top of each tree show the first known virus identical to each putative progenitor, as well as the mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and the widely used Wuhan-Hu-1 reference sequence.
+All sequences in the monophyletic group that contains C28144T are collapsed into a single clade labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; note that this clade contains Wuhan-Hu-1.
+ Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is instead RpYN06 or RmYN02.
+\label{fig:tree_RaTG13}
+ }
+ \end{figure*}
 
@@
 \setcounter{page}{1}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{1}
 
@@
 
+ \begin{figure}[h!]
+ {\bf \LARGE A} \\
+ \centerline{
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RpYN06_without_deleted_seqs.pdf}
+ \hspace{0.04\linewidth}
+ \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RpYN06_without_deleted_seqs.pdf}
+  \hspace{0.04\linewidth}
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RpYN06_without_deleted_seqs.pdf}
+ }
+ \\
+  {\bf \LARGE B} \\
+ \centerline{
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RmYN02_without_deleted_seqs.pdf}
+ \hspace{0.04\linewidth}
+ \includegraphics[width=0.32\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RmYN02_without_deleted_seqs.pdf}
+  \hspace{0.04\linewidth}
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RmYN02_without_deleted_seqs.pdf}
+ }
+ \caption{
+ A version of Figure~\ref{fig:tree_RaTG13} but rooting using an outgroup of (A) RpYN06 or (B) RmYN02 outgroups.
+ The tree topologies are identical to those obtained using RaTG13 in Figure~\ref{fig:tree_RaTG13}, with the only differences being a few minor changes in which mutations on branches are towards the outgroup (purple mutation labels).
+\label{suppfig:tree_RpYN06_RmYN02}
+ }
+ \end{figure}
+
 \end{document}

### 2021-06-15 paper/paper.tex
add trees to results

@@
 Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
-The plot shows all sequences in GISAID collected no later than February 28, 2020, after performing the quality-control described in the Methods.
+The plot shows sequences in GISAID collected no later than February 28, 2020.
 Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
@@
 Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
-
-This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is chosen to be closest to the bat coronavirus outgroup.
-If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three possible ways to root the tree in this fashion (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
 
@@
  \end{figure*}
+
+This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
+If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
+Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
+Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and the epidemiological data suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
+
+Tables 6 and 7 of WHO-China report indicate viruses from Huanan Market all in clade B except S13, which has T8782, T28144, and some mutation at 4946.
+However, clades are connected by single mutations, for instance Sichuan/SC-MY-054/2020 has T8782 but C28144.
 

### 2021-06-16 paper/paper.tex
explanations of rooting discrepancy

@@
 
-\title{Recovery of a deleted SARS-CoV-2 deep sequencing data set from the early Wuhan epidemic}
+\title{Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic}
 
@@
 This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
-One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to the virus's bat coronavirus relatives---but this is not the case.
+One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to these bat coronavirus relatives---but this is not the case.
 
@@
 
-This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest overall identity to SARS-CoV-2.
-The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than SARS-CoV-2 sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
+This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
+The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
 The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
@@
 Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
-Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and the epidemiological data suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
+Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological data suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
+
+Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence indicating that Wuhan was the origin of the pandemic.
+\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
+\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in phylogenetic substitution models.
+\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
+Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref[fig:tree_RaTG13} of simply placing the root at the three possible locations that are closest to the bat outgroup.
+However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} or my more simplistic approach in Figure~\ref[fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
 

### 2021-06-16 paper/paper.tex
start discussion of less plausible explanations

@@
 If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
-Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
-Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological data suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
+Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which is reported to have been collected on January 26, 2020).
+Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological evidence suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
 
-Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence indicating that Wuhan was the origin of the pandemic.
+Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
 \citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
@@
 \citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
-Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref[fig:tree_RaTG13} of simply placing the root at the three possible locations that are closest to the bat outgroup.
-However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} or my more simplistic approach in Figure~\ref[fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
+Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the three possible locations closest to the bat outgroup.
+However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
+
+Before moving onto the next section, I will also briefly address several less plausible explanations for the discordance that have gained traction in popular discussion of SARS-CoV-2's origins.
+The first explanation suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
+This explanation can be discounted, because although there are some unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the fundamental conundrum about the progenitor described above holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before the emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
 
@@
 However, clades are connected by single mutations, for instance Sichuan/SC-MY-054/2020 has T8782 but C28144.
+
+\subsection{Sequences recovered from the deleted project help reconcile inferences about SARS-CoV-2's progenitor}
 

### 2021-06-16 paper/paper.tex
finished discussion on analysis of existing sequences

@@
 The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
-The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a site of potential zoonosis~\citep{WHO2021origins}.
-All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original site of spread of a bat coronavirus into humans.
+The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
+All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
 Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
@@
  The three trees are identical except they are rooted to make the putative progenitor each of the three sequences with the highest identity to the RaTG13 bat coronavirus outgroup.
-The nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected (see key at bottom).
+Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected (see key at bottom).
 The mutations on each branch are labeled, with mutations towards the nucleotide identity in the RaTG13 outgroup in purple.
-The labels at the top of each tree show the first known virus identical to each putative progenitor, as well as the mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and the widely used Wuhan-Hu-1 reference sequence.
-All sequences in the monophyletic group that contains C28144T are collapsed into a single clade labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; note that this clade contains Wuhan-Hu-1.
+The labels at the top of each tree show the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and the Wuhan-Hu-1 reference sequence.
+All sequences in the monophyletic group containing C28144T are collapsed into a single node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
  Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is instead RpYN06 or RmYN02.
@@
 \citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
-Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the three possible locations closest to the bat outgroup.
+Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the bat outgroup.
 However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
 
-Before moving onto the next section, I will also briefly address several less plausible explanations for the discordance that have gained traction in popular discussion of SARS-CoV-2's origins.
-The first explanation suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
-This explanation can be discounted, because although there are some unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the fundamental conundrum about the progenitor described above holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before the emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
-
-Tables 6 and 7 of WHO-China report indicate viruses from Huanan Market all in clade B except S13, which has T8782, T28144, and some mutation at 4946.
-However, clades are connected by single mutations, for instance Sichuan/SC-MY-054/2020 has T8782 but C28144.
+Before moving onto the next section, I will also briefly address two less plausible explanations for the discordance between the phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
+The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
+This explanation can be discounted, because although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
+The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other unknown market being the source of viruses that lack the T8782C and C28144T mutations.
+However, inspection of the phylogenetic trees in Figure~\ref{fig:tree_RaTG13} shows why this idea can be discounted: clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but then also the exceedingly improbable evolution of one of these progenitors into the other after it had jumped to humans.
 

### 2021-06-16 paper/paper.tex
added figures with deleted seqs

@@
 \includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
-\caption{The reported collection date of each SARS-CoV-2 sequence versus its relative mutational distance from the RaTG13 bat coronavirus outgroup.
+\caption{The reported collection dates of SARS-CoV-2 sequences in GISAID versus their relative mutational distances from the RaTG13 bat coronavirus outgroup.
 Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
@@
  \caption{
- Phylogenetic trees of all SARS-CoV-2 sequences with multiple observations among viruses collected no later than January 31, 2020.
+ Phylogenetic trees of all existing SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected no later than January 31, 2020.
  The three trees are identical except they are rooted to make the putative progenitor each of the three sequences with the highest identity to the RaTG13 bat coronavirus outgroup.
@@
 \subsection{Sequences recovered from the deleted project help reconcile inferences about SARS-CoV-2's progenitor}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\linewidth]{figures/deltadist_jitter.png}
+\caption{
+CAPTION
+\label{fig:deltadist_jitter}
+}
+\end{figure}
+
+ \begin{figure*}
+ \centerline{
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RaTG13_with_deleted_seqs.pdf}
+ \hspace{0.02\linewidth}
+ \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RaTG13_with_deleted_seqs.pdf}
+  \hspace{0.02\linewidth}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RaTG13_with_deleted_seqs.pdf}
+ }
+ \caption{
+CAPTION
+\label{fig:tree_with_deleted}
+ }
+ \end{figure*}
 

### 2021-06-16 paper/paper.tex
fix tree figure filenames

@@
  \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RaTG13_without_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_without_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
- \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RaTG13_without_deleted_seqs.pdf}
+ \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
   \hspace{0.02\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RaTG13_without_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_without_deleted_seqs.pdf}
  }
@@
  \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RaTG13_with_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_with_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
- \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RaTG13_with_deleted_seqs.pdf}
+ \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
   \hspace{0.02\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RaTG13_with_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_with_deleted_seqs.pdf}
  }
@@
  \centerline{
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RpYN06_without_deleted_seqs.pdf}
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RpYN06_without_deleted_seqs.pdf}
  \hspace{0.04\linewidth}
- \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RpYN06_without_deleted_seqs.pdf}
+ \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RpYN06_without_deleted_seqs.pdf}
   \hspace{0.04\linewidth}
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RpYN06_without_deleted_seqs.pdf}
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RpYN06_without_deleted_seqs.pdf}
  }
@@
  \centerline{
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%USA\%WA1\%2020_RmYN02_without_deleted_seqs.pdf}
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RmYN02_without_deleted_seqs.pdf}
  \hspace{0.04\linewidth}
- \includegraphics[width=0.32\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Guangdong\%HKU-SZ-002\%2020_RmYN02_without_deleted_seqs.pdf}
+ \includegraphics[width=0.32\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RmYN02_without_deleted_seqs.pdf}
   \hspace{0.04\linewidth}
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19\%Shandong\%LY005-2\%2020_RmYN02_without_deleted_seqs.pdf}
+ \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RmYN02_without_deleted_seqs.pdf}
  }

### 2021-06-16 paper/figures/tree_images/replace_percent.py
fix tree figure filenames

@@ -1,0 +1,8 @@
+import glob
+import os
+
+
+for f in glob.glob('*%*'):
+    new_f = f.replace('%', '-')
+    print(f"Renaming {f} to {new_f}")
+    os.rename(f, new_f)

### 2021-06-16 paper/paper.tex
draft hyperlinks to fix citation reference compiling
start describing addition of deleted seqs in Results

@@
 \PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
+\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references
 
@@
 \subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
-The approach described by \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
+The approach of \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
 The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
-In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by a single rather than nested PCR, which reduces the risk of PCR artifacts.
-I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the very termini of the region.
+In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by a single rather than nested PCR.
+I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the termini of the region.
 
@@
 
-\subsection{Analysis of existing sequences emphasizes the perplexing discordance between collection date and distance to SARS-CoV-2's bat coronavirus relatives}
+I also determined the consensus sequence of the plasmid control from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
+Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T, this fact indicates plasmid contamination did not afflict the viral samples in the deleted sequencing project.
+
+\subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
 To contextualize the viral sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
 
-Evolutionary analyses have shown that known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
+Known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
 However, attempts to infer this SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
@@
 Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
-\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as suggested by phylogenetic rooting.
-\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in phylogenetic substitution models.
+\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as implied by phylogenetic rooting.
+\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in substitution models.
 \citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
@@
 
-Before moving onto the next section, I will also briefly address two less plausible explanations for the discordance between the phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
+Before moving onto the next section, I will also briefly address two less plausible explanations for the discordance between phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
 The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
@@
 The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other unknown market being the source of viruses that lack the T8782C and C28144T mutations.
-However, inspection of the phylogenetic trees in Figure~\ref{fig:tree_RaTG13} shows why this idea can be discounted: clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but then also the exceedingly improbable evolution of one of these progenitors into the other after it had jumped to humans.
+However, inspection of the phylogenetic trees in Figure~\ref{fig:tree_RaTG13} shows this idea can also be discounted: clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but then also the exceedingly improbable evolution of one of these progenitors towards the other after it had jumped to humans.
 
-\subsection{Sequences recovered from the deleted project help reconcile inferences about SARS-CoV-2's progenitor}
+\subsection{Sequences recovered from the deleted project help resolve inferences about SARS-CoV-2's progenitor}
+To examine if the sequences recovered from the deleted data set help resolve the conundrum described in the previous section, I repeated the analyses including those sequences.
+In the process, I noted another salient fact: four GISAID sequences collected in Guangdong that fall in a putative progenitor node are from patients who traveled to Wuhan in late December of 2019 and developed symptoms before or on the day that they returned to Guangdong, where their viruses were ultimately sequenced~\citep{chan2020familial, kang2020evidence}.
+Since these patients were clearly infected in Wuhan even though they were sequenced in Guangdong, I annotated them separately from both the other Wuhan and other China sequences.
 
@@
 \caption{
-CAPTION
+Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by the sequences recovered from the deleted data set (sites 21,570--29,550).
+The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.
+Mutational distance is calculated relative to proCoV2, and points are slightly jittered on the y-axis.
+Go to {\color{red} add link} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of y-axis jittering.
 \label{fig:deltadist_jitter}
@@
  \caption{
-CAPTION
+Phylogenetic trees identical to those in Figure~\ref{fig:tree_RaTG13} except with the addition of the early Wuhan epidemic sequences from the deleted data set, and with Guangdong patients infected in Wuhan prior to January 5 annotated separately.
+Because the deleted data set only provides partial sequences, they cannot all be placed unambiguously on the tree.
+Therefore, they are added to each node with which they are compatible proportional to the number of sequences already in that node.
+The sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as key lineage-defining mutations occur in the sequenced region, but the sequences that lack either of these mutations can be placed in a large number of nodes including that of the proCoV2 putative progenitor.
+ Figure~\ref{suppfig:tree_RpYN06_RmYN02} has already demonstrated that the results are identical if RpYN06 or RmYN02 are instead used as the outgroup.
 \label{fig:tree_with_deleted}
@@
  \end{figure*}
+
+Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any of the sequences from the Huanan Seafood Market (Figure~\ref{fig:tree_with_deleted}).
+To visualize these sequences in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees rooted to put the putative progenitor at each of the three sequences with highest identity to the bat coronavirus outgroup (Figure~\ref{fig:tree_with_deleted}).
+
+It is immediately apparent that the discrepancy between outgroup rooting and epidemiological data is substantially alleviated by adding the deleted sequences and properly annotating the Wuhan infections that were sequenced in Guangdong.
 

### 2021-06-16 paper/paper.tex
finished draft of last Results section

@@
 
-Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any of the sequences from the Huanan Seafood Market (Figure~\ref{fig:tree_with_deleted}).
-To visualize these sequences in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees rooted to put the putative progenitor at each of the three sequences with highest identity to the bat coronavirus outgroup (Figure~\ref{fig:tree_with_deleted}).
+Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any sequences from the Huanan Seafood Market across the region in which all were fully sequenced (Figure~\ref{fig:deltadist_jitter}).
+To place this observation in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees described in the previous section (Figure~\ref{fig:tree_with_deleted}).
 
-It is immediately apparent that the discrepancy between outgroup rooting and epidemiological data is substantially alleviated by adding the deleted sequences and properly annotating the Wuhan infections that were sequenced in Guangdong.
+It is immediately apparent that the discrepancy between outgroup rooting and epidemiological evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong (Figure~\ref{fig:tree_with_deleted}).
+The middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
+The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
+In this scenario, the progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
+The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a putative progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
+Note that although three deleted early epidemic sequences are compatible with this proCoV2 progenitor, the defining C18060T mutation is not covered in the sequenced region---so further sequencing of these samples would be informative.
+The rightmost tree in Figure~\ref{fig:tree_with_deleted} looks less plausible, as it has almost no weight from Wuhan and the first sequence identical to its progenitor was not collected until January 24.
+
+We can also qualitatively evaluate the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the sample principle employed by \citet{worobey2020emergence} to help evaluate scenarios for the emergence of SARS-CoV-2 in Europe and North America: namely that during an exponentially growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
+This principle is especially likely to hold for the scenarios represented in Figure~\ref{fig:tree_with_deleted}, where the are multiple observed representatives of each putative progenitor which each have the potential to yield descendants with new mutations.
+Using this qualitative idea, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
+I acknowledge that these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in more qualitative reasoning when there are valid concerns about the nature of the underlying data.
 

### 2021-06-16 paper/paper.tex
some more edits to results

@@
 
-\begin{figure*}[]
-\centering
-\includegraphics[width=\linewidth]{figures/coverage_region.pdf}
-\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early epidemic samples and the 16 samples from hospitalized patients in February.
-Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
-The dashed gray line is the minimum coverage required to call a consensus identity at a site.
-Note that the y-axis uses a symlog scale.
-An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
-A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Figure~\ref{suppfig:coverage_all}.
-}
-\label{fig:coverage}
-\end{figure*}
-
 I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
-Figure~\ref{fig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
+Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
 
@@
 
-I also determined the consensus sequence of the plasmid control from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
-Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T, this fact indicates plasmid contamination did not afflict the viral samples in the deleted sequencing project.
+I also determined the consensus sequence of the plasmid control used by \citet{wang2020medRxiv} from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
+Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), this fact indicates plasmid contamination did not afflict the viral samples in the deleted sequencing project.
 
 \subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
-To contextualize the viral sequences recovered from the deleted project described above, here I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
+To contextualize the viral sequences recovered from the deleted project described above, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
+The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.
 
 Known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
-However, attempts to infer this SARS-CoV-2 progenitor sequence have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
+However, attempts to infer this progenitor have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
 This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
@@
 
+\section{Discussion}
+
+Something about lacking formal analyses of some other studies, but maybe this is not so bad.
+
 \section{Methods}
@@
 \section{Supplementary Material}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\linewidth]{figures/coverage_region.pdf}
+\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early epidemic samples and the 16 samples from hospitalized patients in February.
+Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
+The dashed gray line is the minimum coverage required to call a consensus identity at a site.
+Note that the y-axis uses a symlog scale.
+An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
+A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Figure~\ref{suppfig:coverage_all}.
+}
+\label{suppfig:coverage}
+\end{figure}
 

### 2021-06-17 paper/paper.tex
more edits to Results

@@
 \usepackage[export]{adjustbox}
+\usepackage{xurl}
 
-\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
-\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references
+%\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
+%\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references
 
@@
 The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
-An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail sent to SRA staff by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} to request the deletion of two sequencing runs.
+An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} requesting deletion of two sequencing runs.
 Subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.
@@
 \caption{Example of the process to delete SRA data.
-The image shows excerpted e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff.
-The full e-mails are at \url{https://usrtk.org/tag/pangolin-papers/}.
+The image shows e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff excerpted from \url{https://usrtk.org/tag/pangolin-papers/}.
 }
@@
 In addition, none of the runs could be accessed using the command-line tools of the SRA Toolkit.
-For instance, as of June 3, 2020, running \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} returned an error message stating ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313485'``.
+For instance, running \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} returned the message ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313485'``.
 
@@
 Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
-This strategy was successful; for instance, as of June 3, 2020, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
+This strategy was successful; for instance, as of June 3, 2021, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
 I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
@@
 The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
-In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by a single rather than nested PCR.
+In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
 I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the termini of the region.
@@
 \subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
-To contextualize the viral sequences recovered from the deleted project described above, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
+To contextualize the viral sequences recovered from the deleted project, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
 The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.
@@
 The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
-The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was widely reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
+The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
 All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
@@
 
-This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
-If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
-Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which is reported to have been collected on January 26, 2020).
+This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
+If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
+Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
 Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological evidence suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
@@
 \citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as implied by phylogenetic rooting.
-\citet{pipes2021assessing} provide a detailed discussion of the conundrum, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in substitution models.
+\citet{pipes2021assessing} discuss the conundrum in detail, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in substitution models.
 \citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
-Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the bat outgroup.
-However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach in Figure~\ref{fig:tree_RaTG13} explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
+Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the outgroup.
+However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
 
-Before moving onto the next section, I will also briefly address two less plausible explanations for the discordance between phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
+Before moving to the next section, I will also briefly address two less plausible explanations for the discordance between phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
 The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
-This explanation can be discounted, because although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
-The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other unknown market being the source of viruses that lack the T8782C and C28144T mutations.
-However, inspection of the phylogenetic trees in Figure~\ref{fig:tree_RaTG13} shows this idea can also be discounted: clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but then also the exceedingly improbable evolution of one of these progenitors towards the other after it had jumped to humans.
+But although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
+The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other market being the source of viruses that lack the T8782C and C28144T mutations.
+However, inspection of Figure~\ref{fig:tree_RaTG13} shows that clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but also the exceedingly improbable evolution of one of these progenitors towards the other after it had jumped to humans.
 
@@
 
-Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any sequences from the Huanan Seafood Market across the region in which all were fully sequenced (Figure~\ref{fig:deltadist_jitter}).
-To place this observation in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees described in the previous section (Figure~\ref{fig:tree_with_deleted}).
+Repeating the analysis of the previous section with these new sequences shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
+This fact suggests that the Huanan Seafood Market sequences, which are the primary focus of the genomic epidemiology analyses in the joint WHO-China report~\citep{WHO2021origins}, may not be representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
 
-It is immediately apparent that the discrepancy between outgroup rooting and epidemiological evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong (Figure~\ref{fig:tree_with_deleted}).
-The middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
+Furthermore, it is immediately apparent that the discrepancy between outgroup rooting and the evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong.
+The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
 The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
-In this scenario, the progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
-The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a putative progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
-Note that although three deleted early epidemic sequences are compatible with this proCoV2 progenitor, the defining C18060T mutation is not covered in the sequenced region---so further sequencing of these samples would be informative.
+This putative progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
+The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
+However, analysis of this rooting is limited by the fact that the defining C18060T mutation is in a region not covered in the deleted sequences.
 The rightmost tree in Figure~\ref{fig:tree_with_deleted} looks less plausible, as it has almost no weight from Wuhan and the first sequence identical to its progenitor was not collected until January 24.
 
-We can also qualitatively evaluate the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the sample principle employed by \citet{worobey2020emergence} to help evaluate scenarios for the emergence of SARS-CoV-2 in Europe and North America: namely that during an exponentially growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
-This principle is especially likely to hold for the scenarios represented in Figure~\ref{fig:tree_with_deleted}, where the are multiple observed representatives of each putative progenitor which each have the potential to yield descendants with new mutations.
-Using this qualitative idea, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
-I acknowledge that these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in more qualitative reasoning when there are valid concerns about the nature of the underlying data.
+We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
+This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
+Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
+I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the nature of the underlying data.
 

### 2021-06-17 paper/paper.tex
minor edits to Results

@@
  \caption{
- Phylogenetic trees of all existing SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected no later than January 31, 2020.
- The three trees are identical except they are rooted to make the putative progenitor each of the three sequences with the highest identity to the RaTG13 bat coronavirus outgroup.
-Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected (see key at bottom).
-The mutations on each branch are labeled, with mutations towards the nucleotide identity in the RaTG13 outgroup in purple.
-The labels at the top of each tree show the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and the Wuhan-Hu-1 reference sequence.
-All sequences in the monophyletic group containing C28144T are collapsed into a single node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is instead RpYN06 or RmYN02.
+ Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected no later than January 31, 2020.
+ The trees are identical except they are rooted to make the putative progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
+Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected.
+The mutations on each branch are labeled, with mutations towards the nucleotide identity in the outgroup in purple.
+The labels at the top of each tree give the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and Wuhan-Hu-1.
+The monophyletic group containing C28144T is collapsed into a single node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
+ Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is RpYN06 or RmYN02.
 \label{fig:tree_RaTG13}
@@
 Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
-Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and epidemiological evidence suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.
+Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with other evidence suggesting the progenitor is an early virus from Wuhan~\citep{pipes2021assessing}.
 
@@
 
-\subsection{Sequences recovered from the deleted project help resolve inferences about SARS-CoV-2's progenitor}
+\subsection{Sequences recovered from the deleted project and better annotation of Wuhan-derived viruses help reconcile inferences about SARS-CoV-2's progenitor}
 To examine if the sequences recovered from the deleted data set help resolve the conundrum described in the previous section, I repeated the analyses including those sequences.
@@
 \caption{
-Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by the sequences recovered from the deleted data set (sites 21,570--29,550).
+Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (sites 21,570--29,550).
 The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.
@@
  \caption{
-Phylogenetic trees identical to those in Figure~\ref{fig:tree_RaTG13} except with the addition of the early Wuhan epidemic sequences from the deleted data set, and with Guangdong patients infected in Wuhan prior to January 5 annotated separately.
-Because the deleted data set only provides partial sequences, they cannot all be placed unambiguously on the tree.
-Therefore, they are added to each node with which they are compatible proportional to the number of sequences already in that node.
-The sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as key lineage-defining mutations occur in the sequenced region, but the sequences that lack either of these mutations can be placed in a large number of nodes including that of the proCoV2 putative progenitor.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} has already demonstrated that the results are identical if RpYN06 or RmYN02 are instead used as the outgroup.
+Phylogenetic trees like those in Figure~\ref{fig:tree_RaTG13} with the addition of the early Wuhan epidemic sequences from the deleted data set, and Guangdong patients infected in Wuhan prior to January 5 annotated separately.
+Because the deleted sequences are partial, they cannot all be placed unambiguously on the tree.
+Therefore, they are added to each compatible node proportional to the number of sequences already in that node.
+The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
+ Figure~\ref{suppfig:tree_RpYN06_RmYN02} has already demonstrated that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
 \label{fig:tree_with_deleted}
@@
 
-Repeating the analysis of the previous section with these new sequences shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
-This fact suggests that the Huanan Seafood Market sequences, which are the primary focus of the genomic epidemiology analyses in the joint WHO-China report~\citep{WHO2021origins}, may not be representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
+Repeating the analysis of the previous section with these changes shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
+This fact suggests that the Huanan Seafood Market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, may not be representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
 
 Furthermore, it is immediately apparent that the discrepancy between outgroup rooting and the evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong.
-The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
+The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} is now highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
 The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.

### 2021-06-17 paper/paper.tex
first two intro paragraphs

@@
 
-\lettrine[lines=2]{\color{color2}T}{}he origins...
+\lettrine[lines=2]{\color{color2}U}{}nderstanding the spread of SARS-CoV-2 in Wuhan is crucial to tracing the origins of the virus, including identifying events that led to infection of patient zero.
+The first reports outside of China at the end of December 2019 emphasized the Huanan Seafood Market~\citep{ProMED2109}, which was initially suggested as a site of zoonosis.
+However, this theory became increasingly tenuous as it was revealed that many early cases had no connection to the market~\citep{huang2020clinical, chen2020epidemiological}.
+By May 2020, Chinese CDC Director Gao Fu dismissed the theory, stating ``At first, we assumed the seafood market might have the virus, but now the market is more like a victim.
+The novel coronavirus had existed long before''~\citep{GlobalTimes2020}.
+
+Instead, there were reports of cases that far preceded the outbreak at the Huanan Seafood Market.
+The \textit{Lancet} described a confirmed case having no association with the market whose symptoms began on December 1, 2019~\citep{huang2020clinical}.
+The \textit{South China Morning Post} described nine cases from November 2019, noting that none of them were confirmed to be ``patient zero''~\citep{ma2020SCMP}.
+Around February 25, 2020, Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} there were two confirmed cases in mid-November of 2019, and one suspected case on September 29~\citep{healthtimes2020}.
+That same day, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic and publication of research results without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
+Then in early March of 2020, China's State Council issued a much broader order requiring approval of all papers related to COVID-19 to ensure publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
+After this, reports of early cases slowed---and in 2021, the joint WHO-China report dismissed all reported cases prior to December 8 of 2020 as not being COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
 
@@
 The pre-print~\citep{wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
-The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be approved by a centralized group to ensure that scientific publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
+The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be centrally approved~\citep{Kang2020}.
 The final published manuscript~\citep{wang2020small} from June of 2020 updated the description from ``early in the epidemic'' to ``early in the epidemic (January 2020).''

### 2021-06-17 paper/paper.tex
full draft of Intro

--- 2021-06-17 paper/paper.tex
@@
 
-\lettrine[lines=2]{\color{color2}T}{}he origins...
+\lettrine[lines=2]{\color{color2}U}{}nderstanding the spread of SARS-CoV-2 in Wuhan is crucial to tracing the origins of the virus, including identifying events that led to infection of patient zero.
+The first reports outside of China at the end of December 2019 emphasized the role of the Huanan Seafood Market~\citep{ProMED2109}, which was initially suggested as a site of zoonosis.
+However, this theory became increasingly tenuous as it was learned that many early cases had no connection to the market~\citep{huang2020clinical, chen2020epidemiological}.
+Soon, Chinese CDC Director Gao Fu dismissed the theory, stating ``At first, we assumed the seafood market might have the virus, but now the market is more like a victim.
+The novel coronavirus had existed long before''~\citep{GlobalTimes2020}.
+
+Indeed, there were reports of cases that far preceded the outbreak at the Huanan Seafood Market.
+The \textit{Lancet} described a confirmed case having no association with the market whose symptoms began on December 1, 2019~\citep{huang2020clinical}.
+The \textit{South China Morning Post} described nine cases from November 2019 including details on patient age and sex, noting that none were confirmed to be ``patient zero''~\citep{ma2020SCMP}.
+Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} that records he had reviewed showed two cases in mid-November, and one suspected case on September 29~\citep{healthtimes2020}.
+At about the same time as Professor Chuanhua's interview, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
+Then China's State Council issued a much broader order requiring central approval of all papers related to COVID-19 to ensure publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
+After this, reports of early cases slowed---and in 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
+
+In other outbreaks where direct identification of early cases has been stymied, it has increasingly become possible to use genomic epidemiology to infer the timing and dynamics of spread from analysis of viral sequences.
+For instance, analysis of SARS-CoV-2 sequences has enabled reconstruction of the initial spread of SARS-CoV-2 in North America and Europe~\citep{bedford2020cryptic, worobey2020emergence, deng2020genomic, fauver2020coast}.
+
+But in the case of Wuhan, genomic epidemiology has also proven frustratingly inconclusive.
+Some of the problem is simply limited data: despite the fact that Wuhan has advanced virology labs, there is only patchy sampling of SARS-CoV-2 sequences from the first months of the city's explosive outbreak.
+Other than a set of multiply sequenced samples collected in late December of 2019 from a dozen patients connected to the Huanan Seafood Market~\citep{WHO2021origins}, just a handful of Wuhan sequences are available from before late January of 2020 (see analysis in this study below).
+The paucity of sequences could be due in part to an order that unauthorized Chinese labs destroy all coronavirus samples from early in the outbreak, reportedly for ``laboratory biological safety'' reasons~\citep{pinghui2020SCMP}.
+
+However, the Wuhan sequences that are available have also confounded phylogenetic analyses designed to infer the ``progenitor'' of all known SARS-CoV-2~\citep{kumar2021evolutionary}.
+Although there is debate about exactly how SARS-CoV-2 entered the human population, it is universally accepted that the virus's deep ancestors are bat coronaviruses~\citep{lytras2021exploring}.
+But the earliest known SARS-CoV-2 sequences, which are mostly derived from the Huanan Seafood Market, are notably more different from these bat coronaviruses than other SARS-CoV-2 sequences collected at later dates outside Wuhan.
+As a result, there is a direct conflict between the two major principles used to infer an outbreak's progenitor: namely that it should be among the earliest sequences, and that it should be most closely related to deeper ancestors~\citep{pipes2021assessing}.
+
+Here I take a step towards resolving these questions by identifying and recovering a deleted data set of partial SARS-CoV-2 sequences from outpatient samples collected early in the Wuhan epidemic.
+Analysis of these new sequences in conjunction with careful annotation of existing ones suggests that the early Wuhan samples that have been the focus of most studies including the joint WHO-China report~\citep{WHO2021origins} may not be fully representative of the viruses actually present early in the Wuhan epidemic.
+These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to the progenitor inferred by \citet{kumar2021evolutionary}.
+Furthermore, the approach taken here hints that it may be possible to advance understanding of the origins or early spread of SARS-CoV-2 even without further on-the-ground studies in China, such as by more deeply probing data archived by the NIH and other entities outside China.
 

### 2021-06-17 paper/paper.tex
revisions to Intro

@@
 At about the same time as Professor Chuanhua's interview, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
-Then China's State Council issued a much broader order requiring central approval of all papers related to COVID-19 to ensure publications were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
+Then China's State Council issued a much broader order requiring central approval of all publications related to COVID-19 to ensure they were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
 After this, reports of early cases slowed---and in 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
@@
 
-However, the Wuhan sequences that are available have also confounded phylogenetic analyses designed to infer the ``progenitor'' of all known SARS-CoV-2~\citep{kumar2021evolutionary}.
+However, the Wuhan sequences that are available have also confounded phylogenetic analyses designed to infer the ``progenitor'' of SARS-CoV-2~\citep{kumar2021evolutionary}.
 Although there is debate about exactly how SARS-CoV-2 entered the human population, it is universally accepted that the virus's deep ancestors are bat coronaviruses~\citep{lytras2021exploring}.
@@
 Here I take a step towards resolving these questions by identifying and recovering a deleted data set of partial SARS-CoV-2 sequences from outpatient samples collected early in the Wuhan epidemic.
-Analysis of these new sequences in conjunction with careful annotation of existing ones suggests that the early Wuhan samples that have been the focus of most studies including the joint WHO-China report~\citep{WHO2021origins} may not be fully representative of the viruses actually present early in the Wuhan epidemic.
-These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to the progenitor inferred by \citet{kumar2021evolutionary}.
-Furthermore, the approach taken here hints that it may be possible to advance understanding of the origins or early spread of SARS-CoV-2 even without further on-the-ground studies in China, such as by more deeply probing data archived by the NIH and other entities outside China.
+Analysis of these new sequences in conjunction with careful annotation of existing ones suggests that the early Wuhan samples that have been the focus of most studies including the joint WHO-China report~\citep{WHO2021origins} are not fully representative of the viruses actually present in Wuhan at that time.
+These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to that inferred by \citet{kumar2021evolutionary}.
+Furthermore, the approach taken here hints it may be possible to advance understanding of SARS-CoV-2's origins or early spread even without further on-the-ground studies, such as by more deeply probing data archived by the NIH and other entities outside China.
 

### 2021-06-17 paper/paper.tex
some progress on Conclusion

@@
 
-\title{Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic}
+\title{Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic}
 
@@
 Other than a set of multiply sequenced samples collected in late December of 2019 from a dozen patients connected to the Huanan Seafood Market~\citep{WHO2021origins}, just a handful of Wuhan sequences are available from before late January of 2020 (see analysis in this study below).
-The paucity of sequences could be due in part to an order that unauthorized Chinese labs destroy all coronavirus samples from early in the outbreak, reportedly for ``laboratory biological safety'' reasons~\citep{pinghui2020SCMP}.
+This paucity of sequences could be due in part to an order that unauthorized Chinese labs destroy all coronavirus samples from early in the outbreak, reportedly for ``laboratory biological safety'' reasons~\citep{pinghui2020SCMP}.
 
@@
 \subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
-The approach of \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
-The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
+\citet{wang2020medRxiv} sequenced PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
+They also sequenced a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
 In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
-I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the termini of the region.
+I slightly trimmed the region of interest to 21,570 to 29,550 because many samples had poor coverage at the termini.
 
@@
 I also determined the consensus sequence of the plasmid control used by \citet{wang2020medRxiv} from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
-Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), this fact indicates plasmid contamination did not afflict the viral samples in the deleted sequencing project.
+Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), plasmid contamination did not afflict the viral samples in the deleted sequencing project.
 
@@
 The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
-The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
-All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
+The discrepancy is especially pronounced for sequences from patients who had visited the Huanan Seafood Market~\citep{WHO2021origins}.
+All sequences associated with this market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
 Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
@@
  \caption{
- Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected no later than January 31, 2020.
- The trees are identical except they are rooted to make the putative progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
+ Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected before Februrary, 2020.
+ The trees are identical except they are rooted to make the progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
 Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected.
@@
 The labels at the top of each tree give the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and Wuhan-Hu-1.
-The monophyletic group containing C28144T is collapsed into a single node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is RpYN06 or RmYN02.
+The monophyletic group containing C28144T is collapsed into a node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
+ Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
 \label{fig:tree_RaTG13}
@@
 
-This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
+This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the progenitor sequence is closest to the bat coronavirus outgroup.
 If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
@@
 \caption{
-Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (sites 21,570--29,550).
+Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (21,570--29,550).
 The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.
-Mutational distance is calculated relative to proCoV2, and points are slightly jittered on the y-axis.
-Go to {\color{red} add link} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of y-axis jittering.
+Mutational distance is calculated relative to proCoV2, and points are jittered on the y-axis.
+Go to {\color{red} add link} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of jittering.
 \label{fig:deltadist_jitter}
@@
 The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} has already demonstrated that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
+ Figure~\ref{suppfig:tree_RpYN06_RmYN02} demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
 \label{fig:tree_with_deleted}
@@
 Repeating the analysis of the previous section with these changes shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
-This fact suggests that the Huanan Seafood Market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, may not be representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
+This fact suggests that the market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, are not representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
 
@@
 
-We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
+We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple branching lineages.
 This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
@@
 \section{Discussion}
+I have identified and recovered a deleted data set of partial SARS-CoV-2 sequences from the early Wuhan epidemic.
+Analysis of these sequences leads to several important conclusions.
+First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are clearly not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
+Instead, the deleted data set as well as analysis of existing sequences from Wuhan-infected patients hospitalized in Guangdong shows early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than the sequences in the joint WHO-China report~\citep{WHO2021origins}.
+Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2 sequences.
+One of these putative progenitors is proCoV2 described by \citet{kumar2021evolutionary}, and the other is a sequence that carries three mutations (C8782T, T28144C, and C29095T) relative to Wuhan-Hu-1.
+Crucially, both putative progenitors are three mutations closer to SARS-CoV-2's bat coronavirus relatives than sequences from the Huanan Seafood Market.
+Note also that the progenitor of all currently known SARS-CoV-2 sequences could still be downstream of the sequence that actually infected patient zero depending on the transmission dynamics of the earliest infections.
 
+In addition, the fact that such an informative data set was deleted carries implications beyond those gleaned directly from the recovered sequences.
+Samples from early outpatients in the Wuhan epidemic are a gold mine for anyone seeking to understand the early spread of the virus.
+Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than surreptitiously delete the partial sequences.
+There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states proper human subjects approval was obtained, and the sequencing data show no evidence of plasmid or sample-to-sample contamination.
+It therefore seems likely that the data were deleted from the SRA to obscure their existence from scientists searching such databases.
+
+There are several caveats...
+Sequences only partial and lacking full metadata...
 Something about lacking formal analyses of some other studies, but maybe this is not so bad.
+Still not parsimonious
+
+Finally, implications for further study. getting info from NIH on these and others.
 

### 2021-06-17 paper/paper.tex
all bat last paragraph of Discussion

@@
 To examine if the sequences recovered from the deleted data set help resolve the conundrum described in the previous section, I repeated the analyses including those sequences.
-In the process, I noted another salient fact: four GISAID sequences collected in Guangdong that fall in a putative progenitor node are from patients who traveled to Wuhan in late December of 2019 and developed symptoms before or on the day that they returned to Guangdong, where their viruses were ultimately sequenced~\citep{chan2020familial, kang2020evidence}.
+In the process, I noted another salient fact: four GISAID sequences collected in Guangdong that fall in a putative progenitor node are from two different clusters of patients who traveled to Wuhan in late December of 2019 and developed symptoms before or on the day that they returned to Guangdong, where their viruses were ultimately sequenced~\citep{chan2020familial, kang2020evidence}.
 Since these patients were clearly infected in Wuhan even though they were sequenced in Guangdong, I annotated them separately from both the other Wuhan and other China sequences.
@@
 \section{Discussion}
-I have identified and recovered a deleted data set of partial SARS-CoV-2 sequences from the early Wuhan epidemic.
+I have identified and recovered a deleted set of partial SARS-CoV-2 sequences from the early Wuhan epidemic.
 Analysis of these sequences leads to several important conclusions.
-First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are clearly not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
-Instead, the deleted data set as well as analysis of existing sequences from Wuhan-infected patients hospitalized in Guangdong shows early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than the sequences in the joint WHO-China report~\citep{WHO2021origins}.
-Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2 sequences.
+First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
+The deleted data set as well as analysis of existing sequences from Wuhan-infected patients hospitalized in Guangdong shows early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than sequences in the joint WHO-China report~\citep{WHO2021origins}.
+Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2.
 One of these putative progenitors is proCoV2 described by \citet{kumar2021evolutionary}, and the other is a sequence that carries three mutations (C8782T, T28144C, and C29095T) relative to Wuhan-Hu-1.
 Crucially, both putative progenitors are three mutations closer to SARS-CoV-2's bat coronavirus relatives than sequences from the Huanan Seafood Market.
-Note also that the progenitor of all currently known SARS-CoV-2 sequences could still be downstream of the sequence that actually infected patient zero depending on the transmission dynamics of the earliest infections.
+Note also that the progenitor of all known SARS-CoV-2 sequences could still be downstream of the sequence that infected patient zero depending on the transmission dynamics of the first infections.
 
-In addition, the fact that such an informative data set was deleted carries implications beyond those gleaned directly from the recovered sequences.
-Samples from early outpatients in the Wuhan epidemic are a gold mine for anyone seeking to understand the early spread of the virus.
+The fact that such an informative data set was deleted also has implications beyond those gleaned directly from the recovered sequences.
+Samples from early outpatients in Wuhan are a gold mine for anyone seeking to understand spread of the virus.
 Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than surreptitiously delete the partial sequences.
-There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states proper human subjects approval was obtained, and the sequencing data show no evidence of plasmid or sample-to-sample contamination.
-It therefore seems likely that the data were deleted from the SRA to obscure their existence from scientists searching such databases.
+There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing data show no evidence of plasmid or sample-to-sample contamination.
+It therefore seems likely the sequences were deleted from the SRA to obscure their existence.
+Particularly in light of the directive that unauthorized labs destroy early viral samples~\citep{pinghui2020SCMP} and the multiple orders requiring approval of publications related to COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to trace the early spread of the epidemic.
 
-There are several caveats...
-Sequences only partial and lacking full metadata...
+Another important implication is that genomic epidemiology studies of early SARS-CoV-2 need to pay as much attention to the provenance and annotation of the underlying sequences as technical considerations.
+There has been substantial scientific effort expended on topics such as phylogenetic rooting~\citep{pipes2021assessing, morel2021phylogenetic}, novel algorithms~\citep{kumar2021evolutionary}, and correction of sequencing errors~\citep{turakhia2020stability}.
+Future studies should devote equal effort to going beyond the default annotations in GISAID to carefully trace the location of patient infection and sample sequencing.
+The potential importance of such work is revealed by the observation above that many of the sequences closest to SARS-CoV-2's bat coronavirus relatives are from early patients who were infected in Wuhan, but then sequenced in and attributed to Guangdong.
+
+There are several caveats to this study.
+Most obviously, the sequences recovered from the deleted data set are partial and lack full metadata.
+Therefore, it is impossible to unambiguously place them phylogenetically, or determine exactly when they were collected.
+However, little can be done to mitigate this caveat beyond my failed attempt to contact the corresponding authors of \citet{wang2020medRxiv}.
+It is also important to note that my phylogenetic analyses use relatively simple methods to draw qualitative conclusions without formal statistical testing.
+Certainly further application of more advanced methods would be a welcome advance.
+However, qualitative and visual analyses do have advantages when the key questions relate more to the underlying data than the sophistication of the inferences.
+Finally, both plausible putative progenitors described above require that one of the earliest mutations to SARS-CoV-2 was a reversion towards the bat coronavirus outgroups (either C18060T or C29095T) on a branch that subsequently gave rise to multiple distinct descendants.
+Such a non-parsimonious scenario can only be avoided by invoking recombination very early in the pandemic, which is not entirely implausible for a coronavirus~\citep{boni2020evolutionary}.
+However, because the outgroups have $\sim$4\% nucleotide divergence from SARS-CoV-2, a mutation towards the outgroup is also in the realm of reasonable possibility.
+Of course, future identification of additional early sequences could fully resolve these questions.
+
 Something about lacking formal analyses of some other studies, but maybe this is not so bad.
@@
 
-Finally, implications for further study. getting info from NIH on these and others.
+More broadly, ..., implications for further study. getting info from NIH on these and others.
 

### 2021-06-17 paper/paper.tex
full draft of discussion

@@
 \caption{Example of the process to delete SRA data.
-The image shows e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff excerpted from \url{https://usrtk.org/tag/pangolin-papers/}.
+The image shows e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff excerpted from \citet{USRTK2020}.
 }
@@
 Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2.
-One of these putative progenitors is proCoV2 described by \citet{kumar2021evolutionary}, and the other is a sequence that carries three mutations (C8782T, T28144C, and C29095T) relative to Wuhan-Hu-1.
+One is proCoV2 described by \citet{kumar2021evolutionary}, and the other is a sequence that carries three mutations (C8782T, T28144C, and C29095T) relative to Wuhan-Hu-1.
 Crucially, both putative progenitors are three mutations closer to SARS-CoV-2's bat coronavirus relatives than sequences from the Huanan Seafood Market.
@@
 There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing data show no evidence of plasmid or sample-to-sample contamination.
-It therefore seems likely the sequences were deleted from the SRA to obscure their existence.
+It therefore seems likely the sequences were deleted to obscure their existence.
 Particularly in light of the directive that unauthorized labs destroy early viral samples~\citep{pinghui2020SCMP} and the multiple orders requiring approval of publications related to COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to trace the early spread of the epidemic.
@@
 It is also important to note that my phylogenetic analyses use relatively simple methods to draw qualitative conclusions without formal statistical testing.
-Certainly further application of more advanced methods would be a welcome advance.
+Further application of more advanced methods would be a welcome advance.
 However, qualitative and visual analyses do have advantages when the key questions relate more to the underlying data than the sophistication of the inferences.
-Finally, both plausible putative progenitors described above require that one of the earliest mutations to SARS-CoV-2 was a reversion towards the bat coronavirus outgroups (either C18060T or C29095T) on a branch that subsequently gave rise to multiple distinct descendants.
-Such a non-parsimonious scenario can only be avoided by invoking recombination very early in the pandemic, which is not entirely implausible for a coronavirus~\citep{boni2020evolutionary}.
-However, because the outgroups have $\sim$4\% nucleotide divergence from SARS-CoV-2, a mutation towards the outgroup is also in the realm of reasonable possibility.
+Finally, both plausible putative progenitors require that one of the earliest mutations to SARS-CoV-2 was a reversion towards the bat coronavirus outgroups (either C18060T or C29095T) on a branch that subsequently gave rise to multiple distinct descendants.
+Such a scenario can only be avoided by invoking recombination very early in the pandemic, which is not entirely implausible for a coronavirus~\citep{boni2020evolutionary}.
+However, because the outgroups have $\sim$4\% nucleotide divergence from SARS-CoV-2, a mutation towards the outgroup is also entirely possible.
 Of course, future identification of additional early sequences could fully resolve these questions.
 
-Something about lacking formal analyses of some other studies, but maybe this is not so bad.
-Still not parsimonious
-
-More broadly, ..., implications for further study. getting info from NIH on these and others.
+More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without a complex international investigation.
+Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.
+In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions, which are already known to include SRR11119760 and SRR11119761~\citep{USRTK2020}.
+Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for their staff to validate the rationale for all requests.
+However, the current study suggests that at least in one case, the trusting structures of scientific data sharing have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
+Therefore, a careful re-evaluation of other routine forms of scientific communication, reporting, and data sharing could shed additional light on early events in the emergence of the pandemic.
 

### 2021-06-17 paper/paper.tex
full draft of Discussion

@@
 Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for their staff to validate the rationale for all requests.
-However, the current study suggests that at least in one case, the trusting structures of scientific data sharing have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
-Therefore, a careful re-evaluation of other routine forms of scientific communication, reporting, and data sharing could shed additional light on early events in the emergence of the pandemic.
+However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
+Therefore, a careful re-evaluation of archived forms of scientific communication, reporting, and data sharing could shed additional light on early events in the emergence of the pandemic.
 

### 2021-06-17 paper/paper.tex
revisions to Intro and Discussion

@@
 The \textit{South China Morning Post} described nine cases from November 2019 including details on patient age and sex, noting that none were confirmed to be ``patient zero''~\citep{ma2020SCMP}.
-Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} that records he had reviewed showed two cases in mid-November, and one suspected case on September 29~\citep{healthtimes2020}.
+Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} that records he reviewed showed two cases in mid-November, and one suspected case on September 29~\citep{healthtimes2020}.
 At about the same time as Professor Chuanhua's interview, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
@@
 I have identified and recovered a deleted set of partial SARS-CoV-2 sequences from the early Wuhan epidemic.
-Analysis of these sequences leads to several important conclusions.
+Analysis of these sequences leads to several conclusions.
 First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
-The deleted data set as well as analysis of existing sequences from Wuhan-infected patients hospitalized in Guangdong shows early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than sequences in the joint WHO-China report~\citep{WHO2021origins}.
+The deleted data as well as existing sequences from Wuhan-infected patients hospitalized in Guangdong show early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than sequences in the joint WHO-China report~\citep{WHO2021origins}.
 Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2.
@@
 
-The fact that such an informative data set was deleted also has implications beyond those gleaned directly from the recovered sequences.
+The fact that such an informative data set was deleted has implications beyond those gleaned directly from the recovered sequences.
 Samples from early outpatients in Wuhan are a gold mine for anyone seeking to understand spread of the virus.
 Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than surreptitiously delete the partial sequences.
-There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing data show no evidence of plasmid or sample-to-sample contamination.
+There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
 It therefore seems likely the sequences were deleted to obscure their existence.
-Particularly in light of the directive that unauthorized labs destroy early viral samples~\citep{pinghui2020SCMP} and the multiple orders requiring approval of publications related to COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to trace the early spread of the epidemic.
+Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to trace early spread of the epidemic.
 
@@
 There has been substantial scientific effort expended on topics such as phylogenetic rooting~\citep{pipes2021assessing, morel2021phylogenetic}, novel algorithms~\citep{kumar2021evolutionary}, and correction of sequencing errors~\citep{turakhia2020stability}.
-Future studies should devote equal effort to going beyond the default annotations in GISAID to carefully trace the location of patient infection and sample sequencing.
-The potential importance of such work is revealed by the observation above that many of the sequences closest to SARS-CoV-2's bat coronavirus relatives are from early patients who were infected in Wuhan, but then sequenced in and attributed to Guangdong.
+Future studies should devote equal effort to going beyond the annotations in GISAID to carefully trace the location of patient infection and sample sequencing.
+The potential importance of such work is revealed by the observation that many of the sequences closest to SARS-CoV-2's bat coronavirus relatives are from early patients who were infected in Wuhan, but then sequenced in and attributed to Guangdong.
 
@@
 However, qualitative and visual analyses do have advantages when the key questions relate more to the underlying data than the sophistication of the inferences.
-Finally, both plausible putative progenitors require that one of the earliest mutations to SARS-CoV-2 was a reversion towards the bat coronavirus outgroups (either C18060T or C29095T) on a branch that subsequently gave rise to multiple distinct descendants.
+Finally, both plausible putative progenitors require that an early mutation to SARS-CoV-2 was a reversion towards the bat coronavirus outgroups (either C18060T or C29095T) on a branch that subsequently gave rise to multiple distinct descendants.
 Such a scenario can only be avoided by invoking recombination very early in the pandemic, which is not entirely implausible for a coronavirus~\citep{boni2020evolutionary}.
@@
 
-More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without a complex international investigation.
+More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without an international investigation.
 Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.

### 2021-06-17 paper/paper.tex
draft abstract

@@
 \begin{abstract}
-Abstract here
+The origin and early spread of SARS-CoV-2 remains shrouded in mystery.
+Here I identify a data set containing SARS-CoV-2 sequences from early in the Wuhan epidemic that has been deleted from the NIH's Sequence Read Archive.
+I recover the deleted files from the Google Cloud, and use them to reconstruct partial sequences of 13 early epidemic viruses.
+Phylogenetic analysis of these sequences in the context of carefully annotated existing data suggest that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
+Instead, the progenitor of currently known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
 \end{abstract}

### 2021-06-17 paper/paper.tex
draft of abstract

@@
 I recover the deleted files from the Google Cloud, and use them to reconstruct partial sequences of 13 early epidemic viruses.
-Phylogenetic analysis of these sequences in the context of carefully annotated existing data suggest that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
+Phylogenetic analysis of these sequences in the context of carefully annotated existing data suggests that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
 Instead, the progenitor of currently known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
@@
 
-\subsection{Identification of a SARS-CoV-2 deep sequencing data set that has been removed from the NCBI's Sequence Read Archive}
-During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed SARS-CoV-2 deep sequencing data from the Sequence Read Archive (SRA), which is a repository maintained by the NIH's National Center for Biotechnology Information (NCBI).
+\subsection{Identification of a SARS-CoV-2 deep sequencing data set that has been removed from the Sequence Read Archive}
+During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed SARS-CoV-2 deep sequencing data from the Sequence Read Archive (SRA), which is a repository maintained by the NIH's National Center for Biotechnology Information.
 The first supplementary table of \citet{farkas2020insights} lists all SARS-CoV-2 deep sequencing data available from the SRA as of March 30, 2020.

### 2021-06-17 paper/paper.tex
some formatting

@@
 \usepackage{xurl}
+\usepackage{stfloats}
+
+\renewcommand{\topfraction}{0.9}	% max fraction of floats at top
+    \renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
+    %   Parameters for TEXT pages (not float pages):
+    \setcounter{topnumber}{2}
+    \setcounter{bottomnumber}{2}
+    \setcounter{totalnumber}{4}     % 2 may work better
+    \setcounter{dbltopnumber}{2}    % for 2-column pages
+    \renewcommand{\dbltopfraction}{0.9}	% fit big float above 2-col. text
+    \renewcommand{\textfraction}{0.07}	% allow minimal text w. figs
+    %   Parameters for FLOAT pages (not text pages):
+    \renewcommand{\floatpagefraction}{0.7}	% require fuller float pages
+	% N.B.: floatpagefraction MUST be less than topfraction !!
+    \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages
 
@@
 Here I identify a data set containing SARS-CoV-2 sequences from early in the Wuhan epidemic that has been deleted from the NIH's Sequence Read Archive.
-I recover the deleted files from the Google Cloud, and use them to reconstruct partial sequences of 13 early epidemic viruses.
+I recover the deleted files from the Google Cloud, and reconstruct partial sequences of 13 early epidemic viruses.
 Phylogenetic analysis of these sequences in the context of carefully annotated existing data suggests that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
-Instead, the progenitor of currently known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
+Instead, the progenitor of known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
 \end{abstract}
@@
 \centering
-\fbox{\includegraphics[trim={0.8in 0.8in 1.1in 1in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
+\fbox{\includegraphics[trim={0in 0in 0in 0in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
 \caption{Example of the process to delete SRA data.
@@
 
-Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
-This strategy was successful; for instance, as of June 3, 2021, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
-I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
-
-I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February (files for SRR11313490 and SRR11313499 were not accessible via the cloud).
-I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for each SRA file.
-For all files, the object timestamp is February 15, 2020, and the time is March 16, 2020.
-Although the SRA Toolkit does not clearly document these two properties, my \emph{guess} is that the object timestamp may refer to when the SRA file was created from a FASTQ file uploaded to the SRA, and the time may refer to when the accession was made public.
-
-\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
-\citet{wang2020medRxiv} sequenced PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
-They also sequenced a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
-In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
-I slightly trimmed the region of interest to 21,570 to 29,550 because many samples had poor coverage at the termini.
-
-I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
-Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
-
-\begin{table*}[]
+\begin{table*}[b!]
 {\small
@@
 
+Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
+This strategy was successful; for instance, as of June 3, 2021, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
+I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
+
+I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February (files for SRR11313490 and SRR11313499 were not accessible via the cloud).
+I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for each SRA file.
+For all files, the object timestamp is February 15, 2020, and the time is March 16, 2020.
+Although the SRA Toolkit does not clearly document these two properties, my \emph{guess} is that the object timestamp may refer to when the SRA file was created from a FASTQ file uploaded to the SRA, and the time may refer to when the accession was made public.
+
+\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
+\citet{wang2020medRxiv} sequenced PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
+They also sequenced a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
+In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
+I slightly trimmed the region of interest to 21,570 to 29,550 because many samples had poor coverage at the termini.
+
+I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
+Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
+
 I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
@@
 Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), plasmid contamination did not afflict the viral samples in the deleted sequencing project.
-
-\subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
-To contextualize the viral sequences recovered from the deleted project, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
-The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.
-
-Known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
-However, attempts to infer this progenitor have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
-This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
-One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to these bat coronavirus relatives---but this is not the case.
 
@@
 \end{figure*}
+
+\subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
+To contextualize the viral sequences recovered from the deleted project, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
+The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.
+
+Known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
+However, attempts to infer this progenitor have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
+This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
+One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to these bat coronavirus relatives---but this is not the case.
 

### 2021-06-17 paper/paper.tex
start adding backmatter

@@
 
-\subsection{Recovery of deleted sequencing data from the Google cloud} 
+\subsection{Recovery of deleted sequencing data from the Google Cloud} 
 As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
@@
 
+{\small
+
 \section{Methods}
-
 \subsection{Code and data availability}
-The deleted SRA files recovered from the Google cloud are all available at  \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
+The deleted SRA files recovered from the Google Cloud are all available at  \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
 I have suffixed the file extension \texttt{.sra} to all of these files.
@@
 \subsection{Archiving of key weblinks}
-I have digitally archived key weblinks in the Wayback Machine.
-These include:
+I have digitally archived key weblinks in the Wayback Machine, including a subset of the SRA files from PRJNA612766 on the Google Cloud:
 \begin{itemize}
-\item The first supplementary table of \citet{farkas2020insights}, which lists all SARS-CoV-2 accessions in the SRA as of March 30, 2020, is archived at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
+\item The first supplementary table of \citet{farkas2020insights} is archived at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
+\item SRR11313485: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}
+\item SRR11313486: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313486/SRR11313486}
+\item SRR11313274: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313274/SRR11313274}
+\item SRR11313275: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313275/SRR11313275}
+\item SRR11313285: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313285/SRR11313285}
+\item SRR11313286: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313286/SRR11313286}
+\item SRR11313448: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313448/SRR11313448}
+\item SRR11313449: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313449/SRR11313449}
+\item SRR11313427: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313427/SRR11313427}
+\item SRR11313429: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313429/SRR11313429}
 \end{itemize}
+
+\section{Acknowledgments}
+I thank the many citizens and scientists on Twitter who helped inspire this study and inform its background through their careful analyses and discussions of the early SARS-CoV-2 epidemic.
+The scientific computing infrastructure used in this work was supported by the NIH Office of Research Infrastructure Programs under S10OD028685.
+The author is an Investigator of the Howard Hughes Medical Institute.
+
+\section{Competing interests}
+The author consults for Moderna on SARS-CoV-2 evolution and epidemiology, consults for Flagship Labs 77 on viral deep mutational scanning, and has the potential to receive a share of IP revenue as an inventor on a Fred Hutch licensed technology/patent (application WO2020006494) related to deep mutational scanning of viral proteins.
+
+}
 

### 2021-06-18 data/comparator_genomes_gisaid/parse_gisaid_to_acc_ack.py
add GISAID acknowledgments and accessions

@@ -1,0 +1,31 @@
+"""Parse xzipped GISAID TSV to accessions and acknowledgments."""
+
+
+import lzma
+
+import pandas as pd
+
+
+with lzma.open('1622468911409.metadata.tsv.xz') as f:
+    df = (pd.read_csv(f, sep='\t')
+          # next two lines handle entries with newlines
+          .query('gisaid_epi_isl.notnull()')
+          .query('gisaid_epi_isl.str.startswith("EPI_ISL_")')
+          .reset_index(drop=True)
+          )
+
+accessions = df['gisaid_epi_isl'].tolist()
+assert len(accessions) == len(set(accessions))
+acc_file = 'accessions.txt'
+print(f"Writing the {len(accessions)} to {acc_file}")
+with open(acc_file, 'w') as f:
+    f.write('\n'.join(accessions))
+
+acknowledgments = (
+    df
+    [['originating_lab', 'submitting_lab', 'authors']]
+    .drop_duplicates()
+    )
+ack_file = 'acknowledgments.txt'
+print(f"Writing the {len(acknowledgments)} to {ack_file}")
+acknowledgments.to_csv(ack_file, index=False)

### 2021-06-18 data/gisaid_sequences_through_Feb2020/parse_gisaid_to_acc_ack.py

@@ -1,0 +1,31 @@
+"""Parse xzipped GISAID TSV to accessions and acknowledgments."""
+
+
+import lzma
+
+import pandas as pd
+
+
+with lzma.open('1622384383620.metadata.tsv.xz') as f:
+    df = (pd.read_csv(f, sep='\t')
+          # next two lines handle entries with newlines
+          .query('gisaid_epi_isl.notnull()')
+          .query('gisaid_epi_isl.str.startswith("EPI_ISL_")')
+          .reset_index(drop=True)
+          )
+
+accessions = df['gisaid_epi_isl'].tolist()
+assert len(accessions) == len(set(accessions))
+acc_file = 'accessions.txt'
+print(f"Writing the {len(accessions)} to {acc_file}")
+with open(acc_file, 'w') as f:
+    f.write('\n'.join(accessions))
+
+acknowledgments = (
+    df
+    [['originating_lab', 'submitting_lab', 'authors']]
+    .drop_duplicates()
+    )
+ack_file = 'acknowledgments.txt'
+print(f"Writing the {len(acknowledgments)} to {ack_file}")
+acknowledgments.to_csv(ack_file, index=False)

### 2021-06-18 data/comparator_genomes_gisaid/parse_gisaid_to_acc_ack.py
acknowledgments in CSV rather than text

@@ -26,6 +26,6 @@ acknowledgments = (
     [['originating_lab', 'submitting_lab', 'authors']]
     .drop_duplicates()
     )
-ack_file = 'acknowledgments.txt'
+ack_file = 'acknowledgments.csv'
 print(f"Writing the {len(acknowledgments)} to {ack_file}")
 acknowledgments.to_csv(ack_file, index=False)

### 2021-06-18 paper/paper.tex
add GISAID acknowledgments and accessions
Methods on SRA downloads

@@
 The first reports outside of China at the end of December 2019 emphasized the role of the Huanan Seafood Market~\citep{ProMED2109}, which was initially suggested as a site of zoonosis.
-However, this theory became increasingly tenuous as it was learned that many early cases had no connection to the market~\citep{huang2020clinical, chen2020epidemiological}.
+However, this theory became increasingly tenuous as it was learned that many early cases had no connection to the market~\citep{cohen2020wuhan,huang2020clinical, chen2020epidemiological}.
 Soon, Chinese CDC Director Gao Fu dismissed the theory, stating ``At first, we assumed the seafood market might have the virus, but now the market is more like a victim.
@@
 These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to that inferred by \citet{kumar2021evolutionary}.
-Furthermore, the approach taken here hints it may be possible to advance understanding of SARS-CoV-2's origins or early spread even without further on-the-ground studies, such as by more deeply probing data archived by the NIH and other entities outside China.
+Furthermore, the approach taken here hints it may be possible to advance understanding of SARS-CoV-2's origins or early spread even without further on-the-ground studies, such as by more deeply probing data archived by the NIH and other entities.
 
@@
 In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions, which are already known to include SRR11119760 and SRR11119761~\citep{USRTK2020}.
-Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for their staff to validate the rationale for all requests.
+Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for its staff to validate the rationale for all requests.
 However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
-Therefore, a careful re-evaluation of archived forms of scientific communication, reporting, and data sharing could shed additional light on early events in the emergence of the pandemic.
+A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the pandemic.
 
@@
 \subsection{Code and data availability}
+The computer code and input data necessary to reproduce all analyses described in this paper are available on GitHub at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766}.
+This GitHub repository includes a \texttt{Snakemake}~\citep{molder2021sustainable} pipeline that fully automates all steps in the analysis except for downloading of sequences from GISAID, which must be done manually as described in the GitHub repository's README in order to comply with GISAID data sharing terms.
+
 The deleted SRA files recovered from the Google Cloud are all available at  \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
-I have suffixed the file extension \texttt{.sra} to all of these files.
+I have suffixed the file extension \texttt{.sra} to all these files.
+The consensus sequences recovered from these deleted SRA files are linked to in the relevant Methods subsection below.
 
@@
 
+\subsection{Recovery of SRA files from deleted project PRJNA612766}
+I parsed the first supplementary table of \citet{farkas2020insights} to extract the accessions for sequencing runs for deleted SRA BioProject PRJNA612766.
+By cross-referencing the samples described in this table to \citet{wang2020medRxiv, wang2020small}, I identified the accessions corresponding to the 34 early outpatient samples who were positive, as well as the accessions corresponding to the 16 hospitalized patient samples from February.
+Samples had both 10 minute and 4 hour sequencing runtime accessions, which were combined in the subsequent analysis.
+I also identified the samples corresponding to the high-copy plasmid controls to enable analysis of the plasmid sequence to rule out contamination.
+The code used to parse the Excel table is available as a Jupyter notebook at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/manual_analyses/PRJNA612766}.
+
+The main \texttt{Snakemake} pipeline then downloads relevant SRA files from the Google Cloud links by using \texttt{wget} to download files from the Google Cloud from paths like \path{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
+Note that I cannot guarantee that these Google Cloud links will remain active, as my analyses of other deleted SRA runs beyond the scope of this study indicates that only sometimes are deleted SRA files still available via the Google Cloud.
+For this reason, key runs have been archived on the Wayback Machine as described above, and all downloaded SRA files relevant to this study are included in the GitHub repository.
+Note also that as described in this paper's main text, two SRA files could not be downloaded from the Google Cloud using the aforementioned method, and so are not part of this study.
+
+The downloaded SRA files were converted to FASTQ files using \texttt{fasterq-dump} from the SRA Toolkit.
+
 \section{Acknowledgments}
 I thank the many citizens and scientists on Twitter who helped inspire this study and inform its background through their careful analyses and discussions of the early SARS-CoV-2 epidemic.
+I thank the many scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.
 The scientific computing infrastructure used in this work was supported by the NIH Office of Research Infrastructure Programs under S10OD028685.

### 2021-06-18 paper/paper.tex
calling of deleted seq consensus

@@
 However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
-A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the pandemic.
+A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the virus.
 
@@
 
+\subsection{Alignment of recovered reads and calling of consensus sequences}
 The downloaded SRA files were converted to FASTQ files using \texttt{fasterq-dump} from the SRA Toolkit.
+The FASTQ files were pre-processed with \texttt{fastp}~\citep{chen2018fastp} to trim reads and remove low-quality ones (the exact settings using in this pre-processing are specified in the \texttt{Snakemake} file in the GitHub repository).
+
+The reads in these FASTQ files were then aligned to a SARS-CoV-2 reference genome using \texttt{minimap2}~\citep{li2018minimap2} with default settings.
+The reference genome used for the entirety of this study is proCoV2~\citep{kumar2021evolutionary}, which was generated by making the following three single-nucleotide changes to the Wuhan-Hu-1 reference (ASM985889v2) available on NCBI: C8782T, C18060T, and T28144C.
+
+I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to generate files covering the coverage at each site by aligned nucleotides with a quality score of at least 20.
+These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the coverage plot that enable zooming and mouseovers to get statistics for specific sites.
+I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
+These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} was designed to only cover part of the genome.
+
+I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$95\% of the sites in the region of interest from site 21,570 to 29,550.
+These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
 

### 2021-06-18 paper/paper.tex
methods on bat CoVs

@@
 
+\subsection{Bat coronavirus outgroup sequences}
+For analyses that involved comparisons to SARS-CoV-2's bat coronavirus relatives~\citep{lytras2021exploring}, the bat coronavirus sequences were manually downloaded from GISAID~\citep{shu2017gisaid}.
+The sequences used were RaTG13~\citep{zhou2020pneumonia}, RmYN02~\citep{zhou2020novel}, and RpYN06~\citep{zhou2021identification}---although the multiple sequence alignment of these viruses to SARS-CoV-2 also contains PrC31~\citep{li2021novel}, which was not used in the final analyses as it more diverged from SARS-CoV-2 than the other three bat coronaviruses at a whole-genome level.
+The GISAID accessions for these sequences are listed at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/comparator_genomes_gisaid/accessions.txt}, and  a table acknowledging the labs and authors is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/comparator_genomes_gisaid/acknowledgments.csv}.
+Sites in SARS-CoV-2 were mapped to their corresponding nucleotide identities in the bat coronavirus outgroups via a multiple sequence alignment of proCoV2 to the bat coronaviruses generated using \texttt{mafft}~\citep{katoh2013mafft}.
+
+\subsection{Curation and analysis of early SARS-CoV-2 sequences from GISAID}
+I manually
+
 \section{Acknowledgments}

### 2021-06-18 paper/paper.tex

@@
 Then China's State Council issued a much broader order requiring central approval of all publications related to COVID-19 to ensure they were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
-After this, reports of early cases slowed---and in 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
+In 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
 
@@
 
-The main \texttt{Snakemake} pipeline then downloads relevant SRA files from the Google Cloud links by using \texttt{wget} to download files from the Google Cloud from paths like \path{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
-Note that I cannot guarantee that these Google Cloud links will remain active, as my analyses of other deleted SRA runs beyond the scope of this study indicates that only sometimes are deleted SRA files still available via the Google Cloud.
+I recovered the SRA files from the Google Cloud by using \texttt{wget} to download files with from paths like \path{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
+Note that I cannot guarantee that these Google Cloud links will remain active, as my analyses of other deleted SRA runs (beyond the scope of this study) indicates that only sometimes are deleted SRA files still available via the Google Cloud.
 For this reason, key runs have been archived on the Wayback Machine as described above, and all downloaded SRA files relevant to this study are included in the GitHub repository.
@@
 
-I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to generate files covering the coverage at each site by aligned nucleotides with a quality score of at least 20.
-These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the coverage plot that enable zooming and mouseovers to get statistics for specific sites.
+I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to determine the coverage at each site by aligned nucleotides with a quality score of at least 20.
+These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
 I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
-These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} was designed to only cover part of the genome.
+These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} only covers part of the genome.
 
@@
 These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
+These sequences in the region of interest (21,570 to 29,550) are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/recovered_seqs.fa}.
 
@@
 \subsection{Curation and analysis of early SARS-CoV-2 sequences from GISAID}
-I manually
+For the broader analyses of existing SARS-CoV-2 sequences, I downloaded all sequences from collected prior to March of 2020 from GISAID.
+The accessions of these sequences are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/accessions.txt}, and a table acknowledging the labs and authors is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/acknowledgments.csv}.
+
+I then used \texttt{mafft}~\citep{katoh2013mafft} to align these sequences to the proCoV2 reference described above, stripped any sites that were gapped relative to the reference, and filtered the sequences using the following criteria:
+\begin{itemize}
+\item I removed any sequences collected after February 28, 2020.
+\item I removed any sequences that had $\ge$4 mutations within any 10-nucleotide stretch, as such runs of mutations often indicate sequencing errors.
+\item I removed any sequence for which the alignment covered $<$90\% of the proCoV2 sequence.
+\item I removed any sequence with $\ge$15 mutations relative to the reference.
+\item I removed any sequence with $\ge$5,000 ambiguous nucleotides.
+\end{itemize}
+
+I then annotated the sequences using some additional information.
+First, I annotated sequences based on the joint WHO-China report~\citep{WHO2021origins} and also \citet{zhu2020brief} to keep only one representative from multiply sequenced patients, and to indicate which sequences were from patients associated with the Huanan Seafood Market.
+My version of these annotations is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/WHO_China_Report_Dec2019_cases.yaml}.
+Next, I identified some sequences in the set that were clearly duplicates from the same patient, and removed these.
+The annotations used to remove these duplicates are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/seqs_to_exclude.yaml}.
+Finally, I used information from ~\citet{chan2020familial} and \citet{kang2020evidence} to identify patients who were infected in Wuhan before January 5 of 2020, but ultimately sequenced in Guangdong: these annotations are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/Wuhan_exports.yaml}.
+
+I next removed any of the handful of mutations noted by \citet{turakhia2020stability} to be lab artifacts that commonly afflict SARS-CoV-2 sequences.
+I also limited the analyses to the region of the genome that spans from the start of the first coding region (ORF1ab) to the end of the last (ORF10), because I noticed that some sequences had suspicious patterns (such as many mutations or runs of mutations) near the termini of the genome.
+
+The plot in Figure~\ref{fig:deltadist_RaTG13} contains all of the GISAID sequences after this filtering, and was rendered using \texttt{Altair}~\citep{vanderplas2018altair}.
+The plot in Figure~\ref{fig:deltadist_jitter} shows the filtered GISAID sequences collected before February of 2020 plus the 13 good coverage recovered partial early outpatient sequences (Table~\ref{tab:mutations}), considering only the region covered by the partial sequences (21,570 to 29,550).
+
+\subsection{Phylogenetic analyses}
 

### 2021-06-18 paper/paper.tex
full draft of Methods

@@
 \subsection{Phylogenetic analyses}
+The phylogenetic trees were inferred using the GISAD sequences after the filtering and annotations described above, only considering sequences with $\ge$95\% coverage over the region of interest that were collected before February of 2020.
+In addition, after generating this sequence set I removed any sequence variants with a combination of mutations that was not observed at least twice so the analysis only includes multiply observed sequence variants.
+A file indicating the unique sequences used for the phylogenetic analysis, their mutations relative to proCoV2, and other sequences in that cluster is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/phylogenetics/all_alignment.csv}.
+
+I then used \texttt{IQ-Tree}~\citep{minh2020iq} to infer a maximum-likelihood phylogenetic tree using a GTR nucleotide substitution model with empirical nucleotide frequencies, and collapsing zero-length branches to potentially allow a multifurcating tree.
+The inference yielded the tree topology and branch lengths shown in all figures in this study with phylogenetic trees.
+I then rendered the images of the tree using \texttt{ETE 3}~\citep{huerta2016ete}, manually re-rooting the tree to place the first (progenitor) node at each of the three nodes that have the highest identity to the bat coronavirus outgroup.
+In these images, node sizes are proportional to the number of sequences in that node, and are colored in proportion to the location from which those sequences are derived.
+As indicated in the legend to Figure~\ref{fig:tree_RaTG13}, the node containing the monophyletic set of sequences with C28144T is collapsed into a single node in the tree images.
+
+For the trees in which I added the recovered sequences from the deleted data set (Figure~\ref{fig:tree_with_deleted}), the actual trees are exactly the same as those inferred using the GISAID sequences above.
+The difference is that the sequences from the deleted data set are then added to each node with which they are compatible given their mutations in an amount proportional to the size of the node, the logic being that a sequence is more likely to fall into larger clusters.
 
 \section{Acknowledgments}
-I thank the many citizens and scientists on Twitter who helped inspire this study and inform its background through their careful analyses and discussions of the early SARS-CoV-2 epidemic.
-I thank the many scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.
+I thank the citizens and scientists on Twitter who helped inspire this study and inform its background through their analyses and discussions of the early spread of SARS-CoV-2.
+I thank the scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.
 The scientific computing infrastructure used in this work was supported by the NIH Office of Research Infrastructure Programs under S10OD028685.

### 2021-06-18 paper/paper.tex
few more minor text tweaks

@@
 However, this theory became increasingly tenuous as it was learned that many early cases had no connection to the market~\citep{cohen2020wuhan,huang2020clinical, chen2020epidemiological}.
-Soon, Chinese CDC Director Gao Fu dismissed the theory, stating ``At first, we assumed the seafood market might have the virus, but now the market is more like a victim.
+Eventually, Chinese CDC Director Gao Fu dismissed the theory, stating ``At first, we assumed the seafood market might have the virus, but now the market is more like a victim.
 The novel coronavirus had existed long before''~\citep{GlobalTimes2020}.
@@
 
-However, the Wuhan sequences that are available have also confounded phylogenetic analyses designed to infer the ``progenitor'' of SARS-CoV-2~\citep{kumar2021evolutionary}.
+However, the Wuhan sequences that are available have also confounded phylogenetic analyses designed to infer the ``progenitor'' of SARS-CoV-2, which is the sequence from which all other currently known sequences are descended~\citep{kumar2021evolutionary}.
 Although there is debate about exactly how SARS-CoV-2 entered the human population, it is universally accepted that the virus's deep ancestors are bat coronaviruses~\citep{lytras2021exploring}.
-But the earliest known SARS-CoV-2 sequences, which are mostly derived from the Huanan Seafood Market, are notably more different from these bat coronaviruses than other SARS-CoV-2 sequences collected at later dates outside Wuhan.
+But the earliest known SARS-CoV-2 sequences, which are mostly derived from the Huanan Seafood Market, are notably more different from these bat coronaviruses than other sequences collected at later dates outside Wuhan.
 As a result, there is a direct conflict between the two major principles used to infer an outbreak's progenitor: namely that it should be among the earliest sequences, and that it should be most closely related to deeper ancestors~\citep{pipes2021assessing}.
@@
 \section{Competing interests}
-The author consults for Moderna on SARS-CoV-2 evolution and epidemiology, consults for Flagship Labs 77 on viral deep mutational scanning, and has the potential to receive a share of IP revenue as an inventor on a Fred Hutch licensed technology/patent (application WO2020006494) related to deep mutational scanning of viral proteins.
+The author consults for Moderna on SARS-CoV-2 evolution and epidemiology, consults for Flagship Labs 77 on viral evolution and deep mutational scanning, and has the potential to receive a share of IP revenue as an inventor on a Fred Hutch licensed technology/patent (application WO2020006494) related to deep mutational scanning of viral proteins.
 

### 2021-06-18 paper/paper.tex
add links to interactive plots to paper

@@
 Points are slightly jittered on the y-axis.
-Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
+Go to \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/deltadist.html} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
 Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
@@
 Mutational distance is calculated relative to proCoV2, and points are jittered on the y-axis.
-Go to {\color{red} add link} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of jittering.
+Go to \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/deltadist_jitter.html} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of jittering.
 \label{fig:deltadist_jitter}
@@
 
-The plot in Figure~\ref{fig:deltadist_RaTG13} contains all of the GISAID sequences after this filtering, and was rendered using \texttt{Altair}~\citep{vanderplas2018altair}.
+The plot in Figure~\ref{fig:deltadist_RaTG13} contains all of the GISAID sequences after this filtering.
 The plot in Figure~\ref{fig:deltadist_jitter} shows the filtered GISAID sequences collected before February of 2020 plus the 13 good coverage recovered partial early outpatient sequences (Table~\ref{tab:mutations}), considering only the region covered by the partial sequences (21,570 to 29,550).
@@
 The difference is that the sequences from the deleted data set are then added to each node with which they are compatible given their mutations in an amount proportional to the size of the node, the logic being that a sequence is more likely to fall into larger clusters.
+
+\subsection{Interactive versions of some figures}
+Interactive versions of some figures are available at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/}, and were created using \texttt{Altair}~\citep{vanderplas2018altair}
 
@@
 \renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{1}
+\setcounter{figure}{0}
 
@@
 Note that the y-axis uses a symlog scale.
-An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at {\color{red} add link}.
+An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/coverage_region.html}.
 A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Figure~\ref{suppfig:coverage_all}.
@@
 \caption{A version of Figure~\ref{fig:coverage} that shows coverage over the full length of the SARS-CoV-2 genome.
-An interactive version of this plot is at {\color{red} add link}.
+An interactive version of this plot is at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/coverage_all.html}.
 }

### 2021-06-23 paper/paper.tex
fixed typo noted by Len Granowetter

@@
 With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
-Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely using Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
+Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely used Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
 Although requiring coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations have coverage $\ge$10.

### 2021-06-27 paper/paper.tex
add Joe Dudley suggested wording change
update text to describe SRR11313490 and SRR11313499

@@
 
-I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February (files for SRR11313490 and SRR11313499 were not accessible via the cloud).
-I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for each SRA file.
+I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February.
+The SRA files for two runs (SRR11313490 and SRR11313499) were not accessible via the Google Cloud, but after I posted the first version of this manuscript as a pre-print, several individuals found archived data for these runs that had been downloaded when they were still available on the SRA, and I have used those data in the updated analysis described here (see Methods for details).
+I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for all SRA files.
 For all files, the object timestamp is February 15, 2020, and the time is March 16, 2020.
@@
 Therefore, it is impossible to unambiguously place them phylogenetically, or determine exactly when they were collected.
-However, little can be done to mitigate this caveat beyond my failed attempt to contact the corresponding authors of \citet{wang2020medRxiv}.
+However, little can be done to mitigate this caveat beyond my unsuccessful attempt to contact the corresponding authors of \citet{wang2020medRxiv}.
 It is also important to note that my phylogenetic analyses use relatively simple methods to draw qualitative conclusions without formal statistical testing.
@@
 Note also that as described in this paper's main text, two SRA files could not be downloaded from the Google Cloud using the aforementioned method, and so are not part of this study.
+
+\subsection{FASTQ files for SRR11313490 and SRR11313499}
+I was not able to download SRA files for two runs, SRR11313490 and SRR11313499, from the Google Cloud.
+After I posted the initial version of this pre-print, I was contacted by three different individuals (Andrew Fire -- Stanford University, Theo Sanderson -- Crick Institute, Francisco de Asis de Ribera -- DRASTIC Research) who had uncovered FASTQ or FASTA files for these two missing runs that appeared to have been downloaded from the SRA prior to the deletion of PRJNA612766.
+For revised versions of the manuscript, I included analysis of the data for these two runs that I obtained from the links provided at \url{https://github.com/lifebit-ai/SARS-CoV-2/blob/master/assets/ucsc/aws_https_links.txt}.
+Inclusion of data for these two runs did not appreciably change the results of the analysis relative to that in the original version of the pre-print, since both runs corresponded to low coverage samples for which meaningful viral genetic information could not be obtained.
 

### 2021-06-27 paper/paper.tex
describe masking primer binding sites

@@
 }
-\caption{Samples for which the SARS-CoV-2 sequence could be called at $\ge$95\% of sites between 21,570 and 29,550, and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
+\caption{Samples for which the SARS-CoV-2 sequence could be called at $\ge$90\% of sites between 21,570 and 29,550, and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
 Numbers in parentheses after each substitution give the deep sequencing reads with each nucleotide identity.
@@
 
-I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
+I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample, and masking regions that corresponded to the primer binding sites described in \citet{wang2020medRxiv}.
 Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
@@
 I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
-With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
+With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$90\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
 Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely used Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
@@
 I was not able to download SRA files for two runs, SRR11313490 and SRR11313499, from the Google Cloud.
-After I posted the initial version of this pre-print, I was contacted by three different individuals (Andrew Fire -- Stanford University, Theo Sanderson -- Crick Institute, Francisco de Asis de Ribera -- DRASTIC Research) who had uncovered FASTQ or FASTA files for these two missing runs that appeared to have been downloaded from the SRA prior to the deletion of PRJNA612766.
+After I posted the initial version of this pre-print, I was contacted by three different individuals (see Acknowledgments) who had uncovered FASTQ or FASTA files for these two missing runs that were downloaded from the SRA prior to the deletion of PRJNA612766.
 For revised versions of the manuscript, I included analysis of the data for these two runs that I obtained from the links provided at \url{https://github.com/lifebit-ai/SARS-CoV-2/blob/master/assets/ucsc/aws_https_links.txt}.
@@
 I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to determine the coverage at each site by aligned nucleotides with a quality score of at least 20.
+I also masked (set to zero coverage) all sites overlapped by the primers used by \citet{wang2020medRxiv} to PCR the amplicons used for sequencing.
 These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
@@
 
-I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$95\% of the sites in the region of interest from site 21,570 to 29,550.
+I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$90\% of the sites in the region of interest from site 21,570 to 29,550.
 These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
@@
 I thank the citizens and scientists on Twitter who helped inspire this study and inform its background through their analyses and discussions of the early spread of SARS-CoV-2.
+I thank Brendan Larsen for pointing out that I failed to mask the primer binding site regions in the analysis described in the original pre-print (this has been fixed in this version).
+I thank Andrew Fire (Stanford University), Theo Sanderson (Crick Institute), and Francisco de Asis de Ribera (DRASTIC Research) for finding locations of downloaded data for the two accessions (SRR11313490 and SRR11313499) that I could not download from the Google Cloud and so had to omit from the analysis described in the original pre-print (they are included in the analysis in this version).
 I thank the scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.

### 2021-06-27 paper/paper.tex
update tables / figures with coverage after masking primer sites

@@
 \midrule
-A4 &                               0.9827 &        early outpatient &                                                                          none \\
-C1 &                               0.9966 &        early outpatient &  G22081A (A=924, C=4, G=9), C28144T (C=6, T=1185), T29483G (C=1, G=45, T=1) \\
-C2 &                               0.9962 &        early outpatient &                                                 C29095T (C=1, G=1, T=751)  \\
-C9 &                               0.9536 &        early outpatient &                                  C28144T (C=3, T=823), G28514T (G=1, T=36)  \\
-D9 &                               0.9585 &        early outpatient &                                                     C28144T (C=4, T=1653)  \\
-D12 &                               0.9970 &        early outpatient &                                                     C28144T (C=8, T=2400)  \\
-E1 &                               0.9759 &        early outpatient &                                                           C28144T (T=125)  \\
-E5 &                               0.9758 &        early outpatient &              C24034T (A=5, C=3, T=74), T26729C (C=12), G28077C (C=142, G=4)  \\
-E11 &                               0.9877 &        early outpatient &                                 C25460T (C=2, T=246), C28144T (C=1, T=412)  \\
-F11 &                               0.9594 &        early outpatient &                             T25304A (A=9, T=1), C28144T (C=6, G=1, T=1328)  \\
-G1 &                               0.9959 &        early outpatient &                                   none                                        \\
-G11 &                               0.9677 &        early outpatient &                                  none                                         \\
-H9 &                               0.9941 &        early outpatient &                                                     C28144T (C=2, T=1254)  \\
-R11 &                               0.9987 &  hospital patient (Feb) &                               C21707T (T=401), C28144T (A=1, C=18, T=4265)  \\
+A4 &                               0.9266 &        early outpatient &                                                                          none \\
+C1 &                               0.9396 &        early outpatient &  G22081A (A=924, C=4, G=9), C28144T (C=6, T=1185), T29483G (C=1, G=45, T=1) \\
+C2 &                               0.9397 &        early outpatient &                                                 C29095T (C=1, G=1, T=751)  \\
+C9 &                               0.9005 &        early outpatient &                                  C28144T (C=3, T=823), G28514T (G=1, T=36)  \\
+D9 &                               0.9051 &        early outpatient &                                                     C28144T (C=4, T=1653)  \\
+D12 &                               0.9400 &        early outpatient &                                                     C28144T (C=8, T=2400)  \\
+E1 &                               0.9223 &        early outpatient &                                                           C28144T (T=125)  \\
+E5 &                               0.9227 &        early outpatient &              C24034T (A=5, C=3, T=74), T26729C (C=12), G28077C (C=142, G=4)  \\
+E11 &                               0.9321 &        early outpatient &                                 C25460T (C=2, T=246), C28144T (C=1, T=412)  \\
+F11 &                               0.9054 &        early outpatient &                             T25304A (A=9, T=1), C28144T (C=6, G=1, T=1328)  \\
+G1 &                               0.9396 &        early outpatient &                                   none                                        \\
+G11 &                               0.9112 &        early outpatient &                                  none                                         \\
+H9 &                               0.9381 &        early outpatient &                                                     C28144T (C=2, T=1254)  \\
+R11 &                               0.9422 &  hospital patient (Feb) &                               C21707T (T=401), C28144T (A=1, C=18, T=4265)  \\
 \bottomrule
@@
 \includegraphics[width=\linewidth]{figures/coverage_all.pdf}
-\caption{A version of Figure~\ref{fig:coverage} that shows coverage over the full length of the SARS-CoV-2 genome.
+\caption{A version of Figure~\ref{suppfig:coverage} that shows coverage over the full length of the SARS-CoV-2 genome.
 An interactive version of this plot is at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/coverage_all.html}.

### 2021-06-27 paper/paper.tex
remove reference to pangolin accession example

@@
 I recover the deleted files from the Google Cloud, and reconstruct partial sequences of 13 early epidemic viruses.
-Phylogenetic analysis of these sequences in the context of carefully annotated existing data suggests that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
+Phylogenetic analysis of these sequences in the context of carefully annotated existing data supports existing data indicating that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
 Instead, the progenitor of known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
@@
 The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
-An example of this process from another study is in Figure~\ref{fig:pangolin_emails}, which shows an e-mail by the lead author of a paper on pangolin coronaviruses~\citep{xiao2020isolation} requesting deletion of two sequencing runs.
-Subsequent to March 30, 2020, a similar e-mail request must have been made to fully delete SARS-CoV-2 deep sequencing project PRJNA612766.
-
-\begin{figure}[]
-\centering
-\fbox{\includegraphics[trim={0in 0in 0in 0in},clip,width=\linewidth]{figures/pangolin_emails.jpg}}
-\caption{Example of the process to delete SRA data.
-The image shows e-mails between the lead author of the pangolin coronavirus paper~\citet{xiao2020isolation} and SRA staff excerpted from \citet{USRTK2020}.
-}
-\label{fig:pangolin_emails}
-\end{figure}
 
@@
 Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.
-In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions, which are already known to include SRR11119760 and SRR11119761~\citep{USRTK2020}.
+In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions.
 Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for its staff to validate the rationale for all requests.

### 2021-06-28 paper/paper.tex
add Wuhan University e-mail

@@
 
+\begin{figure}
+\centerline{\fbox{\includegraphics[width=0.95\linewidth]{figures/SRA_email.png}}}
+\caption{A redacted version of the e-mails from Wuhan University to the SRA staff requesting deletion of the sequencing data.
+This e-mail was provided to me by the NIH's NCBI Director Stephen Sherry on June, 19, 2021, the day after I e-mailed the NIH an advance copy of this manuscript.
+The redactions and highlighting were done by the NIH, and I am showing the e-mail exactly as it was provided to me.
+}
+\label{fig:SRA_email}
+\end{figure}
+
 The fact that such an informative data set was deleted has implications beyond those gleaned directly from the recovered sequences.
 Samples from early outpatients in Wuhan are a gold mine for anyone seeking to understand spread of the virus.
-Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than surreptitiously delete the partial sequences.
-There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
-It therefore seems likely the sequences were deleted to obscure their existence.
-Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to trace early spread of the epidemic.
+Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than delete the partial sequence data.
+There is no scientific reason for the deletion: the sequences are concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
+After I e-mailed the NIH the original version of this manuscript, they sent me the e-mails requesting deletion of the data, which are in Figure~\ref{fig:SRA_email}.
+While I cannot rule out that the authors posted the data on some unknown website, they are not in GISAID, NCBI, or any database used by the joint WHO-China report.
+Therefore, even though the data were on the Google Cloud (as described above) and the mutations were listed in a table in the \textit{Small} paper by \citet{wang2020small}, the practical consequence of removing the data from the SRA was that nobody was aware of its existence.
+Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to understand the early spread of the epidemic.
 

### 2021-06-28 paper/paper.tex
shorten discussion of two markets, eliminate that of RaTG13

@@
 
-Before moving to the next section, I will also briefly address two less plausible explanations for the discordance between phylogenetic rooting and epidemiological data that have gained traction in discussion of SARS-CoV-2's origins.
-The first explanation, which has circulated on social media, suggests that the RaTG13 sequence might be faked in a way that confounds phylogenetic inference of SARS-CoV-2's progenitor.
-But although there are unusual aspects of RaTG13's primary sequencing data~\citep{singla2020novo,rahalkar2020anomalous}, the conundrum about inferring the progenitor holds for other outgroups such as RpYN06, RmYN02, and more distant bat coronaviruses reported before emergence of SARS-CoV-2 such as ZC45~\citep{tang2020origin}.
-The second explanation, which was proposed in a blog post by \citet{garry2021early} and amplified by a popular podcast~\citep{twiv2021}, is that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other market being the source of viruses that lack the T8782C and C28144T mutations.
-However, inspection of Figure~\ref{fig:tree_RaTG13} shows that clade B is connected to viruses lacking T8782C and C28144T by single mutational steps via other human isolates, so this explanation requires not only positing two markets with two progenitors differing by just two mutations, but also the exceedingly improbable evolution of one of these progenitors towards the other after it had jumped to humans.
+Another explanation that I consider less plausible is offered by \citet{garry2021early}: that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other market being the source of viruses that lack T8782C and C28144T (Figure~\ref{fig:tree_RaTG13}).
+However, this explanation requires positing zoonoses in two markets by two progenitors differing by just two mutations, which seems non-parsimonious in the absence of direct evidence for zoonosis in any market.
 

### 2021-06-28 paper/paper.tex
more text revisions

@@
 I recover the deleted files from the Google Cloud, and reconstruct partial sequences of 13 early epidemic viruses.
-Phylogenetic analysis of these sequences in the context of carefully annotated existing data supports existing data indicating that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
-Instead, the progenitor of known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
+Phylogenetic analysis of these sequences in the context of carefully annotated existing data further supports the idea that the Huanan Seafood Market sequences are not fully representative of the viruses in Wuhan early in the epidemic.
+Instead, the progenitor of currently known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
 \end{abstract}
@@
 
-Here I take a step towards resolving these questions by identifying and recovering a deleted data set of partial SARS-CoV-2 sequences from outpatient samples collected early in the Wuhan epidemic.
+Here I take a small step towards resolving these questions by identifying and recovering a deleted data set of partial SARS-CoV-2 sequences from outpatient samples collected early in the Wuhan epidemic.
 Analysis of these new sequences in conjunction with careful annotation of existing ones suggests that the early Wuhan samples that have been the focus of most studies including the joint WHO-China report~\citep{WHO2021origins} are not fully representative of the viruses actually present in Wuhan at that time.
-These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to that inferred by \citet{kumar2021evolutionary}.
+These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to that inferred by \citet{kumar2021evolutionary} and the other of which contains the C29095T mutation.
 Furthermore, the approach taken here hints it may be possible to advance understanding of SARS-CoV-2's origins or early spread even without further on-the-ground studies, such as by more deeply probing data archived by the NIH and other entities.
@@
 Importantly, the study notes that this approach yields information about the sequence of the virus as well enabling diagnosis of infection.
+In fact, \citet{wang2020small} even list the mutations determined from this sequencing---but because this paper was published in the chemistry journal \textit{Small}, after the sequences were removed from the SRA their existence appears to have been entirely overlooked.
 
@@
 
-Neither the pre-print~\citep{wang2020medRxiv} nor published manuscript~\citep{wang2020small} contain any correction or note that indicates a scientific reason for deleting the study's sequencing data from the SRA.
+Neither the pre-print~\citep{wang2020medRxiv} nor published manuscript~\citep{wang2020small} contain any correction or note that suggests a scientific reason for deleting the study's sequencing data from the SRA.
 I e-mailed both corresponding authors of \citet{wang2020medRxiv} to ask why they had deleted the deep sequencing data and to request details on the collection dates of the early outpatient samples, but received no reply.
@@
 In addition, the mutations I called from the raw sequence data in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.
+Again, this fact emphasizes that the information contained in the deleted sequencing data is largely present in \citet{wang2020small}, but because it was only published in a table in the chemistry journal \textit{Small} rather than placed on the SRA, its existence was overlooked.
 
@@
 Analysis of these sequences leads to several conclusions.
-First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
+First, they provide further evidence Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
 The deleted data as well as existing sequences from Wuhan-infected patients hospitalized in Guangdong show early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than sequences in the joint WHO-China report~\citep{WHO2021origins}.
@@
 Crucially, both putative progenitors are three mutations closer to SARS-CoV-2's bat coronavirus relatives than sequences from the Huanan Seafood Market.
-Note also that the progenitor of all known SARS-CoV-2 sequences could still be downstream of the sequence that infected patient zero depending on the transmission dynamics of the first infections.
+Note also that the progenitor of all known SARS-CoV-2 sequences could still be downstream of the sequence that infected patient zero---and it is possible that the future discovery of additional early SARS-CoV-2 sequences could lead to further revisions of inferences about the earliest viruses in the outbreak.
 
 \begin{figure}
-\centerline{\fbox{\includegraphics[width=0.95\linewidth]{figures/SRA_email.png}}}
+\centerline{\fbox{\includegraphics[width=\linewidth]{figures/SRA_email.png}}}
 \caption{A redacted version of the e-mails from Wuhan University to the SRA staff requesting deletion of the sequencing data.
@@
 
-The fact that such an informative data set was deleted has implications beyond those gleaned directly from the recovered sequences.
+The fact that this informative data set was deleted suggests implications beyond those gleaned directly from the recovered sequences.
 Samples from early outpatients in Wuhan are a gold mine for anyone seeking to understand spread of the virus.
-Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than delete the partial sequence data.
-There is no scientific reason for the deletion: the sequences are concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
+Even my analysis of 13 partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence all 34 samples rather than delete the partial sequence data.
+There is no obvious scientific reason for the deletion: the sequences are concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
 After I e-mailed the NIH the original version of this manuscript, they sent me the e-mails requesting deletion of the data, which are in Figure~\ref{fig:SRA_email}.
-While I cannot rule out that the authors posted the data on some unknown website, they are not in GISAID, NCBI, or any database used by the joint WHO-China report.
-Therefore, even though the data were on the Google Cloud (as described above) and the mutations were listed in a table in the \textit{Small} paper by \citet{wang2020small}, the practical consequence of removing the data from the SRA was that nobody was aware of its existence.
-Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to understand the early spread of the epidemic.
+While I cannot rule out that the authors posted the data on some unknown website, the sequences are not in GISAID, NCBI, or any database used by the joint WHO-China report.
+Therefore, even though the sequencing data were on the Google Cloud (as described above) and the mutations were listed in a table in the \textit{Small} paper by \citet{wang2020small}, the practical consequence of removing the data from the SRA was that nobody was aware these sequences existed.
+Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to maximize information about viral sequences from early in the Wuhan epidemic.
 
@@
 More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without an international investigation.
-Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.
-In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions.
+I suggest it could be worthwhile for the NIH to review e-mail records to identify other SRA deletions.
 Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for its staff to validate the rationale for all requests.
-However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
-A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the virus.
+However, the current study suggests that at least in one case, the trusting structures of science permitted a data deletion that obscured sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
+A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed light on additional overlooked information relevant to the early emergence of the virus.
 
@@
 I thank Andrew Fire (Stanford University), Theo Sanderson (Crick Institute), and Francisco de Asis de Ribera (DRASTIC Research) for finding locations of downloaded data for the two accessions (SRR11313490 and SRR11313499) that I could not download from the Google Cloud and so had to omit from the analysis described in the original pre-print (they are included in the analysis in this version).
+I thank Stephen Goldstein and anonymous commenter for providing critiques of the original version of the pre-print on \textit{bioRxiv} that have helped me revise the manuscript.
 I thank the scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.

### 2021-06-28 paper/paper.tex
update acknowledgements

@@
 I was not able to download SRA files for two runs, SRR11313490 and SRR11313499, from the Google Cloud.
-After I posted the initial version of this pre-print, I was contacted by three different individuals (see Acknowledgments) who had uncovered FASTQ or FASTA files for these two missing runs that were downloaded from the SRA prior to the deletion of PRJNA612766.
+After I posted the initial version of this pre-print, I was contacted by three different individuals who had uncovered FASTQ or FASTA files for these two missing runs that were downloaded from the SRA prior to the deletion of PRJNA612766.
 For revised versions of the manuscript, I included analysis of the data for these two runs that I obtained from the links provided at \url{https://github.com/lifebit-ai/SARS-CoV-2/blob/master/assets/ucsc/aws_https_links.txt}.
@@
 I thank Brendan Larsen for pointing out that I failed to mask the primer binding site regions in the analysis described in the original pre-print (this has been fixed in this version).
-I thank Andrew Fire (Stanford University), Theo Sanderson (Crick Institute), and Francisco de Asis de Ribera (DRASTIC Research) for finding locations of downloaded data for the two accessions (SRR11313490 and SRR11313499) that I could not download from the Google Cloud and so had to omit from the analysis described in the original pre-print (they are included in the analysis in this version).
 I thank Stephen Goldstein and anonymous commenter for providing critiques of the original version of the pre-print on \textit{bioRxiv} that have helped me revise the manuscript.

### 2021-07-19 paper/reviewer_response.tex
added reviewer comments

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb}
\usepackage{hyperref} 
\usepackage[round,semicolon]{natbib}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
\newcommand{\response}[1]{{\color{black}#1}}

\title{Response to reviews of ``Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic'' for \textit{Molecular Biology and Evolution}}
\author{}

\begin{document}
\maketitle

\emph{Below, the reviewer and editor comments {\color{blue} are in blue}, and may responses are in black.}

\color{blue}

\subsection*{Editors’ comments to the author:}

I have received two reviews of your paper.  Both reviewers encourage revisions of the paper, but as you will see from the reviews, one is less positive and provide very few technical comments.  This second review is somewhat unusual as it contains very few suggestions for revisions, but it includes instead a number of value judgements of statements in the paper.  It might be difficult to address these comments directly, but I will encourage you nonetheless to try to do so.  In addition, please also address all other comments by the first reviewer.  I have also read the paper myself, and my major suggestion is for you not to imply similarity with the ancestral sequence inferred by Kumar et al. as a type of validation.  This method in Kumar et all is essentially a form of outgroup rooting that uses an outgroup to infer the sequence in the root of the tree.  So when you use outgroup rooting, it is no surprise that the implied sequence is similar to the one found by Kumar et al.

\subsection*{Reviewer: 1}

Comments to the Author
The author identifies a set of SARS-CoV-2 sequences from the early days of the pandemic which has been deleted from the SRA database. He manages to recover the sequence files (but not the associated metadata) and shows that they are more similar to SARS-CoV-2's bat coronavirus relatives than sequences associated to the Huanan market, suggesting that the progenitor of all known SARS-CoV-2 sequences was three mutations away from the Huanan market sequences.

Even though the phylogeny reconstruction is simple and mostly qualitative, and does not include any statistical test to evaluate the statistical robustness of each tree, the analysis is worth publishing as it is. Indeed, the phylogenetic signal derived from early SARS-CoV-2 sequences is unavoidably low, with only a few informative mutations. Establishing the root of such starting outbreak is thus a difficult and complex problem, and we cannot expect the current paper to solve this issue with a few extra analyses. As highlighted by the author, it will be important in the future to take into account the information from metadata, such as epidemiological relationships between patients, date of contamination, date of sampling, etc. to better assess the root of all known SARS-CoV-2 sequences. I hope that the COVID-19 pandemic will lead to new approaches (such as the one developped by Kumar et al.) and new statistical methods to infer the root of sequences from an outbreak.

The text is very well written. The results have already been widely covered in the media, both for the general public and in scientific journals such as Nature and Science. The main findings are (1) that several informative early SARS-CoV-2 sequences have been removed from databases and (2) that these sequences provide further support that circulation within the human population started at least a few weeks before the Huanan market outbreak (three mutations away).

Minor comments

1) In the introduction the author should make it clearer that the dataset he recovered has not been used in recent thorough phylogenetic analyses of the SARS-CoV-2 outbreak: Pipes et al. 2021; Morel et al. 2021; Kumar et al. 2021; Turakhia et al. 2020; Pekar et al. 2021.

2) In the discussion or the results section it should also be mentioned that there is no mechanism for deleting a pre-print (unlike for SRA data). So the authors were committed to having their preprint permanently available after it was posted in March.

3) page 4, right column, line 54: it is mentioned that the author limited "the analysis to sequences with at least two observations among strains". To include potentially more sequences into the analysis and still take into account sequencing errors, would it be possible to exclude sites (when a particular nucleotide is found in only one single sequence) on specific sequences and thus to keep the corresponding sequence into the analysis (as in Kumar et al. 2021)?

4) In addition to Figures 2 and 4 it would be informative to show a haplotype network combining these early SARS-CoV-2 sequences and RaTG13, with different colors according to where the virus sequences were collected as in Figure 5.

Suggestions for manuscript text and figures revision

page 4, left column, line 43: "plasmid contamination did not afflict the viral samples in the deleted sequencing project" seems a bit long and complicated. Maybe replace by something like "plasmid contamination of viral samples is unlikely".

page 4, Fig. 2 legend: replace "false/true" by something like "associated with Huanan Seafood Market/not associated with Huanan Seafood Market".

page 4, Fig. 2: The distinction between squares and circles is hard to see on the Figure. One suggestion is to put the data points associated with the market in black.

page 4, Fig. 2: The legend reads "Mutational distances are relative to the putative progenitor proCoV2 inferred by Kumar et al. (2021)." This is a bit confusing. I am guessing that the author means that the mutational distance of the putative progenitor proCoV2 inferred by Kumar et al. (2021) is set to zero and other values are adjusted accordingly. My suggestion would be to show the absolute mutational distance from the outgroup RaTG13 (even if it is in the order of 1000) and to indicate the position of proCoV2 in the figure or in the legend (with an arrow for example). But I don't feel strongly about this suggestion.

page 6, Fig. 4 legend: same comment as above, it is surprising to see values below 0.

page 4, right column, line 45-47: I would suggest to remove "a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coron- avirus into humans" as this is redundant with the previous paragraph and it is better in the results section to focus on observations.

page 5, right column, line 42-45: The following sentence: "However, neither the sophisticated algorithm of Kumar et al. (2021) nor my more simplistic approach explain why the progenitor should be so different from the earliest sequences reported from Wuhan" is not clear. My suggestion would be something likee: "The sophisticated algorithm of Kumar et al. (2021) and my more simplistic approach do not take into account sampling dates and they both find that the inferred progenitor is three mutations away from the earliest sequences reported from Wuhan. A possible explanation is that the sampling of early cases is biased and not representative of the initial dynamics of the outbreak."

page 6, Figure 4: instead of jittering points, another possibility would be to show circles whose area is proportional to the number of observations.

page 6, left column, lines 52-56: I would suggest to remove "This fact suggests that the market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report (WHO 2021), are not representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020." as this should go in to the discussion part and is not useful here.

page 6, right column, line 6: "The rooting of the middle tree in Figure 5 is now highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node." This statement is rather qualitative so maybe change  "is now highly plausible" into something like "is now more supported than the previous one in Figure 3".

page 6, right column, line 33: I would suggest to replace "valid concerns about the nature of the underlying data" by "valid concerns about the sampling distribution of the underlying data".

page 9, right colmun, line 11: remove "from" in "I downloaded all sequences from collected prior to March of 2020"

page 11, left column, line 27: The hyper link for \url{https://web.archive.org/web/20210103124552/https://www.scmp.com/news/china/society/article/3084635/china-confirms-unauthorised-labs-were-told-destroy-early}  contains an extra space that should be removed.

\subsection*{Reviewer: 2}

Comments to the Author
The primary shortcoming of this work is its significance. The author has recovered from Google Cloud raw sequence files that had been deleted from the SRA at the request of the authors of Wang et al, cited in the manuscript. However, the relevant data points, notably the SARS-CoV-2 SNPs that distinguish early lineages A and B, are reported in that paper, published June 2020 in Small. Additionally, it was already known that two lineages of SARS-CoV-2 were present in Wuhan early in the epidemic. These sequences confirm that both lineages circulated contemporaneously in late 2019-early 2020 in Wuhan, but do not change the timeline of SARS-CoV-2 introduction nor provide new information as to the route or routes of introduction. These sequences post-date the likely introduction of SARS-CoV-2 into the human population by several weeks up to 2 months, and thus do not provide new information regarding the earliest human infection events.

The author casts doubt on the timing and distribution of early cases based on media reports of unknown reliability. It is impossible for the reader to interpret this.

The author describes a conundrum whereas the first sequences (Dec 2019) seem more distant from related bat viruses than do later sequences (Jan-Feb 2020), suggesting this is nearly impossible. However, sequencing is highly biased to case identification, and it is known that lineage B infections were preferentially identified due to the amplification of early transmission at the Huanan market. Additionally, this presupposes early diversity in SARS-CoV-2 is due to the emergence of SNPs during human to human transmission, rather than in a putative intermediate host. The author additionally notes that "unusual mutational processes" might explain an ancestral position for lineage B, and indeed C>T, T>C mutations are frequently observed in SARS-CoV-2, complicating the inference of directional relationships between viral lineages dependent on these SNPs.

Rooting SARS-CoV-2 trees continues to be problematic but the author's efforts do not resolve this problem. The available bat viruses are too distant to reliably root the tree and while the proCoV-2 sequence the author also uses is appropriately close, it was inferred based on the too-distant bat viruses, making this a somewhat circular process. There is simply too much uncertainty currently, and it may be as likely as otherwise that the root falls between lineages A and B, rather than in either one of them. Alternatively, the author cannot exclude A>B directionality occurring in an intermediate host rather than in humans, unless one discounts the possibility of an intermediate host. The author acknowledges the problematic nature of the rooting issue, but nevertheless heavily weights the conclusions on this analysis.

The author states that multiple spillover events is a less parsimonious start of the outbreak than a single introduction but does not explain this analysis. The phylogenetic analysis would be greatly strengthened by a more thorough discussion of this matter. If A>B directionality in humans is presumed, it depends on a single introduction event, which the author should make the case for.

Figure 6 does not add to the data analysis and only serves to drive home what seems to be the primary objective of the manuscript - that the sequences were deleted from the SRA.

The acknowledgement of "twitter citizens" risks highlighting commentary by individuals who have directed abuse and threats at the author's colleagues. I encourage the author to reconsider, or specify who he refers to in particular in order to avoid granting inadvertent credibility to nefarious individuals.

\color{black}
\bibliographystyle{mbe}
{\small
\bibliography{references.bib}
}

\end{document}

### 2021-07-19 paper/reviewer_response.tex
responded to some of reivewer 1 minor comments

@@
 So when you use outgroup rooting, it is no surprise that the implied sequence is similar to the one found by Kumar et al.
+
+\response{Thanks for the helpful comments.
+I have addressed all of the comments from Reviewer \#1 and all the comments from Reviewer \#2 that I am able to address in the context of the limitations of the study.
+Also, thanks for the point about Kumar et al's proCoV2 inference also essentially being an outgroup method.
+I still mention this prominently to credit their excellent work, but also make clear that there are some shared assumptions of the two methods, so the concordance does not represent a validation of those assumptions but only an indication of reaching similar inferences under some of the same assumptions.}
 
@@
 
+\response{This is a good suggestion, and I have removed the sentence in question.}
+
 page 6, right column, line 6: "The rooting of the middle tree in Figure 5 is now highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node." This statement is rather qualitative so maybe change  "is now highly plausible" into something like "is now more supported than the previous one in Figure 3".
+
+\response{This is a good suggestion, and I have re-worded the statement in question.}
 
@@
 
+\response{This is a good suggestion, and I have made the change.}
+
 page 9, right colmun, line 11: remove "from" in "I downloaded all sequences from collected prior to March of 2020"
 
+\response{I have fixed this typo.}
+
 page 11, left column, line 27: The hyper link for \url{https://web.archive.org/web/20210103124552/https://www.scmp.com/news/china/society/article/3084635/china-confirms-unauthorised-labs-were-told-destroy-early}  contains an extra space that should be removed.
+
+\response{This link is fixed, at least in the PDF as it is rendered on my computer.}
 

### 2021-07-19 paper/paper.tex
responded to some of reivewer 1 minor comments

@@
     \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages
-
-%\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
-%\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references
-
-\articletype{inv}
 
@@
 Repeating the analysis of the previous section with these changes shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
-This fact suggests that the market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, are not representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
 
 Furthermore, it is immediately apparent that the discrepancy between outgroup rooting and the evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong.
-The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} is now highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
+The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} is now more consistent with the evidence that the pandemic originated in Wuhan, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
 The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
@@
 Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
-I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the nature of the underlying data.
+I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the sampling distribution of the underlying data.
 
@@
 \subsection{Curation and analysis of early SARS-CoV-2 sequences from GISAID}
-For the broader analyses of existing SARS-CoV-2 sequences, I downloaded all sequences from collected prior to March of 2020 from GISAID.
+For the broader analyses of existing SARS-CoV-2 sequences, I downloaded all sequences collected prior to March of 2020 from GISAID.
 The accessions of these sequences are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/accessions.txt}, and a table acknowledging the labs and authors is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/acknowledgments.csv}.

### 2021-07-20 paper/reviewer_response.tex
made reviewer 1 minor edits that do not require figure adjustment

@@
 
+\response{This is a good suggestion, and I have made the recommended change.}
+
 page 4, Fig. 2 legend: replace "false/true" by something like "associated with Huanan Seafood Market/not associated with Huanan Seafood Market".
@@
 
-page 4, right column, line 45-47: I would suggest to remove "a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coron- avirus into humans" as this is redundant with the previous paragraph and it is better in the results section to focus on observations.
+page 4, right column, line 45-47: I would suggest to remove "a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans" as this is redundant with the previous paragraph and it is better in the results section to focus on observations.
+
+\response{This is a good suggestion, and I have made the recommended change.}
 
@@
 (2021) nor my more simplistic approach explain why the progenitor should be so different from the earliest sequences reported from Wuhan" is not clear.
-My suggestion would be something likee: "The sophisticated algorithm of Kumar et al.
+My suggestion would be something like: "The sophisticated algorithm of Kumar et al.
 (2021) and my more simplistic approach do not take into account sampling dates and they both find that the inferred progenitor is three mutations away from the earliest sequences reported from Wuhan.
@@
 
+\response{This is a good suggestion, and I have replaced the original sentence with a sentence similar to the first one suggested by the reviewer.
+I have deferred the second sentence suggested by the reviewer to the Discussion.}
+
 page 6, Figure 4: instead of jittering points, another possibility would be to show circles whose area is proportional to the number of observations.
+
+\response{This is an interesting suggestion, and probably makes the static plot in Figure 4 better!
+However, I have kept the jittered format as my hope is that readers will actually go to the interactive version of the plot linked in the Figure 4 legend (\url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/deltadist_jitter.html}), and the interactive plot is better as jittered points as it allows mousing over points to identify specific sequences, which could not be done with the aggregated circle approach.
+Given this fact, I think it's better to keep the static plot in the same format as the interactive one.}
 

### 2021-07-20 paper/paper.tex
made reviewer 1 minor edits that do not require figure adjustment

@@
 I also determined the consensus sequence of the plasmid control used by \citet{wang2020medRxiv} from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
-Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), plasmid contamination did not afflict the viral samples in the deleted sequencing project.
+Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), plasmid contamination of the viral samples is unlikely.
 
@@
 The discrepancy is especially pronounced for sequences from patients who had visited the Huanan Seafood Market~\citep{WHO2021origins}.
-All sequences associated with this market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
+All sequences associated with this market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13}).
 Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
@@
 Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the outgroup.
-However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
+However, the sophisticated algorithm of \citet{kumar2021evolutionary} and my more simplistic approach both ignore sampling dates, which is probably a major reason that they both infer a progenitor that is three mutations away from the earliest sequences reported from Wuhan.
 

### 2021-07-21 paper/reviewer_response.tex
better explain relative mutational distances

@@
 
+\response{I see the rationale behind this suggestion.
+But after experimenting with this, I have decided it is actually more confusing because it gets hard to read off the numbers on the y-axis when they are much larger.
+Instead, I have made much clearer in the legend what is plotted, and also given the distance of the proCoV2 reference to the outgroup (1,132 mutations).
+In addition, I have explained that the point below zero corresponds to a sequence (Guangdong/FS-30-P00502/2020) reportedly collected in late February that is actually two mutations more similar to RaTG13 than proCoV2.}
+
 page 6, Fig. 4 legend: same comment as above, it is surprising to see values below 0.
+
+\response{See my response to the previous point.
+There are more points $< 0$ because the distances are only computed over the portion of the SARS-CoV-2 genome spanned by the recovered sequences, which I have now explained more clearly in the legend.}
 

### 2021-07-21 paper/paper.tex
better explain relative mutational distances

@@
 \caption{The reported collection dates of SARS-CoV-2 sequences in GISAID versus their relative mutational distances from the RaTG13 bat coronavirus outgroup.
-Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
-The plot shows sequences in GISAID collected no later than February 28, 2020.
+Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}, which itself differs from RaTG13 by 1,132 mutations---so a sequence with a relative mutational distance of 2 actually has 1,134 differences from RaTG13.
+Note that the lower-right point in the middle (green) panel corresponds to a sequence (Guangdong/FS-30-P00502/2020) reportedly collected in late February that is actually two mutations more similar to RaTG13 than proCoV2.
+The plot shows only sequences in GISAID collected no later than February 28, 2020.
 Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
@@
 Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (21,570--29,550).
+Because the calculated distances here are only over a portion of the genome, there are more negative points than in Figure~\ref{fig:deltadist_RaTG13}.
 The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.

### 2021-07-31 paper/paper.tex
filter to remove rare mutations rather than rare variants

@@
  \hspace{0.02\linewidth}
- \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
+ \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
   \hspace{0.02\linewidth}
@@
  \caption{
- Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected before Februrary, 2020.
+ Phylogenetic trees of SARS-CoV-2 sequences in GISAID collected before February, 2020.
  The trees are identical except they are rooted to make the progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
@@
 The monophyletic group containing C28144T is collapsed into a node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
+Singleton mutations (mutations observed only once in the sequence set) are removed as described in more detail in the Methods.
+ Figures~\ref{suppfig:tree_RpYN06} and \ref{suppfig:tree_RmYN02} shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
 \label{fig:tree_RaTG13}
@@
 This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the progenitor sequence is closest to the bat coronavirus outgroup.
-If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
+If we limit the analysis to sequences collected no later than January 2020 and remove singleton mutations observed only once in the sequence set, then there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06}, Figure~\ref{suppfig:tree_RmYN02}).
 Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
@@
  \hspace{0.02\linewidth}
- \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
+ \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
   \hspace{0.02\linewidth}
@@
 The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
+ Figures~\ref{suppfig:tree_RpYN06} and \ref{suppfig:tree_RmYN02} demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
 \label{fig:tree_with_deleted}
@@
 The phylogenetic trees were inferred using the GISAD sequences after the filtering and annotations described above, only considering sequences with $\ge$95\% coverage over the region of interest that were collected before February of 2020.
-In addition, after generating this sequence set I removed any sequence variants with a combination of mutations that was not observed at least twice so the analysis only includes multiply observed sequence variants.
+In addition, after generating this sequence set I removed any mutations from sequences that were singletons in the sense that they were only observed once in the sequence set---a similar approach was used by \citet{kumar2021evolutionary}, and this approach should eliminate rare mutations that could be sequencing errors.
 A file indicating the unique sequences used for the phylogenetic analysis, their mutations relative to proCoV2, and other sequences in that cluster is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/phylogenetics/all_alignment.csv}.
@@
  \begin{figure}[h!]
- {\bf \LARGE A} \\
  \centerline{
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RpYN06_without_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RpYN06_without_deleted_seqs.pdf}
  \hspace{0.04\linewidth}
@@
   \hspace{0.04\linewidth}
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RpYN06_without_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RpYN06_without_deleted_seqs.pdf}
  }
- \\
-  {\bf \LARGE B} \\
+ \caption{
+ A version of Figure~\ref{fig:tree_RaTG13} but rooting using an outgroup of RpYN06.
+ The tree topologies are identical to those obtained using RaTG13 in Figure~\ref{fig:tree_RaTG13}, with the only differences being a few minor changes in which mutations on branches are towards the outgroup (purple mutation labels).
+\label{suppfig:tree_RpYN06}
+ }
+ \end{figure}
+
+  \begin{figure}[h!]
 \centerline{
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RmYN02_without_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RmYN02_without_deleted_seqs.pdf}
  \hspace{0.04\linewidth}
@@
   \hspace{0.04\linewidth}
- \includegraphics[width=0.3\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RmYN02_without_deleted_seqs.pdf}
+ \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RmYN02_without_deleted_seqs.pdf}
  }
  \caption{
- A version of Figure~\ref{fig:tree_RaTG13} but rooting using an outgroup of (A) RpYN06 or (B) RmYN02 outgroups.
+ A version of Figure~\ref{fig:tree_RaTG13} but rooting using an outgroup of RmYN02 outgroups.
  The tree topologies are identical to those obtained using RaTG13 in Figure~\ref{fig:tree_RaTG13}, with the only differences being a few minor changes in which mutations on branches are towards the outgroup (purple mutation labels).
-\label{suppfig:tree_RpYN06_RmYN02}
+\label{suppfig:tree_RmYN02}
  }

### 2021-07-31 paper/reviewer_response.tex
update points in Figure 2 to better label market sequences
filter to remove rare mutations rather than rare variants

@@
 
+\response{This is a good suggestion, and I have made the suggested change.
+The trees now show more sequences exactly as the reviewer suggests, but the rooting and other basic evolutionarily relevant observations remain unchanged.}
+
 4) In addition to Figures 2 and 4 it would be informative to show a haplotype network combining these early SARS-CoV-2 sequences and RaTG13, with different colors according to where the virus sequences were collected as in Figure 5.
@@
 
+\response{This is a good suggestion, and I have made the recommended change.}
+
 page 4, Fig. 2: The distinction between squares and circles is hard to see on the Figure. One suggestion is to put the data points associated with the market in black.
+
+\response{This is a good suggestion, and I have added black outlines to the squares indicating the market-associated sequences.}
 

### 2021-08-04 paper/paper.tex
finish reviewer response

@@
 Then China's State Council issued a much broader order requiring central approval of all publications related to COVID-19 to ensure they were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
-In 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
+In 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, reviving the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
 
@@
 \citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
-Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the outgroup.
-However, the sophisticated algorithm of \citet{kumar2021evolutionary} and my more simplistic approach both ignore sampling dates, which is probably a major reason that they both infer a progenitor that is three mutations away from the earliest sequences reported from Wuhan.
+Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the outgroup---a perhaps unsurprising result, since both approaches  use outgroups but ignore sampling dates.
+It is also possible that there was an early ascertainment bias towards market-associated viruses (clade B), since some early case definitions included exposure to the Huanan Seafood Market.
 
@@
 There is no obvious scientific reason for the deletion: the sequences are concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
-After I e-mailed the NIH the original version of this manuscript, they sent me the e-mails requesting deletion of the data, which are in Figure~\ref{fig:SRA_email}.
-While I cannot rule out that the authors posted the data on some unknown website, the sequences are not in GISAID, NCBI, or any database used by the joint WHO-China report.
-Therefore, even though the sequencing data were on the Google Cloud (as described above) and the mutations were listed in a table in the \textit{Small} paper by \citet{wang2020small}, the practical consequence of removing the data from the SRA was that nobody was aware these sequences existed.
-Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to maximize information about viral sequences from early in the Wuhan epidemic.
+After I e-mailed the NIH the original version of this manuscript, they sent me the e-mail requesting deletion of the data, which is in Figure~\ref{fig:SRA_email}.
+Despite the statement in the deletion-request e-mail that the sequences were being uploaded to ``another website'' (Figure~\ref{fig:SRA_email}), I could find no evidence that they were actually uploaded to any other public website---and certainly they were not posted to GISAID, NCBI, or any database used by the joint WHO-China report.
+Therefore, even though the sequencing data were listed in a table in the \textit{Small} paper by \citet{wang2020small}, the practical consequence of removing the data from the SRA was that nobody was aware these sequences existed.
+Notably, it is \emph{not} possible to delete pre-prints from \textit{bioRxiv} and \textit{medRxiv}, so once \citet{wang2020medRxiv} had posted their pre-print, it was permanently committed to the public record (withdrawn pre-prints are still accessible, for instance see \citet{yang2020epidemiological}).
+Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, the deletion of the data from the SRA suggests a less than wholehearted effort to maximize availability of information about viral sequences from early in the Wuhan epidemic.
 
-Another important implication is that genomic epidemiology studies of early SARS-CoV-2 need to pay as much attention to the provenance and annotation of the underlying sequences as technical considerations.
+However, about a month after I publicly posted this manuscript to \textit{bioRxiv} as a pre-print, the deleted sequence data were discussed in a July 21, 2021, press conference by the vice minister of China's National Health Commission and subsequently in two blog posts by a China state-affiliated reporter~\citep{wang2021exclusive,wang2021why}.
+The press conference and blog posts stated that the sequence data were removed from the SRA because the journal \textit{Small} had mistakenly deleted the data availability statement in the paper of \citet{wang2020small}, leading the authors to think that they should also delete the actual data.
+Confusingly, this explanation is inconsistent with the reason provided by the authors to the SRA (Figure~\ref{fig:SRA_email}), and I am unable to reconcile the discrepancy.
+The press conference and blog posts also stated that the sequences were all collected on or after January 30, 2020, rather than ``early in the epidemic'' as originally described in \citet{wang2020medRxiv,wang2020small}.
+Finally, the press conference and blog noted that on July 8, 2021, the deleted data were made available on the Chinese NGDC database under accession PRJCA005725.
+
+Regardless of exactly why the data were deleted, another important implication of this study is that genomic epidemiology studies of early SARS-CoV-2 need to pay as much attention to the provenance and annotation of the underlying sequences as technical considerations.
 There has been substantial scientific effort expended on topics such as phylogenetic rooting~\citep{pipes2021assessing, morel2021phylogenetic}, novel algorithms~\citep{kumar2021evolutionary}, and correction of sequencing errors~\citep{turakhia2020stability}.
@@
 I thank the citizens and scientists on Twitter who helped inspire this study and inform its background through their analyses and discussions of the early spread of SARS-CoV-2.
+I thank two anonymous reviewers for helpful comments on the manuscript.
 I thank Brendan Larsen for pointing out that I failed to mask the primer binding site regions in the analysis described in the original pre-print (this has been fixed in this version).

### 2021-08-04 paper/reviewer_response.tex
finish reviewer response

@@
 \response{Thanks for the helpful comments.
-I have addressed all of the comments from Reviewer \#1 and all the comments from Reviewer \#2 that I am able to address in the context of the limitations of the study.
+I have addressed the comments from Reviewer \#1, and all the comments from Reviewer \#2 that I can address in the context of the limitations of the study.
 Also, thanks for the point about Kumar et al's proCoV2 inference also essentially being an outgroup method.
-I still mention this prominently to credit their excellent work, but also make clear that there are some shared assumptions of the two methods, so the concordance does not represent a validation of those assumptions but only an indication of reaching similar inferences under some of the same assumptions.}
+I still mention this study prominently to credit its excellent work, but also make clear that the two methods share the outgroup assumption, so the concordance does not represent a validation of those assumptions but only an indication of reaching similar inferences under similar assumptions.
+
+In addition, subsequent to the posting of this pre-print, the Chinese Health Council addressed the \textit{bioRxiv} pre-print of my manuscript in response to a question that was asked during a July news conference in which China rejected the WHO's request for further investigations into the origins of COVID-19.
+At that news conference, the Chinese Health Council offered a new explanation for the deletion of the sequences that was distinct from the explanation originally given by the authors to the SRA.
+This new explanation is that the sequences were deleted in response to an editorial error in which the journal \textit{Small} mistakenly deleted the data availability statement for the paper, apparently making the authors feel compelled to delete the data itself.
+In the revised Discussion, I now note this new explanation alongside the one originally offered to the SRA by the authors.
+I also now note that subsequent to the publication of this pre-print, the deleted sequences were made available on the Chinese NGDC database on July-8-2021.
+Finally, I note that at its news conference, the Chinese Health Council updated the description of the sequences from collected ``early in the epidemic'' (as they were described in the pre-print) to all collected on or after January 30.
+Because I do not have the information necessary to distinguish the veracity of the original versus new explanations / descriptions, I provide both in the Discussion.
+}
 
@@
 Comments to the Author
+
 The author identifies a set of SARS-CoV-2 sequences from the early days of the pandemic which has been deleted from the SRA database.
@@
 
+\response{Thank you for the summary of the manuscript.}
+
 Minor comments
@@
 
+\response{This is a good suggestion that I have incorporated.}
+
 2) In the discussion or the results section it should also be mentioned that there is no mechanism for deleting a pre-print (unlike for SRA data).
 So the authors were committed to having their preprint permanently available after it was posted in March.
+
+\response{I have added mention of this fact to to the revised Discussion}
 
@@
 4) In addition to Figures 2 and 4 it would be informative to show a haplotype network combining these early SARS-CoV-2 sequences and RaTG13, with different colors according to where the virus sequences were collected as in Figure 5.
+
+\response{I apologize, but I have been unable to identify software that enables visualization of haplotype networks that enables both pie-chart coloring of nodes and sizing of nodes based on number of sequences.
+If the reviewer is able to suggest specific software that enables this task, I would be happy to do add such a network.
+}
 
@@
 Comments to the Author
+
 The primary shortcoming of this work is its significance.
@@
 
+\response{I fully agree that that this study does not discover the existence of two lineages, and have made certain that the revised version fully acknowledges that this study simply provides more sequences that fall into these lineages.
+I do note, however, that this study does provide more evidence for T29095 possibly being present in the most ancestral form of lineage A at least based on outgroup rooting.
+
+I also fully agree that the SNPs were reported in \textit{Small}, and have made sure to mention this point several times in the revised version.
+However, it is also undeniably true that the sequences had been completely overlooked in any phylogenetic analyses due to their absence in any sequence database after the deletion.
+
+I also fully acknowledge the point about the uncertain timing of collection.
+I have included a paragraph in the Discussion beginning "There are several caveats to this study...'' that discusses this limitation among others.
+Also, I have provided more detail about what has been stated about the collection dates of the sequences.
+These dates were originally described in the pre-print as ``early in the epidemic.''
+These dates were then updated in the final paper to ``early in the epidemic (January 2020).''
+Then in a recent press conference by the China Health Council, this description was further updated to no earlier than January 30.
+I have provided all three descriptions in the revised manuscript.
+
+Finally, I fully acknowledge that these data do not decisively answer questions about the rooting or timing of the introduction---and certainly not the route of introduction (the manuscript does not even touch on this last point).
+As far as rooting and timing, I do think that the addition of even partial data is informative since what we have to work with is so meager, and I also think that this manuscript has productively sparked more discussion and analysis of the rooting question.
+But certainly I recognize the limitations, and have made revisions to ensure they are adequately conveyed.}
+
 The author casts doubt on the timing and distribution of early cases based on media reports of unknown reliability. It is impossible for the reader to interpret this.
+
+\response{I agree that the poor reliability of data about early cases is unfortunate.
+However, I think it is important to include both media and scientific-literature reports.
+For instance, the joint WHO-China report itself was recently revised in response to an article in the \textit{Washington Post} (\url{https://www.washingtonpost.com/world/asia_pacific/covid-wuhan-outbreak-who/2021/07/15/51e7e8a6-e2c6-11eb-88c5-4fd6382c47cb_story.html}), indicating that in at least some cases media reports can be as accurate as scientific literature.
+
+Therefore, I have simply cited all of the competing reports that seem of reasonable reliability, and taken pains not to present any of them as the absolute truth.}
 
@@
 Additionally, this presupposes early diversity in SARS-CoV-2 is due to the emergence of SNPs during human to human transmission, rather than in a putative intermediate host.
-The author additionally notes that "unusual mutational processes" might explain an ancestral position for lineage B, and indeed C>T, T>C mutations are frequently observed in SARS-CoV-2, complicating the inference of directional relationships between viral lineages dependent on these SNPs.
+The author additionally notes that "unusual mutational processes" might explain an ancestral position for lineage B, and indeed C$\rightarrow$T, T$\rightarrow$C mutations are frequently observed in SARS-CoV-2, complicating the inference of directional relationships between viral lineages dependent on these SNPs.
+
+\response{The reviewer makes an excellent point that some early case definitions of COVID-19 included exposure to the Huanan Seafood Market, and that this could bias case ascertainment and follow-on sequencing.
+In fact, this explanation would make special sense given that the sequences analyzed here are from suspected outpatients---which might be more likely to be the status of early infected individuals who didn't meet a market-associate case definition.
+I have added text that mentions this point as another possible explanation for the preponderance of lineage B (market-associated) viruses among the early sequences.
+(Note that although I agree with this point and have seen various reports of market exposure being part of the case definition, I was unable to find any citations for this point so if the reviewer knows any to suggest, I would be happy to add them upon further revision.)
+
+I mention the mutation-bias explanation as it was mentioned by Pipes et al (2021), but mutation biases are unlikely to fully explain the conundrums related to outgroup versus date-based rooting.
+For instance, site 29095 is C in SARS-CoV-2, but conserved as T in \emph{all} other related sarbecoviruses indicating that it is phylogenetically informative (if its mutation rate was saturatingly high, it would not be conserved as T in all related sarbecoviruses).
+
+Regarding the possible role of an intermediate host, see my response to the next point.
+}
 
@@
 There is simply too much uncertainty currently, and it may be as likely as otherwise that the root falls between lineages A and B, rather than in either one of them.
-Alternatively, the author cannot exclude A>B directionality occurring in an intermediate host rather than in humans, unless one discounts the possibility of an intermediate host.
+Alternatively, the author cannot exclude A$\rightarrow$B directionality occurring in an intermediate host rather than in humans, unless one discounts the possibility of an intermediate host.
 The author acknowledges the problematic nature of the rooting issue, but nevertheless heavily weights the conclusions on this analysis.
+
+\response{I fully agree with the reviewer that the currently available outgroups (RaTG13, RmYN02, etc) are more diverged than ideal for rooting, and the confidence in rooting would be greatly improved if the outgroup was more similar.
+Unfortunately, no higher-identity outgroup is known.
+I have been sure to mention the high divergence of the outgroup as a possible limitation in rooting.
+Nonetheless, outgroup rooting with any of the available outgroups consistently places the root in lineage A (specifically, one of the three lineage A variants that is 3 mutations closer to the outgroup than lineage B), so this cannot be entirely dismissed.
+In any case, it's important to discuss these limitations as many other authors (for instance, Rambaut et al (2020)) placed the root in lineage A with rather high confidence in their published work.
+
+I agree using proCoV2 to root is circular, and I meant to only indicate that my outgroup rooting was similar to that of Kumar et al when they proposed proCoV2.
+I have clarified this point, as also suggested by the editor.
+
+As far as an intermediate host, that doesn't really change the conundrum.
+The rooting isn't dependent on which host the virus is replicating in, but only on the assumption that the deeper ancestors are bat coronaviruses.
+Even if the virus went through an intermediate host, or even multiple intermediate hosts, lineage A is still ancestral to lineage B on any tree that uses outgroup rooting.
+}
 
@@
 The phylogenetic analysis would be greatly strengthened by a more thorough discussion of this matter.
-If A>B directionality in humans is presumed, it depends on a single introduction event, which the author should make the case for.
+If A$\rightarrow$B directionality in humans is presumed, it depends on a single introduction event, which the author should make the case for.
+
+\response{
+The reviewer is correct that I am assuming a single introduction, and I have made this assumption explicit in the revised manuscript.
+I would note that a single introduction has been implicitly or explicitly assumed in all the other major published work on this topic, including Rambaut et al (2020), Pipes et al (2021), and Kumar et al (2021).
+
+It is accurate to say that a single introduction is the most parsimonious explanation.
+The definition of parsimony is to assume no more events than necessary to explain the data, and because there is not explicit documentary evidence for even a single introduction (although clearly an introduction did happen), it is certainly more parsimonious to assume just one introduction rather than two.
+In addition, all of the currently known early variants of SARS-CoV-2 are within 2 mutations of each other, which is well within the span of genetic divergence routinely observed in single introduction outbreaks of SARS-CoV-2.
+}
 
@@
 
+\response{
+I recognize different readers may have different views of the relevance of this aspect.
+But I think in the end the manuscript is both about the sequences themselves and the fact that they were not readily available, which is a fact that could reasonably inform views of data sharing practices.
+
+In the end, much of the feedback I have received elsewhere is very curious about this aspect.
+For instance, I received comments from Stephen Goldstein (\url{http://disq.us/p/2hwabcu}) where he explicitly asked (comment 8) for inclusion of the reason for the removal.
+In addition, as mentioned in the response to the editor, this point has been further raised by the Chinese Health Council.
+
+Therefore, I think it would be a glaring omission to at least some readers to not provide the explanations given for the data deletion.
+}
+
 The acknowledgement of "twitter citizens" risks highlighting commentary by individuals who have directed abuse and threats at the author's colleagues.
 I encourage the author to reconsider, or specify who he refers to in particular in order to avoid granting inadvertent credibility to nefarious individuals.
+
+\response{I certainly recognize that there are some individuals who post inaccurate or negative information on Twitter, but there are also many individuals who post useful information.
+This includes three individuals (one a professor, one a postdoc, and one a non-scientist) who sent me information about the two accessions I could not recover from the cloud after I Tweeted about the original pre-print.
+However, all three preferred not to be acknowledged by name.
+
+My understanding of many aspects of SARS-CoV-2 (including its early evolution, later evolution, and antigenicity) have been greatly informed by postings on Twitter, including from scientists, journalists, policy makers, and others whose professions are unknown to me.
+Therefore, I think it is appropriate to acknowledge this fact---recognizing of course that Twitter has millions of users, who range in their intentions and credibility.
+}
 

### 2021-08-04 paper/paper.tex
few more minor updates to paper

@@
 This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
-Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
+Using this qualitative principle, the middle and leftmost scenarios in Figure~\ref{fig:tree_with_deleted} seem more plausible than the rightmost scenario, where the progenitor lacks branching descendants.
 I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the sampling distribution of the underlying data.
@@
 Analysis of these sequences leads to several conclusions.
-First, they provide further evidence Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
+First, they provide further evidence that the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
 The deleted data as well as existing sequences from Wuhan-infected patients hospitalized in Guangdong show early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than sequences in the joint WHO-China report~\citep{WHO2021origins}.
@@
 Therefore, it is impossible to unambiguously place them phylogenetically, or determine exactly when they were collected.
-However, little can be done to mitigate this caveat beyond my unsuccessful attempt to contact the corresponding authors of \citet{wang2020medRxiv}.
 It is also important to note that my phylogenetic analyses use relatively simple methods to draw qualitative conclusions without formal statistical testing.

### 2021-08-04 paper/paper.tex
split supplement into separate document

@@
 I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample, and masking regions that corresponded to the primer binding sites described in \citet{wang2020medRxiv}.
-Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
+Figure~S1 shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~S2.
 
@@
 Go to \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/deltadist.html} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
-Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
+Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~S3.
 }
@@
 All sequences associated with this market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13}).
-Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
+Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~S3).
 
@@
 Singleton mutations (mutations observed only once in the sequence set) are removed as described in more detail in the Methods.
- Figures~\ref{suppfig:tree_RpYN06} and \ref{suppfig:tree_RmYN02} shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
+ Figures~S4 and S5 shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
 \label{fig:tree_RaTG13}
@@
 This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the progenitor sequence is closest to the bat coronavirus outgroup.
-If we limit the analysis to sequences collected no later than January 2020 and remove singleton mutations observed only once in the sequence set, then there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06}, Figure~\ref{suppfig:tree_RmYN02}).
+If we limit the analysis to sequences collected no later than January 2020 and remove singleton mutations observed only once in the sequence set, then there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~S4, Figure~S5).
 Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
@@
 The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
- Figures~\ref{suppfig:tree_RpYN06} and \ref{suppfig:tree_RmYN02} demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
+ Figures~S4 and S5 demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
 \label{fig:tree_with_deleted}
@@
 I also masked (set to zero coverage) all sites overlapped by the primers used by \citet{wang2020medRxiv} to PCR the amplicons used for sequencing.
-These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
+These coverage plots are in Figure~S1 and Figure~S2; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
 I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
@@
 
-\clearpage
-\onecolumn
-\renewcommand{\thepage}{S\arabic{page}}
-\setcounter{page}{1}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\section{Supplementary Material}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\linewidth]{figures/coverage_region.pdf}
-\caption{Sequencing depth over the SARS-CoV-2 genome from site 21,570 to 29,550 for the 34 virus-positive early epidemic samples and the 16 samples from hospitalized patients in February.
-Depth is the number of aligned reads that cover that site with a quality score $\ge$20.
-The dashed gray line is the minimum coverage required to call a consensus identity at a site.
-Note that the y-axis uses a symlog scale.
-An interactive version of this plot that enables zooming into specific site ranges and mouseovers to see read count statistics at each site is at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/coverage_region.html}.
-A version of the plot where the x-axis spans the entire SARS-CoV-2 genome is in Figure~\ref{suppfig:coverage_all}.
-}
-\label{suppfig:coverage}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\linewidth]{figures/coverage_all.pdf}
-\caption{A version of Figure~\ref{suppfig:coverage} that shows coverage over the full length of the SARS-CoV-2 genome.
-An interactive version of this plot is at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/coverage_all.html}.
-}
-\label{suppfig:coverage_all}
-\end{figure}
-
-\begin{figure}[h!]
-{\bf \LARGE A} \\
-\centerline{\includegraphics[width=\linewidth]{figures/deltadist_RpYN06.pdf}}
-{\bf \LARGE B} \\
-\centerline{\includegraphics[width=\linewidth]{figures/deltadist_RmYN02.pdf}}
-\caption{Versions of Figure~\ref{fig:deltadist_RaTG13} but calculating the relative mutational distances using an outgroup of (A) RpYN06 or (B) RmYN02.
-}
-\label{suppfig:deltadist_RpYN06_RmYN02}
-\end{figure}
-
- \begin{figure}[h!]
- \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RpYN06_without_deleted_seqs.pdf}
- \hspace{0.04\linewidth}
- \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RpYN06_without_deleted_seqs.pdf}
-  \hspace{0.04\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RpYN06_without_deleted_seqs.pdf}
- }
- \caption{
- A version of Figure~\ref{fig:tree_RaTG13} but rooting using an outgroup of RpYN06.
- The tree topologies are identical to those obtained using RaTG13 in Figure~\ref{fig:tree_RaTG13}, with the only differences being a few minor changes in which mutations on branches are towards the outgroup (purple mutation labels).
-\label{suppfig:tree_RpYN06}
- }
- \end{figure}
-
-  \begin{figure}[h!]
-\centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RmYN02_without_deleted_seqs.pdf}
- \hspace{0.04\linewidth}
- \includegraphics[width=0.32\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RmYN02_without_deleted_seqs.pdf}
-  \hspace{0.04\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RmYN02_without_deleted_seqs.pdf}
- }
- \caption{
- A version of Figure~\ref{fig:tree_RaTG13} but rooting using an outgroup of RmYN02 outgroups.
- The tree topologies are identical to those obtained using RaTG13 in Figure~\ref{fig:tree_RaTG13}, with the only differences being a few minor changes in which mutations on branches are towards the outgroup (purple mutation labels).
-\label{suppfig:tree_RmYN02}
- }
- \end{figure}
-
 \end{document}

### 2021-08-04 paper/paper.tex
made submission-ready figures

@@
 \centering
-\fbox{\includegraphics[trim={0.2in 0 0 9in},clip,width=\linewidth]{figures/acc_removed.png}}
+%\fbox{\includegraphics[trim={0.2in 0 0 9in},clip,width=\linewidth]{figures/acc_removed.png}}
+\fbox{\includegraphics[trim={0.2in 0 0 9in},clip,width=\linewidth]{figures/Figure_1.jpg}}
 \caption{Accessions from deep sequencing project PRJNA612766 have been removed from the SRA.
@@
 \centering
-\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
+%\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
+\includegraphics[width=\linewidth]{figures/Figure_2.pdf}
 \caption{The reported collection dates of SARS-CoV-2 sequences in GISAID versus their relative mutational distances from the RaTG13 bat coronavirus outgroup.
@@
  \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_without_deleted_seqs.pdf}
- \hspace{0.02\linewidth}
- \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
-  \hspace{0.02\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_without_deleted_seqs.pdf}
+ \includegraphics[width=\linewidth]{figures/Figure_3.pdf}
+ %\includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_without_deleted_seqs.pdf}
+ %\hspace{0.02\linewidth}
+ %\includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
+  %\hspace{0.02\linewidth}
+ %\includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_without_deleted_seqs.pdf}
  }
@@
 \centering
-\includegraphics[width=0.75\linewidth]{figures/deltadist_jitter.png}
+%\includegraphics[width=0.75\linewidth]{figures/deltadist_jitter.png}
+\includegraphics[width=0.75\linewidth]{figures/Figure_4.jpg}
 \caption{
@@
  \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_with_deleted_seqs.pdf}
- \hspace{0.02\linewidth}
- \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
-  \hspace{0.02\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_with_deleted_seqs.pdf}
+ \includegraphics[width=\linewidth]{figures/Figure_5.pdf}
+% \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_with_deleted_seqs.pdf}
+% \hspace{0.02\linewidth}
+% \includegraphics[width=0.32\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
+ % \hspace{0.02\linewidth}
+% \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_with_deleted_seqs.pdf}
  }
@@
 \begin{figure}
-\centerline{\fbox{\includegraphics[width=\linewidth]{figures/SRA_email.png}}}
+%\centerline{\fbox{\includegraphics[width=\linewidth]{figures/SRA_email.png}}}
+\centerline{\fbox{\includegraphics[width=\linewidth]{figures/Figure_6.jpg}}}
 \caption{A redacted version of the e-mails from Wuhan University to the SRA staff requesting deletion of the sequencing data.

### 2021-08-19 paper/supp.tex

--- 2021-06-28 paper/paper.tex
@@
 
-%\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
-%\hypersetup{draft}  % comment out hyperlinks for now to deal with line-breaking issue in references
-
-\articletype{inv}
-
-\title{Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic}
-
-\author[]{\Large Jesse D. Bloom}
-
-\affil[]{Fred Hutchinson Cancer Research Center}
-\affil[]{Howard Hughes Medical Institute}
-\affil[]{Seattle, WA, USA}
-
-\keywords{}
-
 \runningtitle{} % For use in the footer 
@@
 
-\begin{abstract}
-The origin and early spread of SARS-CoV-2 remains shrouded in mystery.
-Here I identify a data set containing SARS-CoV-2 sequences from early in the Wuhan epidemic that has been deleted from the NIH's Sequence Read Archive.
-I recover the deleted files from the Google Cloud, and reconstruct partial sequences of 13 early epidemic viruses.
-Phylogenetic analysis of these sequences in the context of carefully annotated existing data supports existing data indicating that the Huanan Seafood Market sequences that are the focus of the joint WHO-China report are not fully representative of the viruses in Wuhan early in the epidemic.
-Instead, the progenitor of known SARS-CoV-2 sequences likely contained three mutations relative to the market viruses that made it more similar to SARS-CoV-2's bat coronavirus relatives. 
-\end{abstract}
+\usepackage{xr}
+\externaldocument[]{paper}
 
@@
 
-\maketitle
-\thispagestyle{firststyle}
-%\marginmark
-\firstpagefootnote
-
-\correspondingauthoraffiliation{}{Corresponding author: \href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org}}
-\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
-
-\lettrine[lines=2]{\color{color2}U}{}nderstanding the spread of SARS-CoV-2 in Wuhan is crucial to tracing the origins of the virus, including identifying events that led to infection of patient zero.
-The first reports outside of China at the end of December 2019 emphasized the role of the Huanan Seafood Market~\citep{ProMED2109}, which was initially suggested as a site of zoonosis.
-However, this theory became increasingly tenuous as it was learned that many early cases had no connection to the market~\citep{cohen2020wuhan,huang2020clinical, chen2020epidemiological}.
-Eventually, Chinese CDC Director Gao Fu dismissed the theory, stating ``At first, we assumed the seafood market might have the virus, but now the market is more like a victim.
-The novel coronavirus had existed long before''~\citep{GlobalTimes2020}.
-
-Indeed, there were reports of cases that far preceded the outbreak at the Huanan Seafood Market.
-The \textit{Lancet} described a confirmed case having no association with the market whose symptoms began on December 1, 2019~\citep{huang2020clinical}.
-The \textit{South China Morning Post} described nine cases from November 2019 including details on patient age and sex, noting that none were confirmed to be ``patient zero''~\citep{ma2020SCMP}.
-Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} that records he reviewed showed two cases in mid-November, and one suspected case on September 29~\citep{healthtimes2020}.
-At about the same time as Professor Chuanhua's interview, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
-Then China's State Council issued a much broader order requiring central approval of all publications related to COVID-19 to ensure they were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
-In 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.
-
-In other outbreaks where direct identification of early cases has been stymied, it has increasingly become possible to use genomic epidemiology to infer the timing and dynamics of spread from analysis of viral sequences.
-For instance, analysis of SARS-CoV-2 sequences has enabled reconstruction of the initial spread of SARS-CoV-2 in North America and Europe~\citep{bedford2020cryptic, worobey2020emergence, deng2020genomic, fauver2020coast}.
-
-But in the case of Wuhan, genomic epidemiology has also proven frustratingly inconclusive.
-Some of the problem is simply limited data: despite the fact that Wuhan has advanced virology labs, there is only patchy sampling of SARS-CoV-2 sequences from the first months of the city's explosive outbreak.
-Other than a set of multiply sequenced samples collected in late December of 2019 from a dozen patients connected to the Huanan Seafood Market~\citep{WHO2021origins}, just a handful of Wuhan sequences are available from before late January of 2020 (see analysis in this study below).
-This paucity of sequences could be due in part to an order that unauthorized Chinese labs destroy all coronavirus samples from early in the outbreak, reportedly for ``laboratory biological safety'' reasons~\citep{pinghui2020SCMP}.
-
-However, the Wuhan sequences that are available have also confounded phylogenetic analyses designed to infer the ``progenitor'' of SARS-CoV-2, which is the sequence from which all other currently known sequences are descended~\citep{kumar2021evolutionary}.
-Although there is debate about exactly how SARS-CoV-2 entered the human population, it is universally accepted that the virus's deep ancestors are bat coronaviruses~\citep{lytras2021exploring}.
-But the earliest known SARS-CoV-2 sequences, which are mostly derived from the Huanan Seafood Market, are notably more different from these bat coronaviruses than other sequences collected at later dates outside Wuhan.
-As a result, there is a direct conflict between the two major principles used to infer an outbreak's progenitor: namely that it should be among the earliest sequences, and that it should be most closely related to deeper ancestors~\citep{pipes2021assessing}.
-
-Here I take a step towards resolving these questions by identifying and recovering a deleted data set of partial SARS-CoV-2 sequences from outpatient samples collected early in the Wuhan epidemic.
-Analysis of these new sequences in conjunction with careful annotation of existing ones suggests that the early Wuhan samples that have been the focus of most studies including the joint WHO-China report~\citep{WHO2021origins} are not fully representative of the viruses actually present in Wuhan at that time.
-These insights help reconcile phylogenetic discrepancies, and suggest two plausible progenitor sequences, one of which is identical to that inferred by \citet{kumar2021evolutionary}.
-Furthermore, the approach taken here hints it may be possible to advance understanding of SARS-CoV-2's origins or early spread even without further on-the-ground studies, such as by more deeply probing data archived by the NIH and other entities.
-
-\section{Results}
-
-\subsection{Identification of a SARS-CoV-2 deep sequencing data set that has been removed from the Sequence Read Archive}
-During the course of my research, I read a paper by \citet{farkas2020insights} that analyzed SARS-CoV-2 deep sequencing data from the Sequence Read Archive (SRA), which is a repository maintained by the NIH's National Center for Biotechnology Information.
-The first supplementary table of \citet{farkas2020insights} lists all SARS-CoV-2 deep sequencing data available from the SRA as of March 30, 2020.
-
-The majority of entries in this table refer to a project (BioProject PRJNA612766) by Wuhan University that is described as nanopore sequencing of SARS-CoV-2 amplicons.
-The table indicates this project represents 241 of the 282 SARS-CoV-2 sequencing run accessions in the SRA as of March, 30, 2020.
-Because I had never encountered any other mention of this project, I performed a Google search for ``PRJNA612766,'' and found no search hits other than the supplementary table itself.
-Searching for ``PRJNA612766'' in the NCBI's SRA search box returned a message of ``No items found.''
-I then searched for individual sequencing run accessions from the project in the NCBI's SRA search box.
-These searches returned messages indicating that the sequencing runs had been removed (Figure~\ref{fig:acc_removed}).
-
-\begin{figure}[]
-\centering
-\fbox{\includegraphics[trim={0.2in 0 0 9in},clip,width=\linewidth]{figures/acc_removed.png}}
-\caption{Accessions from deep sequencing project PRJNA612766 have been removed from the SRA.
-Shown is the result of searching for ``SRR11313485'' in the SRA search toolbar.
-This result has been digitally archived on the Wayback Machine at \url{https://web.archive.org/web/20210502131630/https://trace.ncbi.nlm.nih.gov/Traces/sra/?run=SRR11313485}.
-}%
-\label{fig:acc_removed}
-\end{figure}
-
-The SRA is designed as a permanent archive of deep sequencing data.
-The SRA documentation states that after a sequencing run is uploaded, ``neither its files can be replaced nor filenames can be changed,'' and that data can only be deleted by e-mailing SRA staff~\citep{SRA_deletion}.
-
-\subsection{The deleted data set contains sequencing of viral samples collected early in the Wuhan epidemic}
-The metadata in the first supplementary table of \citet{farkas2020insights} indicates that the samples in deleted project PRNJA612766 were collected by Aisu Fu and Renmin Hospital of Wuhan University.
-Google searching for these terms revealed the samples were related to a study posted as a pre-print on \textit{medRxiv} in early March of 2020~\citep{wang2020medRxiv}, and subsequently published in the journal \textit{Small} in June of 2020~\citep{wang2020small}.
-
-The study describes an approach to diagnose infection with SARS-CoV-2 and other respiratory viruses by nanopore sequencing.
-This approach involved reverse-transcription of total RNA from swab samples, followed by PCR with specific primers to generate amplicons covering portions of the viral genome.
-These amplicons were then sequenced on an Oxford Nanopore GridION, and infection was diagnosed if the sequencing yielded sufficient reads aligning to the viral genome.
-Importantly, the study notes that this approach yields information about the sequence of the virus as well enabling diagnosis of infection.
-
-The pre-print~\citep{wang2020medRxiv} says the approach was applied to ``45 nasopharyngeal swab samples from outpatients with suspected COVID-19 early in the epidemic.''
-The digital object identifier (DOI) for the pre-print indicates that it was processed by \textit{medRxiv} on March 4, 2020, which is one day after China's State Council ordered that all papers related to COVID-19 must be centrally approved~\citep{Kang2020}.
-The final published manuscript~\citep{wang2020small} from June of 2020 updated the description from ``early in the epidemic'' to ``early in the epidemic (January 2020).''
-Both the pre-print and published manuscript say that 34 of the 45 early epidemic samples were positive in the sequencing-based diagnostic approach.
-In addition, both state that the approach was later applied to 16 additional samples collected on February 11--12, 2020, from SARS-CoV-2 patients hospitalized at Renmin Hospital of Wuhan University.
-
-There is complete concordance between the accessions for project PRJNA612766 in the supplementary table of \citet{farkas2020insights} and the samples described by \citet{wang2020medRxiv}.
-There are 89 accessions corresponding to the 45 early epidemic samples, with these samples named like wells in a 96-well plate (A1, A2, etc).
-The number of accessions is approximately twice the number of early epidemic samples because each sample has data for two sequencing runtimes except one sample (B5) with just one runtime.
- There are 31 accessions corresponding to the 16 samples collected in February from Renmin Hospital patients, with these samples named R01, R02, etc.
- Again, all but one sample (R04) have data for two sequencing runtimes.
- In addition, there are 7 accessions corresponding to positive and negative controls, 2 accessions corresponding to other respiratory virus samples, and 112 samples corresponding to plasmids used for benchmarking of the approach.
- Together, these samples and controls account for all 241 accessions listed for PRJNA612766 in the supplementary table of \citet{farkas2020insights}.
-
-Neither the pre-print~\citep{wang2020medRxiv} nor published manuscript~\citep{wang2020small} contain any correction or note that indicates a scientific reason for deleting the study's sequencing data from the SRA.
-I e-mailed both corresponding authors of \citet{wang2020medRxiv} to ask why they had deleted the deep sequencing data and to request details on the collection dates of the early outpatient samples, but received no reply.
-
-\subsection{Recovery of deleted sequencing data from the Google Cloud} 
-As indicated in Figure~\ref{fig:acc_removed}, none of the deleted sequencing runs could be accessed through the SRA's web interface.
-In addition, none of the runs could be accessed using the command-line tools of the SRA Toolkit.
-For instance, running \texttt{fastq-dump SRR11313485} or \texttt{vdb-dump SRR11313485} returned the message ``err: query unauthorized while resolving query within virtual file system module - failed to resolve accession 'SRR11313485'``.
-
-However, the SRA has begun storing all data on the Google and Amazon clouds.
-While inspecting the SRA's web interface for other sequencing accessions, I noticed that SRA files are often available from links to the cloud such as \texttt{https://storage.googleapis.com/nih-sequence-read-archive/run/<ACCESSION>/<ACCESSION>}.
-
-\begin{table*}[b!]
-{\small
-\begin{tabular}{lrll}
-\toprule
-sample &  fraction sites called (21570-29550) &           patient group &                                        substitutions relative to proCoV2 \\
-\midrule
-A4 &                               0.9266 &        early outpatient &                                                                          none \\
-C1 &                               0.9396 &        early outpatient &  G22081A (A=924, C=4, G=9), C28144T (C=6, T=1185), T29483G (C=1, G=45, T=1) \\
-C2 &                               0.9397 &        early outpatient &                                                 C29095T (C=1, G=1, T=751)  \\
-C9 &                               0.9005 &        early outpatient &                                  C28144T (C=3, T=823), G28514T (G=1, T=36)  \\
-D9 &                               0.9051 &        early outpatient &                                                     C28144T (C=4, T=1653)  \\
-D12 &                               0.9400 &        early outpatient &                                                     C28144T (C=8, T=2400)  \\
-E1 &                               0.9223 &        early outpatient &                                                           C28144T (T=125)  \\
-E5 &                               0.9227 &        early outpatient &              C24034T (A=5, C=3, T=74), T26729C (C=12), G28077C (C=142, G=4)  \\
-E11 &                               0.9321 &        early outpatient &                                 C25460T (C=2, T=246), C28144T (C=1, T=412)  \\
-F11 &                               0.9054 &        early outpatient &                             T25304A (A=9, T=1), C28144T (C=6, G=1, T=1328)  \\
-G1 &                               0.9396 &        early outpatient &                                   none                                        \\
-G11 &                               0.9112 &        early outpatient &                                  none                                         \\
-H9 &                               0.9381 &        early outpatient &                                                     C28144T (C=2, T=1254)  \\
-R11 &                               0.9422 &  hospital patient (Feb) &                               C21707T (T=401), C28144T (A=1, C=18, T=4265)  \\
-\bottomrule
-\end{tabular}
-}
-\caption{Samples for which the SARS-CoV-2 sequence could be called at $\ge$90\% of sites between 21,570 and 29,550, and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
-Numbers in parentheses after each substitution give the deep sequencing reads with each nucleotide identity.
-\label{tab:mutations}
-}
-\end{table*}
-
-Based on the hypothesis that deletion of sequencing runs by the SRA might not remove files stored on the cloud, I interpolated the cloud URLs for the deleted accessions and tested if they still yielded the SRA files.
-This strategy was successful; for instance, as of June 3, 2021, going to \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485} downloads the SRA file for accession SRR11313485.
-I have archived this file on the Wayback Machine at \url{https://web.archive.org/web/20210502130820/https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
-
-I automated this strategy to download the SRA files for 97 of the 99 sequencing runs corresponding to the 34 SARS-CoV-2 positive early epidemic samples and the 16 hospital samples from February.
-The SRA files for two runs (SRR11313490 and SRR11313499) were not accessible via the Google Cloud, but after I posted the first version of this manuscript as a pre-print, several individuals found archived data for these runs that had been downloaded when they were still available on the SRA, and I have used those data in the updated analysis described here (see Methods for details).
-I used the SRA Toolkit to get the object timestamp (\texttt{vdb-dump -{}-obj\_timestamp}) and time (\texttt{vdb-dump -{}-info}) for all SRA files.
-For all files, the object timestamp is February 15, 2020, and the time is March 16, 2020.
-Although the SRA Toolkit does not clearly document these two properties, my \emph{guess} is that the object timestamp may refer to when the SRA file was created from a FASTQ file uploaded to the SRA, and the time may refer to when the accession was made public.
-
-\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
-\citet{wang2020medRxiv} sequenced PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
-They also sequenced a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
-In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
-I slightly trimmed the region of interest to 21,570 to 29,550 because many samples had poor coverage at the termini.
-
-I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample, and masking regions that corresponded to the primer binding sites described in \citet{wang2020medRxiv}.
-Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.
-
-I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
-With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$90\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
-Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely used Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
-Although requiring coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations have coverage $\ge$10.
-In addition, the mutations I called from the raw sequence data in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.
-
-I also determined the consensus sequence of the plasmid control used by \citet{wang2020medRxiv} from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
-Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), plasmid contamination did not afflict the viral samples in the deleted sequencing project.
-
-\begin{figure*}
-\centering
-\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
-\caption{The reported collection dates of SARS-CoV-2 sequences in GISAID versus their relative mutational distances from the RaTG13 bat coronavirus outgroup.
-Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
-The plot shows sequences in GISAID collected no later than February 28, 2020.
-Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
-Points are slightly jittered on the y-axis.
-Go to \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/deltadist.html} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
-Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
-}
-\label{fig:deltadist_RaTG13}
-\end{figure*}
-
-\subsection{Analysis of existing SARS-CoV-2 sequences emphasizes the perplexing discordance between collection date and distance to bat coronavirus relatives}
-To contextualize the viral sequences recovered from the deleted project, I first analyze early SARS-CoV-2 sequences already available in the GISAID database~\citep{shu2017gisaid}.
-The analyses described in this section are not entirely novel, but synthesize observations from multiple prior studies~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing} to provide key background.
-
-Known human SARS-CoV-2 sequences are consistent with expansion from a single progenitor sequence~\citep{kumar2021evolutionary,pekar2021timing,rambaut2020dynamic,forster2020phylogenetic,pipes2021assessing}.
-However, attempts to infer this progenitor have been confounded by a perplexing fact: the earliest reported sequences from Wuhan are \emph{not} the sequences most similar to SARS-CoV-2's bat coronavirus relatives~\citep{pipes2021assessing}.
-This fact is perplexing because although the proximal origin of SARS-CoV-2 remains unclear (i.e., zoonosis versus lab accident), all reasonable explanations agree that at a deeper level the SARS-CoV-2 genome is derived from bat coronaviruses~\citep{lytras2021exploring}.
-One would therefore expect the first reported SARS-CoV-2 sequences to be the most similar to these bat coronavirus relatives---but this is not the case.
-
-This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
-The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
-The discrepancy is especially pronounced for sequences from patients who had visited the Huanan Seafood Market~\citep{WHO2021origins}.
-All sequences associated with this market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
-Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).
-
- \begin{figure*}
- \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_without_deleted_seqs.pdf}
- \hspace{0.02\linewidth}
- \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
-  \hspace{0.02\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_without_deleted_seqs.pdf}
- }
- \caption{
- Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected before Februrary, 2020.
- The trees are identical except they are rooted to make the progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
-Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected.
-The mutations on each branch are labeled, with mutations towards the nucleotide identity in the outgroup in purple.
-The labels at the top of each tree give the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and Wuhan-Hu-1.
-The monophyletic group containing C28144T is collapsed into a node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
-\label{fig:tree_RaTG13}
- }
- \end{figure*}
-
-This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the progenitor sequence is closest to the bat coronavirus outgroup.
-If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
-Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
-Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with other evidence suggesting the progenitor is an early virus from Wuhan~\citep{pipes2021assessing}.
-
-Several plausible explanations have been proposed for the discordance of phylogenetic rooting with evidence that Wuhan was the origin of the pandemic.
-\citet{rambaut2020dynamic} suggest that viruses from the clade labeled ``B'' in Figure~\ref{fig:tree_RaTG13} may just ``happen'' to have been sequenced first, but that other SARS-CoV-2 sequences are really more ancestral as implied by phylogenetic rooting.
-\citet{pipes2021assessing} discuss the conundrum in detail, and suggest that phylogenetic rooting could be incorrect due to technical reasons such as high divergence of the outgroup or unusual mutational processes not captured in substitution models.
-\citet{kumar2021evolutionary} agree that phylogenetic rooting is problematic, and circumvent this problem by using an alternative algorithm to infer a progenitor for SARS-CoV-2 that they name proCoV2.
-Notably, proCoV2 turns out to be identical to one of the putative progenitors yielded by my approach in Figure~\ref{fig:tree_RaTG13} of simply placing the root at the nodes closest to the outgroup.
-However, neither the sophisticated algorithm of \citet{kumar2021evolutionary} nor my more simplistic approach explain \emph{why} the progenitor should be so different from the earliest sequences reported from Wuhan.
-
-Another explanation that I consider less plausible is offered by \citet{garry2021early}: that there were multiple zoonoses from distinct markets, with the Huanan Seafood Market being the source of viruses in clade B, and some other market being the source of viruses that lack T8782C and C28144T (Figure~\ref{fig:tree_RaTG13}).
-However, this explanation requires positing zoonoses in two markets by two progenitors differing by just two mutations, which seems non-parsimonious in the absence of direct evidence for zoonosis in any market.
-
-\subsection{Sequences recovered from the deleted project and better annotation of Wuhan-derived viruses help reconcile inferences about SARS-CoV-2's progenitor}
-To examine if the sequences recovered from the deleted data set help resolve the conundrum described in the previous section, I repeated the analyses including those sequences.
-In the process, I noted another salient fact: four GISAID sequences collected in Guangdong that fall in a putative progenitor node are from two different clusters of patients who traveled to Wuhan in late December of 2019 and developed symptoms before or on the day that they returned to Guangdong, where their viruses were ultimately sequenced~\citep{chan2020familial, kang2020evidence}.
-Since these patients were clearly infected in Wuhan even though they were sequenced in Guangdong, I annotated them separately from both the other Wuhan and other China sequences.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\linewidth]{figures/deltadist_jitter.png}
-\caption{
-Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (21,570--29,550).
-The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.
-Mutational distance is calculated relative to proCoV2, and points are jittered on the y-axis.
-Go to \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/deltadist_jitter.html} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of jittering.
-\label{fig:deltadist_jitter}
-}
-\end{figure}
-
- \begin{figure*}
- \centerline{
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_with_deleted_seqs.pdf}
- \hspace{0.02\linewidth}
- \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
-  \hspace{0.02\linewidth}
- \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_with_deleted_seqs.pdf}
- }
- \caption{
-Phylogenetic trees like those in Figure~\ref{fig:tree_RaTG13} with the addition of the early Wuhan epidemic sequences from the deleted data set, and Guangdong patients infected in Wuhan prior to January 5 annotated separately.
-Because the deleted sequences are partial, they cannot all be placed unambiguously on the tree.
-Therefore, they are added to each compatible node proportional to the number of sequences already in that node.
-The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
- Figure~\ref{suppfig:tree_RpYN06_RmYN02} demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
-\label{fig:tree_with_deleted}
- }
- \end{figure*}
-
-Repeating the analysis of the previous section with these changes shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
-This fact suggests that the market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, are not representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.
-
-Furthermore, it is immediately apparent that the discrepancy between outgroup rooting and the evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong.
-The rooting of the middle tree in Figure~\ref{fig:tree_with_deleted} is now highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
-The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
-This putative progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
-The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
-However, analysis of this rooting is limited by the fact that the defining C18060T mutation is in a region not covered in the deleted sequences.
-The rightmost tree in Figure~\ref{fig:tree_with_deleted} looks less plausible, as it has almost no weight from Wuhan and the first sequence identical to its progenitor was not collected until January 24.
-
-We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple branching lineages.
-This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
-Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
-I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the nature of the underlying data.
-
-\section{Discussion}
-I have identified and recovered a deleted set of partial SARS-CoV-2 sequences from the early Wuhan epidemic.
-Analysis of these sequences leads to several conclusions.
-First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
-The deleted data as well as existing sequences from Wuhan-infected patients hospitalized in Guangdong show early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than sequences in the joint WHO-China report~\citep{WHO2021origins}.
-Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2.
-One is proCoV2 described by \citet{kumar2021evolutionary}, and the other is a sequence that carries three mutations (C8782T, T28144C, and C29095T) relative to Wuhan-Hu-1.
-Crucially, both putative progenitors are three mutations closer to SARS-CoV-2's bat coronavirus relatives than sequences from the Huanan Seafood Market.
-Note also that the progenitor of all known SARS-CoV-2 sequences could still be downstream of the sequence that infected patient zero depending on the transmission dynamics of the first infections.
-
-\begin{figure}
-\centerline{\fbox{\includegraphics[width=0.95\linewidth]{figures/SRA_email.png}}}
-\caption{A redacted version of the e-mails from Wuhan University to the SRA staff requesting deletion of the sequencing data.
-This e-mail was provided to me by the NIH's NCBI Director Stephen Sherry on June, 19, 2021, the day after I e-mailed the NIH an advance copy of this manuscript.
-The redactions and highlighting were done by the NIH, and I am showing the e-mail exactly as it was provided to me.
-}
-\label{fig:SRA_email}
-\end{figure}
-
-The fact that such an informative data set was deleted has implications beyond those gleaned directly from the recovered sequences.
-Samples from early outpatients in Wuhan are a gold mine for anyone seeking to understand spread of the virus.
-Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than delete the partial sequence data.
-There is no scientific reason for the deletion: the sequences are concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states human subjects approval was obtained, and the sequencing shows no evidence of plasmid or sample-to-sample contamination.
-After I e-mailed the NIH the original version of this manuscript, they sent me the e-mails requesting deletion of the data, which are in Figure~\ref{fig:SRA_email}.
-While I cannot rule out that the authors posted the data on some unknown website, they are not in GISAID, NCBI, or any database used by the joint WHO-China report.
-Therefore, even though the data were on the Google Cloud (as described above) and the mutations were listed in a table in the \textit{Small} paper by \citet{wang2020small}, the practical consequence of removing the data from the SRA was that nobody was aware of its existence.
-Particularly in light of the directive that labs destroy early samples~\citep{pinghui2020SCMP} and multiple orders requiring approval of publications on COVID-19~\citep{chinacdc2020, Kang2020}, this suggests a less than wholehearted effort to understand the early spread of the epidemic.
-
-Another important implication is that genomic epidemiology studies of early SARS-CoV-2 need to pay as much attention to the provenance and annotation of the underlying sequences as technical considerations.
-There has been substantial scientific effort expended on topics such as phylogenetic rooting~\citep{pipes2021assessing, morel2021phylogenetic}, novel algorithms~\citep{kumar2021evolutionary}, and correction of sequencing errors~\citep{turakhia2020stability}.
-Future studies should devote equal effort to going beyond the annotations in GISAID to carefully trace the location of patient infection and sample sequencing.
-The potential importance of such work is revealed by the observation that many of the sequences closest to SARS-CoV-2's bat coronavirus relatives are from early patients who were infected in Wuhan, but then sequenced in and attributed to Guangdong.
-
-There are several caveats to this study.
-Most obviously, the sequences recovered from the deleted data set are partial and lack full metadata.
-Therefore, it is impossible to unambiguously place them phylogenetically, or determine exactly when they were collected.
-However, little can be done to mitigate this caveat beyond my unsuccessful attempt to contact the corresponding authors of \citet{wang2020medRxiv}.
-It is also important to note that my phylogenetic analyses use relatively simple methods to draw qualitative conclusions without formal statistical testing.
-Further application of more advanced methods would be a welcome advance.
-However, qualitative and visual analyses do have advantages when the key questions relate more to the underlying data than the sophistication of the inferences.
-Finally, both plausible putative progenitors require that an early mutation to SARS-CoV-2 was a reversion towards the bat coronavirus outgroups (either C18060T or C29095T) on a branch that subsequently gave rise to multiple distinct descendants.
-Such a scenario can only be avoided by invoking recombination very early in the pandemic, which is not entirely implausible for a coronavirus~\citep{boni2020evolutionary}.
-However, because the outgroups have $\sim$4\% nucleotide divergence from SARS-CoV-2, a mutation towards the outgroup is also entirely possible.
-Of course, future identification of additional early sequences could fully resolve these questions.
-
-More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without an international investigation.
-Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.
-In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions.
-Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for its staff to validate the rationale for all requests.
-However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
-A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the virus.
-
-{\small
-
-\section{Methods}
-\subsection{Code and data availability}
-The computer code and input data necessary to reproduce all analyses described in this paper are available on GitHub at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766}.
-This GitHub repository includes a \texttt{Snakemake}~\citep{molder2021sustainable} pipeline that fully automates all steps in the analysis except for downloading of sequences from GISAID, which must be done manually as described in the GitHub repository's README in order to comply with GISAID data sharing terms.
-
-The deleted SRA files recovered from the Google Cloud are all available at  \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/results/sra_downloads}.
-I have suffixed the file extension \texttt{.sra} to all these files.
-The consensus sequences recovered from these deleted SRA files are linked to in the relevant Methods subsection below.
-
-\subsection{Archiving of key weblinks}
-I have digitally archived key weblinks in the Wayback Machine, including a subset of the SRA files from PRJNA612766 on the Google Cloud:
-\begin{itemize}
-\item The first supplementary table of \citet{farkas2020insights} is archived at \url{https://web.archive.org/web/20210502130356/https://dfzljdn9uc3pi.cloudfront.net/2020/9255/1/Supplementary_Table_1.xlsx}.
-\item SRR11313485: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}
-\item SRR11313486: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313486/SRR11313486}
-\item SRR11313274: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313274/SRR11313274}
-\item SRR11313275: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313275/SRR11313275}
-\item SRR11313285: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313285/SRR11313285}
-\item SRR11313286: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313286/SRR11313286}
-\item SRR11313448: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313448/SRR11313448}
-\item SRR11313449: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313449/SRR11313449}
-\item SRR11313427: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313427/SRR11313427}
-\item SRR11313429: \url{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313429/SRR11313429}
-\end{itemize}
-
-\subsection{Recovery of SRA files from deleted project PRJNA612766}
-I parsed the first supplementary table of \citet{farkas2020insights} to extract the accessions for sequencing runs for deleted SRA BioProject PRJNA612766.
-By cross-referencing the samples described in this table to \citet{wang2020medRxiv, wang2020small}, I identified the accessions corresponding to the 34 early outpatient samples who were positive, as well as the accessions corresponding to the 16 hospitalized patient samples from February.
-Samples had both 10 minute and 4 hour sequencing runtime accessions, which were combined in the subsequent analysis.
-I also identified the samples corresponding to the high-copy plasmid controls to enable analysis of the plasmid sequence to rule out contamination.
-The code used to parse the Excel table is available as a Jupyter notebook at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/tree/main/manual_analyses/PRJNA612766}.
-
-I recovered the SRA files from the Google Cloud by using \texttt{wget} to download files with from paths like \path{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
-Note that I cannot guarantee that these Google Cloud links will remain active, as my analyses of other deleted SRA runs (beyond the scope of this study) indicates that only sometimes are deleted SRA files still available via the Google Cloud.
-For this reason, key runs have been archived on the Wayback Machine as described above, and all downloaded SRA files relevant to this study are included in the GitHub repository.
-Note also that as described in this paper's main text, two SRA files could not be downloaded from the Google Cloud using the aforementioned method, and so are not part of this study.
-
-\subsection{FASTQ files for SRR11313490 and SRR11313499}
-I was not able to download SRA files for two runs, SRR11313490 and SRR11313499, from the Google Cloud.
-After I posted the initial version of this pre-print, I was contacted by three different individuals (see Acknowledgments) who had uncovered FASTQ or FASTA files for these two missing runs that were downloaded from the SRA prior to the deletion of PRJNA612766.
-For revised versions of the manuscript, I included analysis of the data for these two runs that I obtained from the links provided at \url{https://github.com/lifebit-ai/SARS-CoV-2/blob/master/assets/ucsc/aws_https_links.txt}.
-Inclusion of data for these two runs did not appreciably change the results of the analysis relative to that in the original version of the pre-print, since both runs corresponded to low coverage samples for which meaningful viral genetic information could not be obtained.
-
-\subsection{Alignment of recovered reads and calling of consensus sequences}
-The downloaded SRA files were converted to FASTQ files using \texttt{fasterq-dump} from the SRA Toolkit.
-The FASTQ files were pre-processed with \texttt{fastp}~\citep{chen2018fastp} to trim reads and remove low-quality ones (the exact settings using in this pre-processing are specified in the \texttt{Snakemake} file in the GitHub repository).
-
-The reads in these FASTQ files were then aligned to a SARS-CoV-2 reference genome using \texttt{minimap2}~\citep{li2018minimap2} with default settings.
-The reference genome used for the entirety of this study is proCoV2~\citep{kumar2021evolutionary}, which was generated by making the following three single-nucleotide changes to the Wuhan-Hu-1 reference (ASM985889v2) available on NCBI: C8782T, C18060T, and T28144C.
-
-I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to determine the coverage at each site by aligned nucleotides with a quality score of at least 20.
-I also masked (set to zero coverage) all sites overlapped by the primers used by \citet{wang2020medRxiv} to PCR the amplicons used for sequencing.
-These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
-I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
-These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} only covers part of the genome.
-
-I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$90\% of the sites in the region of interest from site 21,570 to 29,550.
-These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
-These sequences in the region of interest (21,570 to 29,550) are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/recovered_seqs.fa}.
-
-\subsection{Bat coronavirus outgroup sequences}
-For analyses that involved comparisons to SARS-CoV-2's bat coronavirus relatives~\citep{lytras2021exploring}, the bat coronavirus sequences were manually downloaded from GISAID~\citep{shu2017gisaid}.
-The sequences used were RaTG13~\citep{zhou2020pneumonia}, RmYN02~\citep{zhou2020novel}, and RpYN06~\citep{zhou2021identification}---although the multiple sequence alignment of these viruses to SARS-CoV-2 also contains PrC31~\citep{li2021novel}, which was not used in the final analyses as it more diverged from SARS-CoV-2 than the other three bat coronaviruses at a whole-genome level.
-The GISAID accessions for these sequences are listed at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/comparator_genomes_gisaid/accessions.txt}, and a table acknowledging the labs and authors is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/comparator_genomes_gisaid/acknowledgments.csv}.
-Sites in SARS-CoV-2 were mapped to their corresponding nucleotide identities in the bat coronavirus outgroups via a multiple sequence alignment of proCoV2 to the bat coronaviruses generated using \texttt{mafft}~\citep{katoh2013mafft}.
-
-\subsection{Curation and analysis of early SARS-CoV-2 sequences from GISAID}
-For the broader analyses of existing SARS-CoV-2 sequences, I downloaded all sequences from collected prior to March of 2020 from GISAID.
-The accessions of these sequences are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/accessions.txt}, and a table acknowledging the labs and authors is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/acknowledgments.csv}.
-
-I then used \texttt{mafft}~\citep{katoh2013mafft} to align these sequences to the proCoV2 reference described above, stripped any sites that were gapped relative to the reference, and filtered the sequences using the following criteria:
-\begin{itemize}
-\item I removed any sequences collected after February 28, 2020.
-\item I removed any sequences that had $\ge$4 mutations within any 10-nucleotide stretch, as such runs of mutations often indicate sequencing errors.
-\item I removed any sequence for which the alignment covered $<$90\% of the proCoV2 sequence.
-\item I removed any sequence with $\ge$15 mutations relative to the reference.
-\item I removed any sequence with $\ge$5,000 ambiguous nucleotides.
-\end{itemize}
-
-I then annotated the sequences using some additional information.
-First, I annotated sequences based on the joint WHO-China report~\citep{WHO2021origins} and also \citet{zhu2020brief} to keep only one representative from multiply sequenced patients, and to indicate which sequences were from patients associated with the Huanan Seafood Market.
-My version of these annotations is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/WHO_China_Report_Dec2019_cases.yaml}.
-Next, I identified some sequences in the set that were clearly duplicates from the same patient, and removed these.
-The annotations used to remove these duplicates are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/seqs_to_exclude.yaml}.
-Finally, I used information from ~\citet{chan2020familial} and \citet{kang2020evidence} to identify patients who were infected in Wuhan before January 5 of 2020, but ultimately sequenced in Guangdong: these annotations are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/Wuhan_exports.yaml}.
-
-I next removed any of the handful of mutations noted by \citet{turakhia2020stability} to be lab artifacts that commonly afflict SARS-CoV-2 sequences.
-I also limited the analyses to the region of the genome that spans from the start of the first coding region (ORF1ab) to the end of the last (ORF10), because I noticed that some sequences had suspicious patterns (such as many mutations or runs of mutations) near the termini of the genome.
-
-The plot in Figure~\ref{fig:deltadist_RaTG13} contains all of the GISAID sequences after this filtering.
-The plot in Figure~\ref{fig:deltadist_jitter} shows the filtered GISAID sequences collected before February of 2020 plus the 13 good coverage recovered partial early outpatient sequences (Table~\ref{tab:mutations}), considering only the region covered by the partial sequences (21,570 to 29,550).
-
-\subsection{Phylogenetic analyses}
-The phylogenetic trees were inferred using the GISAD sequences after the filtering and annotations described above, only considering sequences with $\ge$95\% coverage over the region of interest that were collected before February of 2020.
-In addition, after generating this sequence set I removed any sequence variants with a combination of mutations that was not observed at least twice so the analysis only includes multiply observed sequence variants.
-A file indicating the unique sequences used for the phylogenetic analysis, their mutations relative to proCoV2, and other sequences in that cluster is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/phylogenetics/all_alignment.csv}.
-
-I then used \texttt{IQ-Tree}~\citep{minh2020iq} to infer a maximum-likelihood phylogenetic tree using a GTR nucleotide substitution model with empirical nucleotide frequencies, and collapsing zero-length branches to potentially allow a multifurcating tree.
-The inference yielded the tree topology and branch lengths shown in all figures in this study with phylogenetic trees.
-I then rendered the images of the tree using \texttt{ETE 3}~\citep{huerta2016ete}, manually re-rooting the tree to place the first (progenitor) node at each of the three nodes that have the highest identity to the bat coronavirus outgroup.
-In these images, node sizes are proportional to the number of sequences in that node, and are colored in proportion to the location from which those sequences are derived.
-As indicated in the legend to Figure~\ref{fig:tree_RaTG13}, the node containing the monophyletic set of sequences with C28144T is collapsed into a single node in the tree images.
-
-For the trees in which I added the recovered sequences from the deleted data set (Figure~\ref{fig:tree_with_deleted}), the actual trees are exactly the same as those inferred using the GISAID sequences above.
-The difference is that the sequences from the deleted data set are then added to each node with which they are compatible given their mutations in an amount proportional to the size of the node, the logic being that a sequence is more likely to fall into larger clusters.
-
-\subsection{Interactive versions of some figures}
-Interactive versions of some figures are available at \url{https://jbloom.github.io/SARS-CoV-2_PRJNA612766/}, and were created using \texttt{Altair}~\citep{vanderplas2018altair}
-
-\section{Acknowledgments}
-I thank the citizens and scientists on Twitter who helped inspire this study and inform its background through their analyses and discussions of the early spread of SARS-CoV-2.
-I thank Brendan Larsen for pointing out that I failed to mask the primer binding site regions in the analysis described in the original pre-print (this has been fixed in this version).
-I thank Andrew Fire (Stanford University), Theo Sanderson (Crick Institute), and Francisco de Asis de Ribera (DRASTIC Research) for finding locations of downloaded data for the two accessions (SRR11313490 and SRR11313499) that I could not download from the Google Cloud and so had to omit from the analysis described in the original pre-print (they are included in the analysis in this version).
-I thank the scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.
-The scientific computing infrastructure used in this work was supported by the NIH Office of Research Infrastructure Programs under S10OD028685.
-The author is an Investigator of the Howard Hughes Medical Institute.
-
-\section{Competing interests}
-The author consults for Moderna on SARS-CoV-2 evolution and epidemiology, consults for Flagship Labs 77 on viral evolution and deep mutational scanning, and has the potential to receive a share of IP revenue as an inventor on a Fred Hutch licensed technology/patent (application WO2020006494) related to deep mutational scanning of viral proteins.
-
-}
-
-\bibliography{references}
-
-\clearpage
 \onecolumn

### 2021-08-19 paper/paper_with_refs.tex

--- 2021-08-04 paper/paper.tex
@@
 
-\bibliography{references}
+\begin{thebibliography}{47}
+
+\bibitem[{Bedford {\em et~al.\/}}(2020){Bedford, Greninger, Roychoudhury,
+  Starita, Famulare, Huang, Nalla, Pepper, Reinhardt, Xie, {\em
+  et~al.\/}}]{bedford2020cryptic}
+{Bedford, T., A.~L. Greninger, P.~Roychoudhury, L.~M. Starita, M.~Famulare,
+  {\em et~al.\/}}, 2020 Cryptic transmission of {SARS-CoV-2} in washington
+  state. Science {\bf 370}: 571--575.
+
+\bibitem[{Boni {\em et~al.\/}}(2020){Boni, Lemey, Jiang, Lam, Perry, Castoe,
+  Rambaut, {\rm and} Robertson}]{boni2020evolutionary}
+{Boni, M.~F., P.~Lemey, X.~Jiang, T.~T.-Y. Lam, B.~W. Perry, {\em et~al.\/}},
+  2020 {Evolutionary origins of the SARS-CoV-2 sarbecovirus lineage responsible
+  for the COVID-19 pandemic}. Nature Microbiology {\bf 5}: 1408--1417.
+
+\bibitem[{Chan {\em et~al.\/}}(2020){Chan, Yuan, Kok, To, Chu, Yang, Xing, Liu,
+  Yip, Poon, {\em et~al.\/}}]{chan2020familial}
+{Chan, J. F.-W., S.~Yuan, K.-H. Kok, K.~K.-W. To, H.~Chu, {\em et~al.\/}}, 2020
+  A familial cluster of pneumonia associated with the 2019 novel coronavirus
+  indicating person-to-person transmission: a study of a family cluster. Lancet
+  {\bf 395}: 514--523.
+
+\bibitem[{Chen {\em et~al.\/}}(2020){Chen, Zhou, Dong, Qu, Gong, Han, Qiu,
+  Wang, Liu, Wei, {\em et~al.\/}}]{chen2020epidemiological}
+{Chen, N., M.~Zhou, X.~Dong, J.~Qu, F.~Gong, {\em et~al.\/}}, 2020
+  Epidemiological and clinical characteristics of 99 cases of 2019 novel
+  coronavirus pneumonia in {Wuhan, China}: a descriptive study. Lancet {\bf
+  395}: 507--513.
+
+\bibitem[{Chen {\em et~al.\/}}(2018){Chen, Zhou, Chen, {\rm and}
+  Gu}]{chen2018fastp}
+{Chen, S., Y.~Zhou, Y.~Chen, {\rm and} J.~Gu}, 2018 fastp: an ultra-fast
+  all-in-one fastq preprocessor. Bioinformatics {\bf 34}: i884--i890.
+
+\bibitem[{{China CDC}}(2020)]{chinacdc2020}
+{{China CDC}}, 2020 {Memo to the Offices of the Chinese Center for Disease
+  Control and Prevention}.
+  \url{https://www.documentcloud.org/documents/7340336-China-CDC-Sup-Regs.html}.
+
+\bibitem[{Cohen}(2020)]{cohen2020wuhan}
+{Cohen, J.}, 2020 Wuhan seafood market may not be source of novel virus
+  spreading globally. Science {\bf 10}: 10.1126/science.abb0611.
+
+\bibitem[{Deng {\em et~al.\/}}(2020){Deng, Gu, Federman, Du~Plessis, Pybus,
+  Faria, Wang, Yu, Bushnell, Pan, {\em et~al.\/}}]{deng2020genomic}
+{Deng, X., W.~Gu, S.~Federman, L.~Du~Plessis, O.~G. Pybus, {\em et~al.\/}},
+  2020 Genomic surveillance reveals multiple introductions of {SARS-CoV-2 into
+  Northern California}. Science {\bf 369}: 582--587.
+
+\bibitem[{Farkas {\em et~al.\/}}(2020){Farkas, Fuentes-Villalobos, Garrido,
+  Haigh, {\rm and} Barr{\'\i}a}]{farkas2020insights}
+{Farkas, C., F.~Fuentes-Villalobos, J.~L. Garrido, J.~Haigh, {\rm and} M.~I.
+  Barr{\'\i}a}, 2020 Insights on early mutational events in {SARS-CoV-2} virus
+  reveal founder effects across geographical regions. PeerJ {\bf 8}: e9255.
+
+\bibitem[{Fauver {\em et~al.\/}}(2020){Fauver, Petrone, Hodcroft, Shioda,
+  Ehrlich, Watts, Vogels, Brito, Alpert, Muyombwe, {\em
+  et~al.\/}}]{fauver2020coast}
+{Fauver, J.~R., M.~E. Petrone, E.~B. Hodcroft, K.~Shioda, H.~Y. Ehrlich, {\em
+  et~al.\/}}, 2020 {Coast-to-coast spread of SARS-CoV-2 during the early
+  epidemic in the United States}. Cell {\bf 181}: 990--996.
+
+\bibitem[{Forster {\em et~al.\/}}(2020){Forster, Forster, Renfrew, {\rm and}
+  Forster}]{forster2020phylogenetic}
+{Forster, P., L.~Forster, C.~Renfrew, {\rm and} M.~Forster}, 2020 Phylogenetic
+  network analysis of {SARS-CoV-2} genomes. Proceedings of the National Academy
+  of Sciences {\bf 117}: 9241--9243.
+
+\bibitem[{Garry}(2021)]{garry2021early}
+{Garry, R.~F.}, 2021 Early appearance of two distinct genomic lineages of
+  {SARS-CoV-2} in different wuhan wildlife markets suggests {SARS-CoV-2} has a
+  natural origin.
+  \url{https://virological.org/t/early-appearance-of-two-distinct-genomic-lineages-of-sars-cov-2-in-different-wuhan-wildlife-markets-suggests-sars-cov-2-has-a-natural-origin/691}.
+
+\bibitem[{{Global Times}}(2020)]{GlobalTimes2020}
+{{Global Times}}, 2020 Wuhan's huanan seafood market a victim of {COVID-19}:
+  {CDC} director. \url{https://www.globaltimes.cn/content/1189506.shtml},
+  archived at
+  \url{https://web.archive.org/web/20200528062530/https://www.globaltimes.cn/content/1189506.shtml}.
+
+\bibitem[{{Health Times}}(2020)]{healthtimes2020}
+{{Health Times}}, 2020 Experts judge the source of the new coronavirus:
+  {December} 8 last year may not be the earliest time of onset.
+  \url{https://www.guancha.cn/politics/2020_02_27_538822.shtml}, archived and
+  detailed at
+  \url{https://docs.google.com/document/d/e/2PACX-1vTQxG822DtqP7IZSjLj751Mrm8Ev7leksXfjBLsA9KJ0_tbGV6YJAAjuijPnwz_YmUQGY1PZUl5LcCI/pub}.
+
+\bibitem[{Huang {\em et~al.\/}}(2020){Huang, Wang, Li, Ren, Zhao, Hu, Zhang,
+  Fan, Xu, Gu, {\em et~al.\/}}]{huang2020clinical}
+{Huang, C., Y.~Wang, X.~Li, L.~Ren, J.~Zhao, {\em et~al.\/}}, 2020 Clinical
+  features of patients infected with 2019 novel coronavirus in {Wuhan, China}.
+  Lancet {\bf 395}: 497--506.
+
+\bibitem[{Huerta-Cepas {\em et~al.\/}}(2016){Huerta-Cepas, Serra, {\rm and}
+  Bork}]{huerta2016ete}
+{Huerta-Cepas, J., F.~Serra, {\rm and} P.~Bork}, 2016 Ete 3: reconstruction,
+  analysis, and visualization of phylogenomic data. Molecular biology and
+  evolution {\bf 33}: 1635--1638.
+
+\bibitem[{Kang {\em et~al.\/}}(2020{a}){Kang, Cheng, {\rm and}
+  McNeil}]{Kang2020}
+{Kang, D., M.~Cheng, {\rm and} S.~McNeil}, 2020{a} China clamps down in hidden
+  hunt for coronavirus origins.
+  \url{https://apnews.com/article/united-nations-coronavirus-pandemic-china-only-on-ap-bats-24fbadc58cee3a40bca2ddf7a14d2955},
+  The actual China State Council order described in the article is at
+  \url{https://www.documentcloud.org/documents/7340337-State-Research-regulations.html}.
+
+\bibitem[{Kang {\em et~al.\/}}(2020{b}){Kang, Wu, Ma, He, Lu, Liu, Li, Mei,
+  Ruan, Lin, Zou, Ke, Zhong, Zhang, Chen, Liu, Zhu, Xiao, Yu, Hu, Zeng, Li,
+  Liao, Tang, Xiao, Wang, Song, Zhuang, Liang, Zeng, He, Lin, Deng, {\rm and}
+  Song}]{kang2020evidence}
+{Kang, M., J.~Wu, W.~Ma, J.~He, J.~Lu, {\em et~al.\/}}, 2020{b} Evidence and
+  characteristics of human-to-human transmission of sars-cov-2. medRxiv .
+
+\bibitem[{Katoh {\rm and} Standley}(2013)]{katoh2013mafft}
+{Katoh, K. {\rm and} D.~M. Standley}, 2013 Mafft multiple sequence alignment
+  software version 7: improvements in performance and usability. Molecular
+  biology and evolution {\bf 30}: 772--780.
+
+\bibitem[{Kumar {\em et~al.\/}}(2021){Kumar, Tao, Weaver, Sanderford,
+  Caraballo-Ortiz, Sharma, Pond, {\rm and} Miura}]{kumar2021evolutionary}
+{Kumar, S., Q.~Tao, S.~Weaver, M.~Sanderford, M.~A. Caraballo-Ortiz, {\em
+  et~al.\/}}, 2021 An evolutionary portrait of the progenitor {SARS-CoV-2} and
+  its dominant offshoots in {COVID-19} pandemic. Molecular Biology and
+  Evolution msab118.
+
+\bibitem[{Li}(2018)]{li2018minimap2}
+{Li, H.}, 2018 Minimap2: pairwise alignment for nucleotide sequences.
+  Bioinformatics {\bf 34}: 3094--3100.
+
+\bibitem[{Li {\em et~al.\/}}(2021){Li, Wang, Ma, Li, Yang, Shi, {\rm and}
+  Duan}]{li2021novel}
+{Li, L., J.~Wang, X.~Ma, J.~Li, X.~Yang, {\em et~al.\/}}, 2021 A novel
+  sars-cov-2 related virus with complex recombination isolated from bats in
+  yunnan province, china. bioRxiv .
+
+\bibitem[{Lytras {\em et~al.\/}}(2021){Lytras, Hughes, Martin, de~Klerk,
+  Lourens, Kosakovsky~Pond, Xia, Jiang, {\rm and}
+  Robertson}]{lytras2021exploring}
+{Lytras, S., J.~Hughes, D.~Martin, A.~de~Klerk, R.~Lourens, {\em et~al.\/}},
+  2021 {Exploring the natural origins of SARS-CoV-2 in the light of
+  recombination}. bioRxiv .
+
+\bibitem[{Ma}(2020)]{ma2020SCMP}
+{Ma, J.}, 2020 {Coronavirus: China’s first confirmed Covid-19 case traced
+  back to November 17}. South China Morning Post,
+  \url{https://www.scmp.com/news/china/society/article/3074991/coronavirus-chinas-first-confirmed-covid-19-case-traced-back},
+  archived at
+  \url{https://web.archive.org/web/20200315011702/https://www.scmp.com/news/china/society/article/3074991/coronavirus-chinas-first-confirmed-covid-19-case-traced-back}.
+
+\bibitem[{Minh {\em et~al.\/}}(2020){Minh, Schmidt, Chernomor, Schrempf,
+  Woodhams, Von~Haeseler, {\rm and} Lanfear}]{minh2020iq}
+{Minh, B.~Q., H.~A. Schmidt, O.~Chernomor, D.~Schrempf, M.~D. Woodhams, {\em
+  et~al.\/}}, 2020 {IQ-TREE} 2: New models and efficient methods for
+  phylogenetic inference in the genomic era. Molecular Biology and Evolution
+  {\bf 37}: 1530--1534.
+
+\bibitem[{M{\"o}lder {\em et~al.\/}}(2021){M{\"o}lder, Jablonski, Letcher,
+  Hall, Tomkins-Tinch, Sochat, Forster, Lee, Twardziok, Kanitz, {\em
+  et~al.\/}}]{molder2021sustainable}
+{M{\"o}lder, F., K.~P. Jablonski, B.~Letcher, M.~B. Hall, C.~H. Tomkins-Tinch,
+  {\em et~al.\/}}, 2021 Sustainable data analysis with snakemake. F1000Research
+  {\bf 10}: 33.
+
+\bibitem[{Morel {\em et~al.\/}}(2021){Morel, Barbera, Czech, Bettisworth,
+  H{\"u}bner, Lutteropp, Serdari, Kostaki, Mamais, Kozlov, {\em
+  et~al.\/}}]{morel2021phylogenetic}
+{Morel, B., P.~Barbera, L.~Czech, B.~Bettisworth, L.~H{\"u}bner, {\em
+  et~al.\/}}, 2021 Phylogenetic analysis of {SARS-CoV-2} data is difficult.
+  Molecular Biology and Evolution {\bf 38}: 1777--1791.
+
+\bibitem[{Pekar {\em et~al.\/}}(2021){Pekar, Worobey, Moshiri, Scheffler, {\rm
+  and} Wertheim}]{pekar2021timing}
+{Pekar, J., M.~Worobey, N.~Moshiri, K.~Scheffler, {\rm and} J.~O. Wertheim},
+  2021 Timing the {SARS-CoV-2} index case in hubei province. Science {\bf 372}:
+  412--417.
+
+\bibitem[{Pingui}(2020)]{pinghui2020SCMP}
+{Pingui, Z.}, 2020 China confirms unauthorised labs were told to destroy early
+  coronavirus samples. South China Morning Post,
+  \url{https://www.scmp.com/news/china/society/article/3084635/china-confirms-unauthorised-labs-were-told-destroy-early},
+  archived at
+  \url{https://web.archive.org/web/20210103124552/https://www.scmp.com/news/china/society/article/3084635/china-confirms-unauthorised-labs-were-told-destroy-early}.
+
+\bibitem[{Pipes {\em et~al.\/}}(2021){Pipes, Wang, Huelsenbeck, {\rm and}
+  Nielsen}]{pipes2021assessing}
+{Pipes, L., H.~Wang, J.~P. Huelsenbeck, {\rm and} R.~Nielsen}, 2021 Assessing
+  uncertainty in the rooting of the {SARS-CoV-2} phylogeny. Molecular Biology
+  and Evolution {\bf 38}: 1537--1543.
+
+\bibitem[{ProMED}(2019)]{ProMED2109}
+{ProMED}, 2019 Undiagnosed pneumonia---{China (Hubei)}: Request for
+  information. \url{https://promedmail.org/promed-post/?id=6864153}.
+
+\bibitem[{Rambaut {\em et~al.\/}}(2020){Rambaut, Holmes, O’Toole, Hill,
+  McCrone, Ruis, du~Plessis, {\rm and} Pybus}]{rambaut2020dynamic}
+{Rambaut, A., E.~C. Holmes, {\'A}.~O’Toole, V.~Hill, J.~T. McCrone, {\em
+  et~al.\/}}, 2020 A dynamic nomenclature proposal for {SARS-CoV-2} lineages to
+  assist genomic epidemiology. Nature Microbiology {\bf 5}: 1403--1407.
+
+\bibitem[{Shu {\rm and} McCauley}(2017)]{shu2017gisaid}
+{Shu, Y. {\rm and} J.~McCauley}, 2017 {GISAID}: Global initiative on sharing
+  all influenza data--from vision to reality. Eurosurveillance {\bf 22}: 30494.
+
+\bibitem[{SRA}(2021)]{SRA_deletion}
+{SRA}, 2021 {SRA} data updates.
+  \url{https://www.ncbi.nlm.nih.gov/sra/docs/submitupdate/#how-do-i-withdraw-sra-data},
+  Last accessed June 2, 2021; a copy of the page is digitally archived on the
+  Wayback Machine at
+  \url{https://web.archive.org/web/20210505002105/https://www.ncbi.nlm.nih.gov/sra/docs/submitupdate/}.
+
+\bibitem[{Turakhia {\em et~al.\/}}(2020){Turakhia, De~Maio, Thornlow, Gozashti,
+  Lanfear, Walker, Hinrichs, Fernandes, Borges, Slodkowicz, {\em
+  et~al.\/}}]{turakhia2020stability}
+{Turakhia, Y., N.~De~Maio, B.~Thornlow, L.~Gozashti, R.~Lanfear, {\em
+  et~al.\/}}, 2020 Stability of {SARS-CoV-2} phylogenies. PLoS Genetics {\bf
+  16}: e1009175.
+
+\bibitem[{VanderPlas {\em et~al.\/}}(2018){VanderPlas, Granger, Heer, Moritz,
+  Wongsuphasawat, Satyanarayan, Lees, Timofeev, Welsh, {\rm and}
+  Sievert}]{vanderplas2018altair}
+{VanderPlas, J., B.~E. Granger, J.~Heer, D.~Moritz, K.~Wongsuphasawat, {\em
+  et~al.\/}}, 2018 Altair: interactive statistical visualizations for {Python}.
+  Journal of Open Source Software {\bf 3}: 1057.
+
+\bibitem[{Wang {\em et~al.\/}}(2020{a}){Wang, Fu, Hu, Tong, Liu, Gu, Liu,
+  Jiang, Shen, Zhao, Men, Deng, Yu, Li, {\rm and} Liu}]{wang2020medRxiv}
+{Wang, M., A.~Fu, B.~Hu, Y.~Tong, R.~Liu, {\em et~al.\/}}, 2020{a} Nanopore
+  target sequencing for accurate and comprehensive detection of {SARS-CoV-2}
+  and other respiratory viruses. medRxiv {\bf 10.1101/2020.03.04.20029538}.
+
+\bibitem[{Wang {\em et~al.\/}}(2020{b}){Wang, Fu, Hu, Tong, Liu, Liu, Gu,
+  Xiang, Liu, Jiang, Shen, Zhao, Men, Deng, Yu, Wei, Li, {\rm and}
+  Liu}]{wang2020small}
+{Wang, M., A.~Fu, B.~Hu, Y.~Tong, R.~Liu, {\em et~al.\/}}, 2020{b} Nanopore
+  targeted sequencing for the accurate and comprehensive detection of
+  {SARS-CoV-2} and other respiratory viruses. Small {\bf 16}: 2002169.
+
+\bibitem[{Wang}(2021{a})]{wang2021exclusive}
+{Wang, Z.}, 2021{a} {Exclusive: Chinese researcher, at the center of COVID data
+  withdrawal controversy, speaks up}.
+  \url{https://beijingchannel.substack.com/p/exclusive-chinese-researcher-at-the}.
+
+\bibitem[{Wang}(2021{b})]{wang2021why}
+{Wang, Z.}, 2021{b} {Why did Wuhan University researchers delete COVID-19 data
+  at NIH?}
+  \url{https://pekingnology.substack.com/p/why-did-wuhan-university-researchers}.
+
+\bibitem[{{WHO}}(2021)]{WHO2021origins}
+{{WHO}}, 2021 {WHO-convened global study of origins of SARS-CoV-2: China Part}.
+  \url{https://www.who.int/publications/i/item/who-convened-global-study-of-origins-of-sars-cov-2-china-part}.
+
+\bibitem[{Worobey {\em et~al.\/}}(2020){Worobey, Pekar, Larsen, Nelson, Hill,
+  Joy, Rambaut, Suchard, Wertheim, {\rm and} Lemey}]{worobey2020emergence}
+{Worobey, M., J.~Pekar, B.~B. Larsen, M.~I. Nelson, V.~Hill, {\em et~al.\/}},
+  2020 {The emergence of SARS-CoV-2 in Europe and North America}. Science {\bf
+  370}: 564--570.
+
+\bibitem[{Yang {\em et~al.\/}}(2020){Yang, Lu, Liu, Wang, Zhang, Jalali, Dean,
+  Longini, Halloran, Xu, Zhang, Wang, Liu, {\rm and}
+  Fang}]{yang2020epidemiological}
+{Yang, Y., Q.-B. Lu, M.-J. Liu, Y.-X. Wang, A.-R. Zhang, {\em et~al.\/}}, 2020
+  Epidemiological and clinical features of the 2019 novel coronavirus outbreak
+  in china. medRxiv {\bf 10.1101/2020.02.10.20021675}.
+
+\bibitem[{Zhou {\em et~al.\/}}(2020{a}){Zhou, Chen, Hu, Li, Song, Liu, Wang,
+  Liu, Yang, Holmes, {\em et~al.\/}}]{zhou2020novel}
+{Zhou, H., X.~Chen, T.~Hu, J.~Li, H.~Song, {\em et~al.\/}}, 2020{a} A novel bat
+  coronavirus closely related to sars-cov-2 contains natural insertions at the
+  s1/s2 cleavage site of the spike protein. Current Biology {\bf 30}:
+  2196--2203.
+
+\bibitem[{Zhou {\em et~al.\/}}(2021){Zhou, Ji, Chen, Bi, Li, Wang, Hu, Song,
+  Zhao, Chen, {\em et~al.\/}}]{zhou2021identification}
+{Zhou, H., J.~Ji, X.~Chen, Y.~Bi, J.~Li, {\em et~al.\/}}, 2021 Identification
+  of novel bat coronaviruses sheds light on the evolutionary origins of
+  sars-cov-2 and related viruses. Cell .
+
+\bibitem[{Zhou {\em et~al.\/}}(2020{b}){Zhou, Yang, Wang, Hu, Zhang, Zhang, Si,
+  Zhu, Li, Huang, {\em et~al.\/}}]{zhou2020pneumonia}
+{Zhou, P., X.-L. Yang, X.-G. Wang, B.~Hu, L.~Zhang, {\em et~al.\/}}, 2020{b} A
+  pneumonia outbreak associated with a new coronavirus of probable bat origin.
+  Nature {\bf 579}: 270--273.
+
+\bibitem[{Zhu {\em et~al.\/}}(2020){Zhu, Zhang, Wang, Li, Yang, Song, Zhao,
+  Huang, Shi, Lu, {\em et~al.\/}}]{zhu2020brief}
+{Zhu, N., D.~Zhang, W.~Wang, X.~Li, B.~Yang, {\em et~al.\/}}, 2020 Brief
+  report: A novel coronavirus from patients with pneumonia in china, 2019. The
+  New England Journal of Medicine {\bf 382}: 727.
+
+\end{thebibliography}
 
